---
title: "Exercise — scATAC-seq with Signac"
subtitle: "From peaks matrix to clustered cells (PBMC 10k, 10x)"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this exercise

A parallel to the scRNA-seq tutorial series — but for **chromatin accessibility** (scATAC-seq). The workflow follows the standard **Signac** pipeline from the Stuart Lab, applied to the 10x PBMC 10k scATAC-seq dataset.

-   Reference vignette: [Signac PBMC scATAC-seq vignette](https://stuartlab.org/signac/articles/pbmc_vignette)
-   Workflow source: Khushbu Patel, single-cell ATAC-seq standard workflow
-   Data: [10x Genomics — PBMC 10k scATAC-seq](https://www.10xgenomics.com/resources/datasets/10-k-peripheral-blood-mononuclear-cells-pbm-cs-from-a-healthy-donor-next-gem-v-1-1-1-1-standard-2-0-0)

::: {.callout-tip title="What's different about ATAC-seq?"}
ATAC-seq measures **open chromatin** (regions accessible to Tn5), not transcription. The raw unit is a **fragment** (a pair of Tn5 cut sites). The "features" are **peaks** — genomic intervals of accessibility — rather than genes. Most of the Seurat idioms carry over, but the **normalization**, **dimensionality reduction**, and **QC metrics** are ATAC-specific.
:::

## Learning goals

By the end of this exercise you will be able to:

-   Load peak-count and fragment data into a Signac `ChromatinAssay` inside a Seurat object
-   Compute and interpret ATAC-specific QC metrics (nucleosome signal, TSS enrichment, blacklist ratio, peak-read fraction)
-   Run TF-IDF normalization and SVD (LSI) on accessibility data
-   Produce a UMAP and cluster scATAC-seq cells

## Setup

```{r}
#| label: setup
#| eval: true

# One-time install:
# remotes::install_github("stuart-lab/signac", ref = "develop")
# install.packages("Matrix", type = "source")
# install.packages("irlba",  type = "source")
# BiocManager::install("EnsDb.Hsapiens.v75")

library(Signac)
library(Seurat)
library(EnsDb.Hsapiens.v75)
library(tidyverse)

set.seed(2026)
```

## Data files

You will need three files from the 10x PBMC 10k scATAC-seq release, placed under `../data/`:

| File | Purpose |
|---|---|
| `atac_v1_pbmc_10k_filtered_peak_bc_matrix.h5` | peak × barcode counts |
| `atac_v1_pbmc_10k_fragments.tsv.gz` (+ `.tbi` index) | per-fragment record |
| `atac_v1_pbmc_10k_singlecell.csv` | per-barcode metadata |

------------------------------------------------------------------------

## Step 1 — Peek at the fragment file

```{r}
#| label: peek-fragments

frag.file <- read.delim(
  "../data/atac_v1_pbmc_10k_fragments.tsv.gz",
  header = FALSE,
  nrows  = 10
)
head(frag.file)
```

::: {.callout-important title="Think about it"}
1.  A fragment file has one line per Tn5 fragment. What are the five columns typically recorded?
2.  Why do we keep the fragment file around instead of just using the peak-counts matrix?

<details>

<summary>Show answers</summary>

1.  `chrom`, `start`, `end`, `cell_barcode`, `duplicate_count` (number of reads collapsed to that fragment).
2.  Many ATAC analyses (TSS enrichment, nucleosome signal, coverage tracks, motif activity) work at the fragment level, not the peak level. The fragment file is the raw, peak-independent record; peaks can be recomputed later.

</details>
:::

## Step 2 — Load the peaks matrix and create the object

```{r}
#| label: load-matrix

counts <- Read10X_h5("../data/atac_v1_pbmc_10k_filtered_peak_bc_matrix.h5")
counts[1:5, 1:5]

chrom_assay <- CreateChromatinAssay(
  counts       = counts,
  sep          = c(":", "-"),
  fragments    = "../data/atac_v1_pbmc_10k_fragments.tsv.gz",
  min.cells    = 10,
  min.features = 200
)

metadata <- read.csv(
  file      = "../data/atac_v1_pbmc_10k_singlecell.csv",
  header    = TRUE,
  row.names = 1
)

pbmc <- CreateSeuratObject(
  counts    = chrom_assay,
  meta.data = metadata,
  assay     = "ATAC"
)

pbmc
```

::: {.callout-important title="Think about it"}
1.  The row names of an RNA assay are gene symbols. What do row names of a `ChromatinAssay` look like, and why?
2.  We pass both `counts` and `fragments` when building the ChromatinAssay. Why is that redundant-looking pairing actually useful?

<details>

<summary>Show answers</summary>

1.  Peak coordinates — strings like `chr1:100234-100734`. Peaks are genomic intervals, not annotated features, so the natural identifier is the interval itself.
2.  `counts` is the fast aggregated view (peak × cell). `fragments` retains per-fragment, per-base precision for downstream peak recomputation, coverage plots, TSS enrichment, motif accessibility, etc. You want both.

</details>
:::

## Step 3 — Add gene annotation

```{r}
#| label: annotation

annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v75)
seqlevels(annotations) <- paste0("chr", seqlevels(annotations))  # UCSC naming
Annotation(pbmc) <- annotations
```

::: {.callout-important title="Think about it"}
1.  The data is mapped to **hg19** (GRCh37). Why `EnsDb.Hsapiens.v75` specifically?
2.  What would you change for a dataset mapped to hg38 or to mm10?

<details>

<summary>Show answers</summary>

1.  Ensembl v75 is the last annotation release pinned to GRCh37 / hg19. Using a matching annotation is essential — otherwise TSSs and gene bodies will be in the wrong places.
2.  hg38: use a current human EnsDb (e.g. `EnsDb.Hsapiens.v86` or newer). mm10: use `EnsDb.Mmusculus.v79` or equivalent. Chromosome-name style also differs (`1` vs `chr1`); convert to match the counts.

</details>
:::

## Step 4 — Compute ATAC-specific QC metrics

```{r}
#| label: qc-compute

pbmc <- NucleosomeSignal(pbmc)
pbmc <- TSSEnrichment(object = pbmc, fast = FALSE)

pbmc$blacklist_ratio    <- pbmc$blacklist_region_fragments / pbmc$peak_region_fragments
pbmc$pct_reads_in_peaks <- pbmc$peak_region_fragments / pbmc$passed_filters * 100

colnames(pbmc@meta.data)
```

::: {.callout-important title="Think about it"}
1.  What does a **high nucleosome signal** indicate?
2.  Why is **TSS enrichment** a quality metric, not a biology metric?
3.  What is the ENCODE blacklist, and why exclude it?

<details>

<summary>Show answers</summary>

1.  High nucleosome signal means an excess of multi-nucleosome-length fragments (vs sub-nucleosomal). Good ATAC libraries show a clear bias toward sub-nucleosomal fragments in open chromatin. High signal flags over-digested or low-quality cells.
2.  Because active TSSs are *reliably* open in essentially all cells. Good cells show a sharp accessibility peak at TSSs; low enrichment at TSSs signals poor Tn5 activity or barcode contamination regardless of cell type.
3.  A curated list of regions with anomalously high signal in *any* experiment (repeats, centromeres, mappability artefacts). Reads falling there are technical noise; a high blacklist ratio indicates a problematic cell or library.

</details>
:::

## Step 5 — Visualize QC

```{r}
#| label: qc-viz

a1 <- DensityScatter(pbmc, x = "nCount_ATAC",
                     y = "TSS.enrichment", log_x = TRUE, quantiles = TRUE)
a2 <- DensityScatter(pbmc, x = "nucleosome_signal",
                     y = "TSS.enrichment", log_x = TRUE, quantiles = TRUE)
a1 | a2

VlnPlot(
  object   = pbmc,
  features = c("nCount_ATAC", "nFeature_ATAC", "TSS.enrichment",
               "nucleosome_signal", "blacklist_ratio", "pct_reads_in_peaks"),
  pt.size  = 0.1,
  ncol     = 6
)
```

::: {.callout-important title="Think about it"}
1.  In the `nCount_ATAC` × `TSS.enrichment` density plot, where should *good* cells sit?
2.  What would a blob at high `nCount_ATAC` and **low** TSS enrichment suggest?

<details>

<summary>Show answers</summary>

1.  Middle-to-high `nCount_ATAC` with high TSS enrichment — enough reads, concentrated at bona-fide open regions.
2.  Many reads but they're **not** at TSSs. Classic signature of a bad library, ambient reads, or reads from dead cells — not useful signal.

</details>
:::

## Step 6 — Filter low-quality cells

```{r}
#| label: filter

pbmc <- subset(
  x = pbmc,
  subset = nCount_ATAC       > 3000  &
           nCount_ATAC       < 30000 &
           pct_reads_in_peaks > 15   &
           blacklist_ratio   < 0.05  &
           nucleosome_signal < 4     &
           TSS.enrichment    > 3
)
pbmc
```

::: {.callout-important title="Think about it"}
1.  These thresholds come from the Signac PBMC vignette. Would you use them as-is on another dataset?
2.  What's the analogue of `percent.mt` in ATAC, and is it relevant?

<details>

<summary>Show answers</summary>

1.  No — ATAC metric distributions shift with tissue, depth, and protocol version. Use the vignette thresholds as a **starting point** and always inspect the violins and density plots first.
2.  Direct analogue doesn't exist (ATAC doesn't measure RNA). The quality proxies are **TSS enrichment**, **nucleosome signal**, **blacklist ratio**, and **pct_reads_in_peaks** — each captures a different failure mode.

</details>
:::

## Step 7 — TF-IDF normalization, top features, SVD

```{r}
#| label: lsi

pbmc <- RunTFIDF(pbmc)
pbmc <- FindTopFeatures(pbmc, min.cutoff = "q0")
pbmc <- RunSVD(pbmc)

DepthCor(pbmc)
```

::: {.callout-important title="Think about it"}
1.  RNA uses `LogNormalize`. ATAC uses **TF-IDF**. Why the different normalization?
2.  Why is the **first LSI component often discarded** (see `DepthCor`)?

<details>

<summary>Show answers</summary>

1.  ATAC data is **binary-ish** — most peaks are either accessible (1) or not (0) per cell. TF-IDF (term-frequency × inverse document-frequency) weighs peaks by rarity, which is well-suited to sparse binary data. LogNormalize is tuned for count RNA data.
2.  The first SVD component typically tracks **sequencing depth / library size** rather than biology. `DepthCor` shows correlation of each LSI component with depth; if component 1 is highly correlated, skip it and use dims 2:30 downstream.

</details>
:::

## Step 8 — Non-linear embedding and clustering

```{r}
#| label: umap-cluster

pbmc <- RunUMAP(pbmc, reduction = "lsi", dims = 2:30)
pbmc <- FindNeighbors(pbmc, reduction = "lsi", dims = 2:30)
pbmc <- FindClusters(pbmc, algorithm = 3)   # SLM

DimPlot(pbmc, label = TRUE) + NoLegend()
```

::: {.callout-important title="Think about it"}
1.  Why `dims = 2:30` instead of `1:30`?
2.  `algorithm = 3` selects the SLM (Smart Local Moving) algorithm. Name one case when you'd use it instead of the default Louvain.

<details>

<summary>Show answers</summary>

1.  Because LSI component 1 is the depth-confounded one we saw in `DepthCor`. Starting at `2` keeps only biologically-meaningful dimensions.
2.  SLM often produces more stable partitions than basic Louvain on large, noisy graphs (e.g. scATAC) because it allows within-cluster re-optimization. Leiden (algorithm 4) is usually the modern default; SLM remains a reasonable alternative.

</details>
:::

## Save the object

```{r}
#| label: save

saveRDS(pbmc, file = "../data/pbmc_atac_clustered.rds")
```

## Stretch goals

-   **Gene activity** — convert peaks to per-gene accessibility scores with `GeneActivity()`, then run a Seurat-style RNA-like analysis on top
-   **Motif enrichment** — use `chromVAR` or Signac's `RunChromVAR` for TF-motif activity per cell
-   **Multi-modal integration** — integrate scRNA-seq + scATAC-seq from the same cells (multiome) using `FindTransferAnchors`

## Credits

-   Reference vignette: [Signac PBMC scATAC-seq vignette](https://stuartlab.org/signac/articles/pbmc_vignette)
-   Workflow source: Khushbu Patel, single-cell ATAC-seq standard workflow
-   Data: 10x Genomics PBMC 10k scATAC-seq
