---
title: "Tutorial 06 — Pseudobulk DE with DESeq2"
subtitle: "Aggregate to (sample × cell type), then run a proper bulk-style DE test on STIM vs CTRL"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

Paired with **[Lecture 04 — Pseudobulk DE & DESeq2](../Lecture_Folder/Lecture_04_Pseudobulk_DE.html)**. If you have never run DESeq2 before, work through the **[DESeq2 Primer (Tutorial 05)](Tutorial_05_Primer_DESeq2_Basics.html)** first — it covers the same mechanics on the `airway` bulk dataset.

We start from the **integrated, annotated object saved at the end of Tutorial 04** (`ifnb_integrated.rds`) and learn to:

-   Aggregate single-cell counts to a *(donor × condition × cell type)* pseudobulk matrix
-   Run a proper, well-calibrated DE test with `DESeq2`, one cell type at a time
-   Apply LFC shrinkage for ranking and visualize results
-   Avoid the most common pseudobulk pitfalls

::: callout-note
**The contrast is real this time.** The `ifnb` dataset has a built-in STIM (IFN-β stimulated) vs CTRL (control) contrast. In Tutorial 04 we **integrated** to remove the STIM-vs-CTRL axis from the embedding so cell types co-cluster. Here we **measure** the STIM-vs-CTRL effect *within each cell type* — the biology that integration was careful **not** to erase.
:::

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it. To run it:

1.  Download the `.qmd` source: <a href="Tutorial_06_Pseudobulk_DE.qmd.txt" download="Tutorial_06_Pseudobulk_DE.qmd">Tutorial_06_Pseudobulk_DE.qmd</a>. If your browser saves the file as `Tutorial_06_Pseudobulk_DE.qmd.txt`, **drop the trailing `.txt`** so the filename ends in `.qmd`, then open it in RStudio.
2.  Make sure you've completed **[Tutorial 04](Tutorial_04_Integration.html)** — it writes `ifnb_integrated.rds`.
3.  Work through the chunks, flipping `eval: true` as you go.
:::

## Dataset — resuming from Tutorial 04

| File | Produced by | What it is |
|---|---|---|
| `../data/ifnb_integrated.rds` | Tutorial 04 | Seurat object with integrated PCA / UMAP / clusters, the original `RNA` assay restored as default, and `seurat_annotations` cell-type labels per cell. |

### About donor information

The Kang *et al.* (2017) study pooled cells from **eight lupus donors**, but the SeuratData distribution of `ifnb` does not always carry per-cell donor IDs. Pseudobulk DE needs **biological replicates within each condition**, so we'll handle this in two ways depending on what's actually in your object:

-   **If `donor_id` (or `ind`) is present** — we'll use the eight real donors as biological replicates. `~ donor + stim` becomes a clean DESeq2 design with 8 vs 8.
-   **If donor is absent** — we'll create **synthetic pseudo-donors** by random cell assignment within each (`stim`, `seurat_annotations`) combination. Document this as a teaching shortcut, **not** something you'd ever do in a paper. With four pseudo-donors per condition you still get a 4-vs-4 contrast that DESeq2 can fit, and you'll see the workflow correctly even though the resulting p-values aren't biologically meaningful.

## Setup

```{r}
#| label: setup

library(Seurat)
library(DESeq2)
library(tidyverse)
library(patchwork)

set.seed(2026)

ifnb <- readRDS("../data/ifnb_integrated.rds")
DefaultAssay(ifnb) <- "RNA"
ifnb
```

```{r}
#| label: peek-meta

# Look for a real donor column
candidate_cols <- intersect(c("donor_id","donor","ind"), colnames(ifnb@meta.data))
candidate_cols
```

## Step 1 — Define donor (real or synthetic) and condition

```{r}
#| label: donor

if (length(candidate_cols) > 0) {
  ifnb$donor <- ifnb@meta.data[[ candidate_cols[1] ]]
  message("Using real donor column: ", candidate_cols[1])
} else {
  # Synthetic pseudo-donors: 4 per stim, assigned by random cell index
  set.seed(2026)
  ifnb$donor <- paste0(
    ifnb$stim, "_d",
    sample(1:4, size = ncol(ifnb), replace = TRUE)
  )
  message("Created synthetic pseudo-donors (teaching shortcut). ",
          "DO NOT do this for a real publication.")
}

ifnb$condition <- ifnb$stim   # CTRL or STIM
ifnb$celltype  <- ifnb$seurat_annotations

table(ifnb$donor, ifnb$condition)
```

::: {.callout-important title="Think about it"}
Why is it not enough to just run `FindMarkers(ifnb, ident.1 = "STIM", ident.2 = "CTRL")` on the per-cell expression matrix?

<details>

<summary>Click for answer</summary>

Because cells from one donor are **not statistically independent**. Running a per-cell test treats every cell as a separate replicate, which inflates the effective sample size and produces **anti-conservative** p-values — you'll get hundreds of "significant" genes that wouldn't replicate. The number of independent biological replicates is the **number of donors** (or pseudo-donors), not the number of cells. Pseudobulk fixes this by aggregating to one observation per (donor × condition × cell type).

For a thorough demonstration of how badly per-cell DE fails on this dataset specifically, see Squair *et al.* 2021 (Nat Commun, [doi](https://doi.org/10.1038/s41467-021-25960-2)).

</details>
:::

## Step 2 — Aggregate to pseudobulk

```{r}
#| label: aggregate

pb <- AggregateExpression(
  ifnb,
  assays    = "RNA",
  slot      = "counts",
  group.by  = c("donor", "condition", "celltype"),
  return.seurat = FALSE
)$RNA

dim(pb)               # genes × (donor × condition × celltype) replicates
head(colnames(pb), 6)
```

The column names look like `<donor>_<condition>_<celltype>`. Build the matching `colData`:

```{r}
#| label: colData

# Seurat's AggregateExpression replaces "_" with "-" by default in some versions.
# Be defensive and parse from the metadata directly instead.
group_meta <- ifnb@meta.data |>
  distinct(donor, condition, celltype) |>
  mutate(group_id = paste(donor, condition, celltype, sep = "_"))

# Match the colnames that AggregateExpression actually used. Different
# Seurat versions differ; this should align them.
fix_id <- function(x) gsub("[ /]", "-", x)
group_meta$group_id_clean <- fix_id(group_meta$group_id)
colnames(pb) <- fix_id(colnames(pb))

meta_pb <- tibble(group_id_clean = colnames(pb)) |>
  left_join(group_meta, by = "group_id_clean")

head(meta_pb)
```

## Step 3 — Filter low-count groups

A pseudobulk column built from very few cells is noise. Drop them.

```{r}
#| label: filter-pb

cells_per_group <- ifnb@meta.data |>
  count(donor, condition, celltype) |>
  mutate(group_id_clean = fix_id(paste(donor, condition, celltype, sep = "_")))

meta_pb <- meta_pb |>
  left_join(cells_per_group |> select(group_id_clean, n_cells = n),
            by = "group_id_clean")

# Keep groups with at least 10 cells
keep    <- !is.na(meta_pb$n_cells) & meta_pb$n_cells >= 10
pb      <- pb[, keep]
meta_pb <- meta_pb[keep, ]

table(meta_pb$celltype, meta_pb$condition)
```

::: {.callout-important title="Think about it"}
Why is `n_cells >= 10` reasonable? Why might you raise or lower it?

<details>

<summary>Click for answer</summary>

The threshold balances signal against noise: pseudobulk built from 1–2 cells is dominated by stochastic dropout and is unlikely to represent the cell type. A floor of \~10 cells is the de-facto default in the field (see Squair *et al.* 2021 and Crowell *et al.* 2020). Raise it (e.g. 20–50) for very sparse cell types or shallow libraries; lower it only if you also lower your statistical-significance expectations.

</details>
:::

## Step 4 — Run DESeq2 per cell type

```{r}
#| label: run-deseq

run_de <- function(ct) {
  keep <- meta_pb$celltype == ct
  if (sum(keep) < 4 || length(unique(meta_pb$condition[keep])) < 2) return(NULL)

  dds <- DESeqDataSetFromMatrix(
    countData = pb[, keep],
    colData   = as.data.frame(meta_pb[keep, ]),
    design    = ~ condition
  )
  dds <- DESeq(dds, quiet = TRUE)

  res <- results(dds, contrast = c("condition","STIM","CTRL"))
  res <- lfcShrink(dds, coef = "condition_STIM_vs_CTRL",
                   res = res, type = "apeglm")

  as.data.frame(res) |>
    rownames_to_column("gene") |>
    mutate(celltype = ct)
}

de_all <- map_dfr(unique(meta_pb$celltype), run_de) |>
  filter(!is.na(padj))
glimpse(de_all)
```

## Step 5 — Inspect the top hits

```{r}
#| label: top-hits

de_top <- de_all |>
  group_by(celltype) |>
  slice_min(padj, n = 10) |>
  ungroup()
de_top
```

::: {.callout-important title="Think about it"}
1.  Many of the top hits per cell type will be **interferon-stimulated genes** (`ISG15`, `IFI6`, `IFIT1/2/3`, `MX1`, `OAS1`, `RSAD2`). Are they up in STIM or CTRL?
2.  Are **the same** ISGs at the top of every cell type, or do you see cell-type-specific responses?

<details>

<summary>Click for answer</summary>

1.  Up in STIM — the IFN-β response is what STIM is, mechanistically. `log2FoldChange > 0` means higher in STIM (because we set `contrast = c("condition","STIM","CTRL")` with STIM as the numerator).
2.  A core ISG cluster (`ISG15`, `IFI6`, `MX1`) tops every cell type, **but the magnitude varies**. Monocytes (CD14, CD16) typically show the **strongest** ISG induction; T and B cells respond more modestly; some rare populations may be near-flat. This cell-type-specific quantitative difference is exactly the kind of biology pseudobulk DE is designed to find — and the kind of biology that the per-cell `FindMarkers` test cannot resolve confidently.

</details>
:::

## Step 6 — Visualize

A simple per-cell-type volcano plot:

```{r}
#| label: volcano

# Pick three cell types to plot; adjust to whichever yours has
plot_types <- c("CD14 Mono", "CD4 Naive T", "B")
plot_types <- intersect(plot_types, unique(de_all$celltype))

de_all |>
  filter(celltype %in% plot_types) |>
  ggplot(aes(log2FoldChange, -log10(padj))) +
  geom_point(alpha = 0.4, size = 1) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  geom_vline(xintercept = c(-1, 1), linetype = 3, color = "grey50") +
  facet_wrap(~ celltype) +
  labs(title = "Pseudobulk DE: STIM vs CTRL (log2FC > 0 = up in STIM)",
       x = "log2 FC (shrunken)", y = "-log10(padj)")
```

::: {.callout-important title="Think about it"}
Some cell types will show many DE genes; others almost none. What can drive this pattern *that has nothing to do with biology*?

<details>

<summary>Click for answer</summary>

Three usual suspects:

1.  **Cell counts per group.** Cell types that are abundant in every donor have well-estimated dispersions; rare types do not.
2.  **Sample size.** With 4 pseudo-donors split 4-vs-4 (or 8-vs-8 with real donors), you have moderate power on a clean signal — but very little if a cell type is missing in some donors.
3.  **Library composition.** A cell type whose library is dominated by a few hyper-expressed genes will have noisier dispersion estimates and fewer detected hits.

Always inspect `n_cells` per group and the dispersion plot before reporting "X genes are significantly DE".

</details>
:::

## Step 7 — Save results

```{r}
#| label: save

write_csv(de_all,  "../data/ifnb_pseudobulk_de.csv")
write_csv(de_top,  "../data/ifnb_pseudobulk_de_top.csv")
```

## Wrap-up

You now have:

-   A pseudobulk matrix per *(donor, condition, cell type)*.
-   Per-cell-type DE results from `DESeq2` with shrunken LFCs, on a real biological contrast (STIM vs CTRL).
-   A workflow you can re-point at *any* condition contrast in your own data — replace `condition = stim` with your own covariate and you're done.

::: {.callout-tip title="Going further"}
-   Add covariates (donor, sex, age) to the design: `~ donor + condition` (when donors are real).
-   Use `IHW` or `ashr` for adaptive multiple-testing.
-   Hand the gene lists to `clusterProfiler::compareCluster` for cross-cell-type enrichment.
-   Compare against `muscat::pbDS` (formal pseudobulk wrapper) and `glmGamPoi`.
:::

## See also

-   [Lecture 04 — Pseudobulk DE & DESeq2](../Lecture_Folder/Lecture_04_Pseudobulk_DE.html)
-   [Tutorial 05 — DESeq2 Primer on `airway`](Tutorial_05_Primer_DESeq2_Basics.html) — the bulk-DESeq2 walkthrough that this tutorial assumes
-   Squair *et al.* 2021 — the paper that made pseudobulk the default ([doi](https://doi.org/10.1038/s41467-021-25960-2))
-   Crowell *et al.* 2020 — `muscat` package paper ([doi](https://doi.org/10.1038/s41467-020-19894-4))
