---
title: "Tutorial 03 — scRNA-seq: Markers & Cell Type Annotation"
subtitle: "Find cluster markers, visualize, label by hand and check against ground truth"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

Part three of the series. We start from the **unintegrated clustered object saved in Tutorial 02** and walk through:


::: callout-note
**Companion book chapter:** [Chapter 3 — Markers & Manual Annotation](../Resources_Folder/Chapter_03_Markers_Annotation.html) — the long-form prose treatment of this tutorial's material, with cross-references to the prerequisite appendices. · **HPC version (Talapas):** [Talapas analysis pipeline — `03_markers_annotation.R`](Talapas/Tutorial_Talapas_03_Analysis_Pipeline.html)
:::


-   Finding marker genes per cluster
-   The canonical marker visualizations (Vln / Feature / Heatmap / Ridge / Dot)
-   Manual cell-type annotation with `RenameIdents`
-   Cross-checking your manual labels against the **ground-truth `seurat_annotations`** that ship with `ifnb`
-   Saving the annotated object so Tutorial 04 (Reference Annotation) can pick it up

::: callout-note
We're annotating the **unintegrated** object. Why? Because **markers are cell-type-driven, not batch-driven** — even when an embedding is dominated by a batch effect, the genes that distinguish T cells from B cells are still T-cell genes. Working markers up before integration is a useful sanity check: if a marker gene cleanly identifies a cluster on the unintegrated data, it should still cleanly identify the *biologically equivalent* cluster after integration in Tutorial 05.
:::

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it. To run it:

1.  Download the `.qmd` source: <a href="Tutorial_03_Markers_Annotation.qmd.txt" download="Tutorial_03_Markers_Annotation.qmd">Tutorial_03_Markers_Annotation.qmd</a>. If your browser saves the file as `Tutorial_03_Markers_Annotation.qmd.txt`, **drop the trailing `.txt`** so the filename ends in `.qmd`, then open it in RStudio.
2.  Make sure you have completed **[Tutorial 02](Tutorial_02_DimReduction_Clustering.html)** — it writes the `ifnb_clustered.rds` that this tutorial loads.
3.  Work through the chunks, flipping `eval: true` as you go.
:::

## Dataset — resuming from Tutorial 02

| File | Produced by | What it is |
|---|---|---|
| `../data/ifnb_clustered.rds` | Tutorial 02 | Seurat object with PCA, SNN graph, clusters at five resolutions, UMAP, plus the `stim` and `seurat_annotations` metadata. |

## Learning goals

-   Use `FindMarkers` / `FindAllMarkers` appropriately
-   Read each of the five standard marker-visualization plots and know when to pick which
-   Assign cell-type labels manually using PBMC canonical markers
-   Validate against the `seurat_annotations` ground truth and quantify agreement


::: {.callout-warning title="Common errors / things that bite"}
**`FindAllMarkers` takes 5+ minutes** — install `presto` (`install.packages("presto")`); Seurat detects it automatically and is \~10× faster.

**Markers list is dominated by ISGs (`ISG15`, `IFI6`, `IFIT3`, `MX1`)** — that's expected on the *unintegrated* `ifnb` data because clusters here are partly driven by `stim`. Tutorial 05 (Integration) fixes this; cell-type-specific markers reclaim the top of the lists after integration.

**`Idents()` is set to the wrong column** — clusters at multiple resolutions all live in `meta.data` (`RNA_snn_res.0.1`, `_0.3`, etc.) but only one is the "active" identity. `Idents(seu) <- "RNA_snn_res.0.5"` switches; `levels(Idents(seu))` confirms.

**`FindMarkers` errors with "data layers are not joined. Please run JoinLayers"** — Seurat v5 keeps per-sample split layers (`counts.CTRL`, `counts.STIM`, `data.CTRL`, `data.STIM`) after `merge()`, and `FindMarkers` refuses to run until they're collapsed. Fix with `ifnb <- JoinLayers(ifnb)` before calling `FindMarkers` / `FindAllMarkers`. The resume chunk above does this automatically. `Layers(ifnb)` shows the current layer list.
:::


::: {.callout-tip title="Solutions / instructor copy"}
The *Think about it* prompts on this page have inline answers (click "Show answer" / "Click for answer" to expand). For the full **runnable instructor copy** with every chunk pre-evaluated and outputs shown:

- **For students:** the same `.qmd` you downloaded already contains every solution inline — just expand the collapsed answers
- **For instructors:** render with `quarto render --profile solutions` from the `Exercise_Folder/` directory. The fully-evaluated HTML is written to `docs/Exercise_Folder/_solutions/Tutorial_03_Markers_Annotation.html`. See [Exercise_Folder/_quarto-solutions.yml](https://github.com/wcresko/scRNAseq_tutorial/blob/main/Exercise_Folder/_quarto-solutions.yml) for the build profile

If you're running a graded course, link the rendered solutions page from your LMS (Canvas / Moodle / etc.) — it's behind the same `_solutions/` path on the workshop's deployed site.
:::

## Setup

```{r}
#| label: setup

library(Seurat)
library(tidyverse)
library(patchwork)   # combine + annotate multi-panel Seurat plots

set.seed(2026)

# ---------------------------------------------------------------------------
# Output directory for this module's figures and tables.
# Every figure/table chunk below writes a file named  Mod3_C<chunk>_<name>
# into ../output/Mod3/ so it can be cross-referenced from the rest of the site.
#   Mod3 = Module 3 (this tutorial);  C<n> = the nth code chunk.
# ---------------------------------------------------------------------------
out_dir <- "../output/Mod3"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: resume

ifnb <- readRDS("../data/ifnb_clustered.rds")
Idents(ifnb) <- "RNA_snn_res.0.5"

# In Seurat v5, merging per-sample objects (as Tutorial 01 does after
# scDblFinder) leaves the RNA assay with split layers. FindMarkers requires
# a single joined layer, so rejoin here if it hasn't been done already.
if (length(Layers(ifnb, search = "counts")) > 1) {
  ifnb <- JoinLayers(ifnb)
}

ifnb
```

------------------------------------------------------------------------

## Step 1 — Find markers for one cluster

```{r}
#| label: markers-one

cluster1.markers <- FindMarkers(ifnb, ident.1 = 1, min.pct = 0.25)
head(cluster1.markers, 10)

# --- Table out: top markers for cluster 1 (gene moved from rownames) -------
cluster1.markers |>
  tibble::rownames_to_column("gene") |>
  readr::write_csv(file.path(out_dir, "Mod3_C3_cluster1_markers.csv"))
```

::: {.callout-important title="Think about it"}
1.  What does `min.pct = 0.25` do, and why is that a useful default?
2.  The output has `avg_log2FC`, `pct.1`, `pct.2`. Which gene would you pick as a *specific* marker for this cluster?

<details>

<summary>Show answers</summary>

1.  Only tests genes detected in ≥ 25% of cells in at least one of the two groups. Prunes low-expression noise and speeds the test up significantly.
2.  A specific marker has **high `pct.1` and low `pct.2`** (expressed in almost all cluster-1 cells, almost none elsewhere) *and* high positive `avg_log2FC`. Markers that are broadly expressed (high `pct.2`) may be real but aren't *specific*.

</details>
:::

## Step 2 — Markers for every cluster

```{r}
#| label: markers-all

ifnb.markers <- FindAllMarkers(
  ifnb,
  only.pos        = TRUE,
  min.pct         = 0.25,
  logfc.threshold = 0.25
)

top_markers <- ifnb.markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 5)
top_markers

# --- Tables out: all positive markers + the per-cluster top 5 --------------
readr::write_csv(ifnb.markers, file.path(out_dir, "Mod3_C4_all_markers.csv"))
readr::write_csv(top_markers,  file.path(out_dir, "Mod3_C4_top5_markers_per_cluster.csv"))
```

::: {.callout-important title="Think about it"}
1.  Many of the top markers will be **interferon-stimulated genes** (`ISG15`, `IFI6`, `IFIT3`, `MX1`). Are those *cell-type* markers, or are they marking something else?
2.  For condition-level DE (STIM vs CTRL), is `FindAllMarkers` the right tool?

<details>

<summary>Show answers</summary>

1.  Those are **condition markers** — they distinguish IFN-β–stimulated cells from control cells, *not* cell types. They top the list because cluster identities at this stage are partly driven by `stim`. Tutorial 05 will integrate the data so cluster identities track biology, not condition.
2.  No. `FindAllMarkers` / `FindMarkers` treat each cell as an independent observation, which is **anti-conservative** for condition-level comparisons across biological samples. For STIM vs CTRL within a cell type, use **pseudobulk** + DESeq2/edgeR/limma — that's exactly what Tutorial 06 covers, on this dataset.

</details>
:::

## Step 3 — Visualizing markers

PBMC canonical markers — useful to keep handy:

| Cell type | Canonical markers |
|---|---|
| CD14+ Monocyte | `CD14`, `LYZ`, `S100A8`, `S100A9` |
| CD16+ Monocyte | `FCGR3A`, `MS4A7`, `CDKN1C` |
| Dendritic cell (DC) | `FCER1A`, `CST3`, `CLEC10A` |
| pDC | `LILRA4`, `IL3RA`, `CLEC4C` |
| CD4 Naive T | `IL7R`, `CCR7`, `LEF1` |
| CD4 Memory T | `IL7R`, `S100A4` |
| CD8 T | `CD8A`, `CD8B`, `GZMK` |
| NK | `GNLY`, `NKG7`, `KLRD1` |
| B | `MS4A1`, `CD79A`, `CD79B` |
| T-activated | `CD69`, `IL2RA` |
| Mk (megakaryocyte) | `PPBP`, `PF4` |
| Eryth | `HBA1`, `HBB` |

### 3a. Violin plots — expression distribution per cluster

```{r}
#| label: viz-vln
#| fig-cap: "Violin plots of six canonical PBMC marker genes across clusters (resolution 0.5)."

p_vln <- VlnPlot(ifnb,
        features = c("CD14", "LYZ", "MS4A1", "CD79A", "CD8A", "GNLY"),
        pt.size = 0, ncol = 3) &
  xlab("Cluster (RNA_snn_res.0.5)") &
  ylab("Log-normalized expression") &
  theme(plot.title = element_text(size = 11))
p_vln <- p_vln +
  patchwork::plot_annotation(
    title    = "Canonical PBMC marker expression by cluster",
    subtitle = "Monocyte (CD14/LYZ), B (MS4A1/CD79A), CD8 T (CD8A) and NK (GNLY) markers",
    caption  = "Module 3 · Markers & Annotation"
  )
p_vln

# --- Figure out ------------------------------------------------------------
ggsave(file.path(out_dir, "Mod3_C5_marker_violins.png"), p_vln,
       width = 10, height = 6, dpi = 300)
```

### 3b. Feature plots — expression on UMAP

```{r}
#| label: viz-feature
#| fig-cap: "UMAP feature plots showing where four canonical markers (CD14, MS4A1, CD8A, GNLY) are expressed."

p_feature <- FeaturePlot(ifnb,
            features = c("CD14", "MS4A1", "CD8A", "GNLY"),
            ncol = 2) &
  xlab("UMAP 1") &
  ylab("UMAP 2") &
  theme(plot.title = element_text(size = 11))
p_feature <- p_feature +
  patchwork::plot_annotation(
    title    = "Marker expression overlaid on the UMAP",
    subtitle = "CD14 (monocytes), MS4A1 (B), CD8A (CD8 T), GNLY (NK)",
    caption  = "Module 3 · Markers & Annotation"
  )
p_feature

# --- Figure out ------------------------------------------------------------
ggsave(file.path(out_dir, "Mod3_C6_marker_featureplots.png"), p_feature,
       width = 9, height = 8, dpi = 300)
```

### 3c. Heatmap — top markers across clusters

```{r}
#| label: viz-heatmap
#| fig-cap: "Heatmap of the top-10 markers per cluster (z-scored expression), cells grouped by cluster."

top10 <- ifnb.markers |> group_by(cluster) |> slice_max(avg_log2FC, n = 10)

# Tutorial 01 only scaled the HVGs; re-scale the top-marker genes so any
# non-HVG markers also appear in the heatmap rather than being dropped.
ifnb <- ScaleData(ifnb, features = unique(top10$gene))

p_heatmap <- DoHeatmap(ifnb, features = top10$gene) +
  NoLegend() +
  labs(
    title    = "Top-10 markers per cluster — scaled expression heatmap",
    subtitle = "Each column is a cell, grouped by cluster; rows are marker genes",
    caption  = "Module 3 · Markers & Annotation"
  )
p_heatmap

# --- Figure out ------------------------------------------------------------
ggsave(file.path(out_dir, "Mod3_C7_marker_heatmap.png"), p_heatmap,
       width = 12, height = 10, dpi = 300)
```

### 3d. Dot plot — % expressing × mean expression

```{r}
#| label: viz-dot
#| fig-cap: "Dot plot of a 16-gene canonical marker panel across clusters — dot size is percent expressing, colour is scaled mean expression."

canonical_markers <- c("CD14", "LYZ", "FCGR3A", "MS4A7", "FCER1A", "CST3",
                       "IL7R", "CCR7", "S100A4", "CD8A", "GZMK", "GNLY",
                       "NKG7", "MS4A1", "CD79A", "PPBP")
p_dot <- DotPlot(ifnb, features = canonical_markers) +
  RotatedAxis() +
  labs(
    title    = "Canonical PBMC marker panel across clusters",
    subtitle = "Dot size = % of cells expressing; colour = scaled average expression",
    x        = "Marker gene",
    y        = "Cluster (RNA_snn_res.0.5)",
    colour   = "Avg. expression",
    size     = "Percent expressed"
  )
p_dot

# --- Figure out ------------------------------------------------------------
ggsave(file.path(out_dir, "Mod3_C8_marker_dotplot.png"), p_dot,
       width = 10, height = 6, dpi = 300)
```

::: {.callout-important title="Think about it"}
1.  Which plot is best for **showing a 16-gene marker panel across 12 clusters on one page**?
2.  Which is best for **checking whether a single gene has spatial structure on the UMAP**?
3.  Which is most misleading if your cluster sizes are very unequal, and why?

<details>

<summary>Show answers</summary>

1.  **Dot plot** — compact, shows both percent-expressing (dot size) and mean expression (color) in a genes × clusters grid.
2.  **Feature plot** — overlays expression on the 2-D embedding so you can see geography.
3.  **Heatmap** can be misleading: it shows cells, not clusters, so a dominant cluster (say 40% of cells) visually monopolizes the plot. Dot plot normalizes per cluster and avoids this.

</details>
:::

## Step 4 — Manual annotation

Use the canonical marker table above and the dot plot to assign labels. Your cluster numbering will depend on the random seed and Seurat version, so **do not copy these labels blindly** — pick from the table that matches *your* dot plot.

```{r}
#| label: rename

# EXAMPLE — replace each entry below with the label appropriate for YOUR
# clusters. The order must match levels(Idents(ifnb)) at this point.
#
# new.cluster.ids <- c(
#   "0" = "CD14 Mono",
#   "1" = "CD4 Naive T",
#   "2" = "CD4 Memory T",
#   "3" = "B",
#   "4" = "CD8 T",
#   "5" = "NK",
#   "6" = "CD16 Mono",
#   "7" = "T activated",
#   "8" = "DC",
#   "9" = "Mk",
#   "10" = "B activated",
#   "11" = "pDC",
#   "12" = "Eryth"
# )
#
# names(new.cluster.ids) <- levels(ifnb)
# ifnb <- RenameIdents(ifnb, new.cluster.ids)
# ifnb$celltype_manual <- Idents(ifnb)
#
# DimPlot(ifnb, reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()
```

::: {.callout-important title="Think about it"}
1.  Two clusters share `CD3D` / `CD3E` as top markers. Are they the same cell type?
2.  What's the risk of assigning a label from a single marker gene?

<details>

<summary>Show answers</summary>

1.  Not necessarily — `CD3D/E` are pan-T markers. Both clusters are likely T cells, but could be CD4 Naive vs CD4 Memory vs CD8 vs T-activated. You need *discriminating* markers (`IL7R`+`CCR7` for naive, `S100A4` for memory, `CD8A` for cytotoxic, `IL2RA` for activated) on top of the shared T-cell markers.
2.  Many markers are non-specific, and biological "marker genes" are often graded rather than binary. Label from **combinations** (a marker *plus* absence of another marker) and always cross-check.

</details>
:::

## Step 5 — Validate against the ground-truth labels

The `ifnb` dataset ships with author-curated labels in `seurat_annotations`. This is rare in real life — but it makes a beautiful teaching moment to compare your manual labels to a published ground truth.

```{r}
#| label: validate
#| fig-cap: "Side-by-side UMAPs comparing your manual cluster annotation against the author-curated ground-truth labels that ship with ifnb."

# Side-by-side UMAPs
p_manual <- DimPlot(ifnb, reduction = "umap", label = TRUE, repel = TRUE) +
  NoLegend() +
  labs(
    title  = "Manual annotation (yours)",
    x      = "UMAP 1",
    y      = "UMAP 2"
  )
p_truth  <- DimPlot(ifnb, reduction = "umap", group.by = "seurat_annotations",
                    label = TRUE, repel = TRUE) +
  NoLegend() +
  labs(
    title  = "Author labels (ground truth)",
    x      = "UMAP 1",
    y      = "UMAP 2"
  )

p_manual
p_truth

# Combined two-panel manual-vs-truth comparison, shared annotation
p_validate <- (p_manual | p_truth) +
  patchwork::plot_annotation(
    title    = "Manual annotation vs author ground truth on the same embedding",
    subtitle = "Cells should land in matching positions if your manual labels agree with the authors'",
    caption  = "Module 3 · Markers & Annotation"
  )

# Cross-tabulate. Each row should ideally concentrate on a single column.
manual_vs_truth <- table(manual = Idents(ifnb), truth = ifnb$seurat_annotations)
manual_vs_truth

# --- Figures out -----------------------------------------------------------
ggsave(file.path(out_dir, "Mod3_C10_umap_manual.png"),         p_manual,
       width = 7,  height = 6, dpi = 300)
ggsave(file.path(out_dir, "Mod3_C10_umap_ground_truth.png"),   p_truth,
       width = 7,  height = 6, dpi = 300)
ggsave(file.path(out_dir, "Mod3_C10_umap_manual_vs_truth.png"), p_validate,
       width = 13, height = 6, dpi = 300)

# --- Table out: cross-tab of manual labels vs ground-truth labels ----------
as.data.frame(manual_vs_truth) |>
  readr::write_csv(file.path(out_dir, "Mod3_C10_manual_vs_truth.csv"))
```

::: {.callout-important title="Think about it"}
1.  Where do you and the authors **agree**? Where do you **disagree**?
2.  Which cell type was easiest to call from markers alone? Hardest?
3.  In Tutorial 05, after integration, the same cluster numbers will mean different things. Why is `seurat_annotations` (a per-cell label) more durable than your `RNA_snn_res.0.5` manual labels (a per-cluster mapping)?

<details>

<summary>Show answers</summary>

1.  Big myeloid (CD14 / CD16 / DC) and lymphoid (B / NK) clusters usually agree perfectly. T-cell subdivisions (CD4 Naive vs CD4 Memory vs T activated) and rare populations (pDC, Eryth, Mk) are the common disagreements — they require multi-gene logic and have soft boundaries.
2.  **Easiest:** B cells (`MS4A1`+, `CD79A`+, otherwise quiet) and NK (`GNLY`+, `NKG7`+, no `CD3`). **Hardest:** the various CD4 T subsets, because they share most markers and differ in subtle, graded ways.
3.  Per-cell labels travel with the cell — if a cell is `CD14 Mono` in `seurat_annotations` here, it stays `CD14 Mono` after integration too. Cluster numbers are recomputed every time you re-cluster (and integration recomputes), so *cluster-level* labels are tied to a specific clustering and have to be re-applied.

</details>
:::

## Step 6 — Save the annotated object

```{r}
#| label: save

saveRDS(ifnb, file = "../data/ifnb_annotated.rds")
```

## Continue to Tutorial 04 — Reference Annotation

You've named the clusters by hand. The immediate next step, **[Tutorial 04 — Reference-based Annotation](Tutorial_04_Reference_Annotation.html)**, cross-checks those manual labels against an automated reference (SingleR / Azimuth) on this same annotated object — confirming the confident calls and flagging the cells the reference can't place.

After that, **[Tutorial 05 — Two-Sample Integration](Tutorial_05_Integration.html)** re-does the PCA + neighbor-graph + clustering steps on an **integrated** representation that aligns CTRL and STIM, and you'll watch what happens to:

-   the UMAP (CTRL/STIM islands collapse onto each other)
-   the clusters (cell types stop splitting in two)
-   the markers (ISGs stop being top markers; cell-type markers reclaim the top)

That integrated `.rds` is then the input to **pseudobulk DE** in Tutorial 06.

## Credits

-   Dataset: `ifnb` from `SeuratData` — Kang *et al.* (2017) *Nat Biotechnol*
-   PBMC canonical markers: [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html), [Azimuth PBMC reference](https://azimuth.hubmapconsortium.org/references/human_pbmc/), [PanglaoDB](https://panglaodb.se/)
-   Inspiration: [scNotebooks](https://integrativebioinformatics.github.io/scNotebooks/)
