---
title: "Tutorial 01 — scRNA-seq: QC & Preprocessing"
subtitle: "Load a two-sample dataset (ifnb), filter cells, normalize, pick HVGs, scale"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

The first hands-on tutorial walking the standard **Seurat** scRNA-seq workflow on a **laptop-friendly** dataset. We use the **`ifnb`** dataset (Kang *et al.* 2017) — peripheral blood mononuclear cells (PBMCs) from eight lupus patients, split into a **control** sample and an **interferon-β-stimulated** sample. Two samples, ~14,000 cells, ~25 MB on disk, no 10x click-through.


::: callout-note
**Companion book chapter:** [Chapter 1 — Preprocessing](../Resources_Folder/Chapter_01_Preprocessing.html) — the long-form prose treatment of this tutorial's material, with cross-references to the prerequisite appendices. · **HPC version (Talapas):** [Talapas analysis pipeline — `01_qc_preprocessing.R`](Talapas/Tutorial_Talapas_03_Analysis_Pipeline.html)
:::


Why two samples from the start? Because Tutorial 05 will teach **integration** — and integration only makes sense if you've already loaded, QC-filtered, and clustered samples that come from different conditions / batches. We carry both `STIM` and `CTRL` cells through every step.

::: callout-tip
**Want to scale up to a bigger dataset later?** Once you've worked through the `ifnb` tutorials, the parallel **Talapas Advanced series** in `Exercise_Folder/Talapas/` re-do the same workflow on a 20k cell, 250 MB dataset. They are written to be run on **Talapas** as R + SLURM jobs — see [Tutorial — SLURM Basics](Talapas/Tutorial_Talapas_01_SLURM_Basics.html).
:::

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it (chunks default to `eval: false`). To actually run the analysis:

1.  Download the `.qmd` source: <a href="Tutorial_01_QC_Preprocessing.qmd.txt" download="Tutorial_01_QC_Preprocessing.qmd">Tutorial_01_QC_Preprocessing.qmd</a>. If your browser saves the file as `Tutorial_01_QC_Preprocessing.qmd.txt`, **drop the trailing `.txt`** so the filename ends in `.qmd`, then open it in RStudio.
2.  Install the `ifnb` dataset once (one R command — see [Setup](#setup)).
3.  Work through the chunks, flipping `eval: true` as you go.
:::

## Dataset — `ifnb` (Kang *et al.* 2017)

**What it is.** Peripheral blood mononuclear cells (PBMCs) from eight lupus patients. Each donor's PBMCs were split into two aliquots:

-   **`CTRL`** — control (no stimulation), \~6,500 cells
-   **`STIM`** — treated with interferon-β for 6 hours, \~7,500 cells

Together \~14,000 cells across two samples on the **10x Chromium 3' v1** assay. **Cell-type labels** (e.g. `CD14 Mono`, `CD4 Naive T`, `B`, `NK`, `DC`) are bundled in the metadata so you can sanity-check clustering and integration without doing your own annotation.

-   **Reference:** Kang *et al.* (2017) *Nat Biotechnol* 36: 89–94. [doi:10.1038/nbt.4042](https://doi.org/10.1038/nbt.4042)
-   **Origin:** GEO accession [GSE96583](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE96583)
-   **Distributed via:** the Bioconductor [`muscData`](https://bioconductor.org/packages/muscData/) package as `Kang18_8vs8()` — cached locally through `ExperimentHub` on first download
-   **Used in:** Tutorials 01 → 04 of this workshop, and the official [Seurat integration vignette](https://satijalab.org/seurat/articles/integration_introduction.html)

::: callout-note
**Why this dataset for teaching.** It's small enough to run on a laptop, it has a clean two-condition structure (so integration is meaningful), and the cell types are PBMCs — well-known, well-marked, and easy to reason about. The interferon response is a strong, biologically real signal that drives both batch-style separation (use case for integration) **and** condition-level differential expression (use case for pseudobulk DE in Tutorial 06).
:::

### Disk-space budget

| Item | Size |
|---|---|
| `ifnb` raw counts (compressed cache) | \~25 MB |
| `ifnb` Seurat object in memory | \~600 MB |
| `.rds` snapshots saved between tutorials | \~500 MB each |

Plan for **\~1.5 GB total** including intermediate `.rds` files for Tutorials 01–08.

### Data dictionary — what's in the loaded `ifnb` Seurat object

| Slot | Class | Contents |
|---|---|---|
| `ifnb@assays$RNA@counts` | `dgCMatrix` | Raw UMI counts. \~14,000 genes × \~14,000 cells (sparse). |
| `ifnb@meta.data$stim` | factor | `"CTRL"` or `"STIM"` — the **sample-of-origin** label. This becomes the integration covariate. |
| `ifnb@meta.data$seurat_annotations` | factor | Author-curated cell-type labels from the original paper. Useful as ground truth in Tutorial 03. |
| `ifnb@meta.data$nCount_RNA`, `nFeature_RNA` | numeric | Auto-populated by `CreateSeuratObject` — UMIs and unique genes per cell. |

## Learning goals

By the end of this tutorial you will be able to:

-   Fetch a published scRNA-seq dataset from Bioconductor's `muscData` package and assemble it as a Seurat object
-   Inspect a multi-sample Seurat object and split it by sample
-   Add per-cell QC metrics (`percent.mt`, `percent.rb`) and visualize them per sample
-   Choose filtering thresholds that are defensible for *your* dataset
-   Run `NormalizeData` → `FindVariableFeatures` → `ScaleData` and understand what each does
-   Save the QC'd object so Tutorial 02 can pick it up


::: {.callout-warning title="Common errors / things that bite"}
**`library(muscData)` errors with "there is no package called 'muscData'"** — `muscData` is a Bioconductor experiment-data package, not on CRAN. Install with `BiocManager::install("muscData")` (which itself needs `install.packages("BiocManager")` first). The historical `SeuratData::InstallData("ifnb")` route is no longer used here — the Satija lab's data server has been intermittently unavailable — and the `scRNAseq` package dropped its per-dataset accessor functions (e.g. `KangHumanPBMCData()`) in v2.18 in favor of a moving-target `fetchDataset()` API. `muscData::Kang18_8vs8()` bundles the same Kang *et al.* 2017 data with a stable interface.

**`scDblFinder` returns a different result than the workshop** — `scDblFinder` is stochastic. The workshop uses `set.seed(2026)` *immediately before* calling it; if you set the seed earlier or not at all, the doublet calls will differ. The 5–10% range is what to expect; exact cell IDs will vary.

**SoupX errors with "no raw matrix found"** — this is correct behaviour. The `Kang18_8vs8()` download only ships the *filtered* counts. SoupX needs both. The Step 5 callout flags this — the SoupX block is illustrative on `ifnb`. Run it for real on data that has a Cell Ranger `raw_feature_bc_matrix/` folder.
:::


::: {.callout-tip title="Solutions / instructor copy"}
The *Think about it* prompts on this page have inline answers (click "Show answer" / "Click for answer" to expand). For the full **runnable instructor copy** with every chunk pre-evaluated and outputs shown:

- **For students:** the same `.qmd` you downloaded already contains every solution inline — just expand the collapsed answers
- **For instructors:** render with `quarto render --profile solutions` from the `Exercise_Folder/` directory. The fully-evaluated HTML is written to `docs/Exercise_Folder/_solutions/Tutorial_01_QC_Preprocessing.html`. See [Exercise_Folder/_quarto-solutions.yml](https://github.com/wcresko/scRNAseq_tutorial/blob/main/Exercise_Folder/_quarto-solutions.yml) for the build profile

If you're running a graded course, link the rendered solutions page from your LMS (Canvas / Moodle / etc.) — it's behind the same `_solutions/` path on the workshop's deployed site.
:::

## Setup {#setup}

```{r}
#| label: setup

# One-time install (skip if already installed):
# install.packages(c("Seurat", "tidyverse", "BiocManager"))
# BiocManager::install(c("muscData", "SingleCellExperiment"))

library(Seurat)
library(muscData)
library(SingleCellExperiment)
library(tidyverse)

set.seed(2026)
```

::: callout-note
`muscData` is a **Bioconductor** experiment-data package (not on CRAN), so it has to go through `BiocManager::install()`. The first call to `Kang18_8vs8()` downloads \~25 MB through `ExperimentHub` into your local cache; subsequent calls are instant.

Earlier versions of this tutorial used `SeuratData::InstallData("ifnb")`. That route depends on the Satija lab's data server (`seurat.nygenome.org`), which has been intermittently unavailable. A brief detour through `scRNAseq::KangHumanPBMCData()` also broke because `scRNAseq` v2.18 removed its per-dataset accessor functions. `muscData::Kang18_8vs8()` exposes the same Kang *et al.* 2017 PBMCs through a stable interface.
:::

------------------------------------------------------------------------

## Step 1 — Load the `ifnb` dataset

```{r}
#| label: load

# Fetch the Kang et al. 2017 PBMC dataset (CTRL + IFN-beta STIM) from
# Bioconductor's ExperimentHub. The first call downloads ~25 MB and caches it;
# every subsequent call is a no-op.
sce <- Kang18_8vs8()

# Drop the multiplexed multiplets so we end up with the same set of singletons
# that the historical `ifnb` Seurat object shipped (~14k cells), and drop any
# cells without an author-assigned cell type label.
sce <- sce[, sce$multiplets == "singlet" & !is.na(sce$cell)]

# Convert to a Seurat object, mirroring the metadata column names the rest of
# this tutorial series expects:
#   - `stim` as a factor with levels c("CTRL", "STIM")
#   - `seurat_annotations` for the author-curated cell-type labels
ifnb <- CreateSeuratObject(
  counts    = counts(sce),
  meta.data = as.data.frame(colData(sce))
)
ifnb$stim <- factor(toupper(ifnb$stim), levels = c("CTRL", "STIM"))
ifnb$seurat_annotations <- factor(ifnb$cell)

ifnb

# What samples are in here?
table(ifnb$stim)

# Author-curated cell-type labels (we'll use these as ground truth in Tut 03)
table(ifnb$seurat_annotations)
```

::: {.callout-important title="Think about it"}
1.  Why do real integration tutorials almost always have at least two samples?
2.  What's the difference between a "sample" (`stim`) and a "cell type" (`seurat_annotations`) in this dataset?

<details>

<summary>Show answers</summary>

1.  Integration aligns the **shared biological signal** between samples while preserving any **real differences** between them. With one sample there's nothing to align — clustering on PCs alone is enough. The `ifnb` dataset has CTRL and STIM as two samples drawn from the same donors, which is the perfect minimal setup to teach the alignment step.
2.  `stim` is the **technical / experimental covariate** — which condition the cell came from. `seurat_annotations` is the **biological identity** — what kind of cell it is. Integration aims to remove `stim`-driven structure from the embedding *without* erasing `seurat_annotations`-driven structure.

</details>
:::

## Step 2 — Inspect dimensions, split by sample

The `ifnb` object already arrives as a single Seurat object containing both conditions. For a two-sample QC workflow, it's useful to look at QC metrics **per sample** rather than globally — that way you'll catch a sample-specific quality issue (e.g. one sample with much lower depth) before it confounds the analysis.

```{r}
#| label: split

# Record the starting matrix dimensions
dim_raw <- dim(ifnb)
dim_raw
```

::: {.callout-important title="Think about it"}
1.  The `ifnb` object holds both CTRL and STIM cells in a single counts matrix. What information distinguishes a CTRL barcode from a STIM barcode?
2.  Why might it matter to split visualizations by `stim` even though clustering happens on the joint matrix?

<details>

<summary>Show answers</summary>

1.  The `stim` column in `meta.data`. Each barcode also carries the `-1` / `-2` GEM-well suffix from Cell Ranger, but you should never rely on barcode suffixes for sample membership — the metadata is the source of truth.
2.  Per-sample plots reveal **batch-level QC issues** (lower median depth in one sample, different mitochondrial fractions, different cell counts per cluster) that can be hidden by a global histogram. If one sample's distribution is shifted, you may need different filter thresholds for it — or you may need to investigate before going further.

</details>
:::

## Step 3 — Add QC metrics

```{r}
#| label: qc-metrics

ifnb[["percent.mt"]] <- PercentageFeatureSet(ifnb, pattern = "^MT-")
ifnb[["percent.rb"]] <- PercentageFeatureSet(ifnb, pattern = "^RP[SL]")

head(ifnb@meta.data)
```

::: callout-note
The `ifnb` raw matrix uses HUGO gene symbols, so `^MT-` matches the 13 protein-coding mitochondrial genes (MT-ND1, MT-CO1, …) and `^RP[SL]` matches the ribosomal protein genes (RPS\*, RPL\*). For mouse data the conventions are lowercase: `^mt-` and `^Rp[sl]`.
:::

::: {.callout-important title="Think about it"}
1.  What biological signal does `percent.mt` capture?
2.  Why might `percent.rb` (ribosomal-protein fraction) be useful?

<details>

<summary>Show answers</summary>

1.  Dying or stressed cells leak cytoplasmic RNA while retaining mitochondrial RNA, so their mitochondrial fraction rises. High `percent.mt` flags apoptotic / compromised cells.
2.  Ribosomal-protein genes are **highly expressed** and **broadly variable** across cell states. A very low or very high `percent.rb` per cluster is sometimes a tell-tale sign of a metabolic / activation difference rather than a cell-type difference — useful to separate "this cluster is biologically distinct" from "this cluster is just translationally quiet."

</details>
:::

## Step 4 — Visualize QC, per sample

```{r}
#| label: qc-plots

VlnPlot(ifnb,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rb"),
        group.by = "stim",
        ncol = 4,
        pt.size = 0)

FeatureScatter(ifnb,
               feature1 = "nCount_RNA",
               feature2 = "nFeature_RNA",
               group.by = "stim") +
  geom_smooth(method = "lm")
```

::: {.callout-important title="Think about it"}
1.  Compare the violins between CTRL and STIM. Are the medians similar? Does either sample have a much fatter upper tail on `nFeature_RNA`?
2.  What would a doublet look like on the `nCount_RNA` vs `nFeature_RNA` scatter? A dying cell?

<details>

<summary>Show answers</summary>

1.  In `ifnb` the two samples are very similar in median QC metrics — that's a feature, not a bug, because the donors are matched. A noticeably fatter upper tail in one sample would suggest more **doublets** there (perhaps a higher loading concentration). A noticeably *lower* median `nFeature_RNA` would suggest sequencing depth or cell-quality differences.
2.  Doublets sit **high on both axes**, often above the main trend line. Dying cells are **low** on both and tend to have **high `percent.mt`** (color the scatter by `percent.mt` or facet by it).

</details>
:::

## Step 5 — Optional: ambient-RNA correction with SoupX

In a 10x droplet experiment, lysed-cell debris in the bulk solution leaks **ambient mRNA** into every droplet. After Cell Ranger, this shows up as a low background of "wrong" transcripts in every cell — most visibly when a cell expresses a marker for a *different* cell type at low level. **`SoupX`** estimates the ambient profile and corrects each cell's counts. It runs *before* QC filtering and replaces the counts matrix with a cleaned one.

::: callout-warning
**SoupX needs both the raw and filtered Cell Ranger matrices** (`raw_feature_bc_matrix/` and `filtered_feature_bc_matrix/`) because it learns the soup profile from empty droplets. `Kang18_8vs8()` only ships the **filtered** counts, so you cannot run SoupX on `ifnb` directly. The code below is the canonical pattern — use it on **your own** 10x data, or on the [Talapas analysis pipeline](Talapas/Tutorial_Talapas_03_Analysis_Pipeline.html) where the raw matrix is available.
:::

```{r}
#| label: soupx
#| eval: false

# Pattern for any 10x dataset where you have the raw Cell Ranger output:
# library(SoupX)
# library(DropletUtils)
#
# # 1. Point at the Cell Ranger 'outs/' directory (containing raw_feature_bc_matrix/
# #    and filtered_feature_bc_matrix/)
# sc <- load10X("path/to/cellranger/outs")
#
# # 2. (Recommended) provide preliminary clusters so SoupX can refine the soup
# #    profile per cluster. Compute these with a quick Seurat workflow on the
# #    filtered matrix first, then attach:
# # sc <- setClusters(sc, setNames(seu$seurat_clusters, colnames(seu)))
# # sc <- setDR(sc, Embeddings(seu, "umap"))
#
# # 3. Estimate the ambient contamination fraction
# sc <- autoEstCont(sc)
# rho <- sc$fit$rhoEst
# message("Estimated ambient fraction: ", round(rho, 3))
#
# # 4. Adjust the counts (round to integers so downstream UMI tools are happy)
# adj <- adjustCounts(sc, roundToInt = TRUE)
#
# # 5. Use `adj` (a sparse matrix) as the input to CreateSeuratObject() in Step 1
# #    instead of the raw filtered matrix. Everything else in this tutorial then
# #    runs on the cleaned counts.
```

::: {.callout-important title="Think about it"}
1.  Why does SoupX need the **raw** (pre-filter) matrix? What information does it pull from there?
2.  When is SoupX overkill? When is it essential?

<details>

<summary>Show answers</summary>

1.  The raw matrix contains droplets that captured **no real cell** — they hold only ambient RNA. SoupX uses those empty droplets to estimate the soup's expression profile, then subtracts a fraction of that profile from every called cell.
2.  **Overkill:** small contamination (rho ≲ 0.05), well-prepared samples, no marker bleed-through. **Essential:** complex tissues with high cell death / large cell-type differences (tumor microenvironment, mucosal tissues), or any time you see a marker like `HBB` or `IGKC` showing up faintly in cells that shouldn't express it.

</details>
:::

## Step 6 — Doublet detection with `scDblFinder`

A **doublet** is one barcode's worth of counts that came from two cells getting trapped in the same droplet. They show up as cells with anomalously high `nCount_RNA` and `nFeature_RNA` and a mixed expression profile (e.g. T-cell markers AND B-cell markers in the same cell). The simple `nFeature_RNA < 2500` filter from Step 7 catches some, but doublets that pair two *different* cell types of similar size will pass that filter — you need a model.

`scDblFinder` simulates artificial doublets, trains a classifier to distinguish them from singletons, and returns a per-cell score and class. Critically, **doublets form within a sample, not across samples**, so we run it per sample.

```{r}
#| label: dbl

# install if needed:
# BiocManager::install("scDblFinder")
library(scDblFinder)

set.seed(2026)

# Run scDblFinder per sample, then merge the results back into the Seurat object
ifnb_list <- SplitObject(ifnb, split.by = "stim")
ifnb_list <- lapply(ifnb_list, function(x) {
  sce <- as.SingleCellExperiment(x)
  sce <- scDblFinder(sce, returnType = "sce")
  x$scDblFinder.score <- colData(sce)$scDblFinder.score
  x$scDblFinder.class <- colData(sce)$scDblFinder.class    # "singlet" or "doublet"
  x
})
ifnb <- merge(ifnb_list[[1]], y = ifnb_list[-1])

# `merge()` in Seurat v5 leaves the RNA assay with per-sample split layers
# (counts.CTRL / counts.STIM, etc.). Downstream tools like FindMarkers expect
# a single joined layer, so rejoin them now.
ifnb <- JoinLayers(ifnb)

# How many doublets were called per sample?
table(sample = ifnb$stim, class = ifnb$scDblFinder.class)

# Inspect: doublets should sit at the high end of nCount/nFeature
VlnPlot(ifnb,
        features = c("nFeature_RNA", "nCount_RNA"),
        group.by = "scDblFinder.class", pt.size = 0)
```

::: {.callout-important title="Think about it"}
1.  Why run `scDblFinder` **per sample** instead of on the merged object?
2.  scDblFinder marks \~5–10% of cells as doublets in a typical 10x run. If your run shows 25%, what likely happened?
3.  Some pipelines run doublet detection *after* clustering and remove doublet-enriched clusters wholesale. Why might that be more conservative than removing per-cell doublet calls?

<details>

<summary>Click for answer</summary>

1.  Doublets are made when **two cells from the same droplet load** are co-encapsulated. Cells from different samples (CTRL and STIM here) were prepared in separate channels, so a STIM and a CTRL cell can never be a doublet. Running per-sample also lets each sample's doublet-rate be modeled correctly.
2.  Either you over-loaded the chip (the loading concentration was too high — typical for "nominal 10k cells, recovered 18k") or there's a quality problem (debris, large cell aggregates) making the simulated-doublet classifier confused. Inspect `nCount_RNA`/`nFeature_RNA` distributions and the ambient fraction.
3.  Per-cell doublet calls have false positives. If you trust them and remove individual cells, you may *also* be removing legitimate cycling cells, large activated cells, or megakaryocytes (which look doublet-like by sheer transcriptional output). Removing doublet-enriched **clusters** is more conservative — only act on a cluster where the doublet fraction is clearly elevated and the markers don't make biological sense.

</details>
:::

## Step 7 — Filter low-quality cells

```{r}
#| label: filter

ifnb <- subset(
  ifnb,
  subset = nFeature_RNA > 200 &
           nFeature_RNA < 2500 &
           percent.mt   < 5 &
           scDblFinder.class == "singlet"
)
ifnb
table(ifnb$stim)

dim_after_filter <- dim(ifnb)
```

::: {.callout-important title="Think about it"}
1.  Did the CTRL or STIM sample lose more cells in this filter? If they lost noticeably different fractions, what would that suggest?
2.  How would you set thresholds *empirically* rather than by default?

<details>

<summary>Show answers</summary>

1.  If one sample loses a much larger fraction at `percent.mt < 5` it usually means its cells were under more stress at dissociation or stimulation. For `ifnb`, IFN-β stimulation can mildly elevate `percent.mt` in some donors — flag the sample, but don't necessarily relax the threshold without thinking.
2.  Look at the per-sample violin / histogram shape; choose thresholds that trim the clear **tails** (not the mode). For an automated rule, use `scater::isOutlier()` or `median ± k·MAD` — and consider running it **per sample** so a globally lopsided distribution doesn't push the cutoff in the wrong direction for one sample.

</details>
:::

## Step 8 — Normalize

```{r}
#| label: normalize

ifnb <- NormalizeData(ifnb)
# Equivalent explicit form:
# NormalizeData(ifnb,
#               normalization.method = "LogNormalize",
#               scale.factor = 10000)

dim_after_normalize <- dim(ifnb)
```

::: {.callout-important title="Think about it"}
1.  What does `LogNormalize` do algebraically?
2.  When would you prefer `SCTransform()` instead?

<details>

<summary>Show answers</summary>

1.  For each cell: divide each gene's count by the cell's total count, multiply by `scale.factor` (default 10 000), then take `log1p`. In formula form: `log(1 + (count_gc / total_c) * 10000)`.
2.  `SCTransform` models the mean–variance relationship per gene with a regularized negative binomial, which is usually better for **low-count** datasets and more faithful downstream variance. It also replaces the NormalizeData → FindVariableFeatures → ScaleData chain in a single call. For `ifnb` either approach works fine.

</details>
:::

## Step 9 — Highly variable features

```{r}
#| label: hvgs

ifnb <- FindVariableFeatures(
  ifnb,
  selection.method = "vst",
  nfeatures = 2000
)

top10 <- head(VariableFeatures(ifnb), 10)
plot1 <- VariableFeaturePlot(ifnb)
LabelPoints(plot = plot1, points = top10, repel = TRUE)
```

::: {.callout-important title="Think about it"}
1.  Which genes are at the top of `top10`? Many will be **interferon-stimulated genes** (ISGs) like `ISG15`, `IFI6`, `IFIT1` — why?
2.  What happens if you set `nfeatures = 500`? Or `nfeatures = 10000`?

<details>

<summary>Show answers</summary>

1.  Because half the dataset (the STIM sample) has IFN-β–induced ISG expression while the other half (CTRL) does not, ISGs have **enormous variance across cells** and rise to the top of the HVG list. This is exactly the signal we'll see drive batch-style separation in Tutorial 02 — and it's exactly what integration in Tutorial 05 will need to handle without erasing.
2.  Too few (500) risks missing subtle cell-type signals and is brittle to noisy gene choices. Too many (10 000) pulls in low-information genes and dilutes the signal. 1 500–3 000 is a common sweet spot; 2 000 is a good default.

</details>
:::

## Step 10 — Scale

```{r}
#| label: scale

ifnb <- ScaleData(ifnb)
```

By default `ScaleData` only scales the variable features identified in Step 9 (~2,000 HVGs) — that's all PCA needs, and it keeps the `scale.data` slot (and the saved `.rds`) small. If you later need scaled values for a specific non-HVG gene (e.g. to draw it on a heatmap), re-run `ScaleData(ifnb, features = <those genes>)` at that point.

::: {.callout-important title="Think about it"}
1.  What does `ScaleData` do to each gene numerically?
2.  Why is scaling necessary *before* PCA?
3.  Could you `vars.to.regress = "stim"` here and skip integration in Tutorial 05?

<details>

<summary>Show answers</summary>

1.  Centers each gene's expression to mean 0 and scales to unit variance across cells (z-scoring).
2.  PCA finds directions of maximum variance. Without scaling, highly-expressed genes (which inherently have large raw variance) would dominate PCs, regardless of whether they are biologically informative.
3.  You *could*, but regressing on a binary covariate is a **brittle** form of batch correction — it removes a linear effect uniformly across all cells, but real batch effects are usually non-linear and cell-type-specific. Modern integration methods (Harmony, anchors, scVI) are much better tools. We use `ScaleData` here as the standard pre-PCA step and reserve the integration question for Tutorial 05.

</details>
:::

## Counts matrix progression — summary

```{r}
#| label: dim-summary

dim_summary <- tibble::tribble(
  ~step,                                                    ~features_after,        ~barcodes_after,
  "Step 1 — Kang18_8vs8 (singlets, raw)",                    dim_raw[1],             dim_raw[2],
  "Step 7 — subset() QC filter (nFeature_RNA, percent.mt)",  dim_after_filter[1],    dim_after_filter[2],
  "Step 8 — NormalizeData (no dimension change)",            dim_after_normalize[1], dim_after_normalize[2]
)
dim_summary
```

You should see roughly:

| Step | Features after | Barcodes after |
|---|---:|---:|
| Step 1 — `Kang18_8vs8` (singlets, annotated)                    | ~35,635 | ~24,500 |
| Step 7 — `subset()` QC filter (`nFeature_RNA`, `percent.mt`)    | ~35,635 | ~22,000 |
| Step 8 — `NormalizeData` (no dimension change)                  | ~35,635 | ~22,000 |

(Numbers will vary by a few hundred depending on package versions.)

## Save and continue to Tutorial 02

```{r}
#| label: save

# Make sure ../data/ exists
dir.create("../data", showWarnings = FALSE)

saveRDS(ifnb, file = "../data/ifnb_preprocessed.rds")
```

::: callout-tip
The saved `.rds` is the starting point for **Tutorial 02 — Dimensionality reduction & clustering**. We deliberately keep both `CTRL` and `STIM` cells in the same object — Tutorial 02 will run PCA + clustering on the unintegrated joint data so you can **see the batch effect** before fixing it in Tutorial 05.
:::

## Credits

-   Dataset: Kang *et al.* (2017) *Nat Biotechnol* 36: 89–94, [doi:10.1038/nbt.4042](https://doi.org/10.1038/nbt.4042) — fetched via Bioconductor's [`muscData::Kang18_8vs8()`](https://bioconductor.org/packages/muscData/) (same data the legacy `SeuratData::LoadData("ifnb")` route returned)
-   Reference vignettes: [Seurat integration tutorial](https://satijalab.org/seurat/articles/integration_introduction.html) · [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)
-   Inspiration for the multi-step QC pacing: [scNotebooks](https://integrativebioinformatics.github.io/scNotebooks/) (Sengupta lab)
