---
title: "Tutorial 05 — Reference-based Annotation (SingleR + Azimuth)"
subtitle: "Map cells against a labeled reference for fast, reproducible cell-type calls"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

A short hands-on companion to **[Lecture 04 --- Annotation](../Lecture_Folder/Lecture_04_Annotation.html)**. We start from the **clustered object saved in Tutorial 02** and:

-   Run `SingleR` against a built-in `celldex` reference
-   Run `Azimuth` against the PBMC reference (workshop default; substitute `pbmcref` with another for your own data)
-   Compare the two automatic calls against your manual labels from Tutorial 03
-   Reconcile disagreements and save a final, reference-validated annotation

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it. To run it:

1.  [Download the `.qmd` source](Tutorial_05_Reference_Annotation.qmd) and open it in RStudio.
2.  Make sure you've completed **[Tutorial 02](Tutorial_02_DimReduction_Clustering.html)** and **[Tutorial 03](Tutorial_03_Markers_Annotation.html)** first.
3.  Work through the chunks, flipping `eval: true` as you go.
:::

## Setup

```{r}
#| label: setup
library(Seurat)
library(SingleR)
library(celldex)
library(SingleCellExperiment)
library(tidyverse)
library(patchwork)

set.seed(2026)

seu <- readRDS("../data/nsclc_annotated.rds")
DimPlot(seu, group.by = "celltype", label = TRUE) + NoLegend()
```

## Path 1 --- `SingleR` against a `celldex` reference

`SingleR` correlates each query cell's expression against per-label reference profiles and returns the best-matching label.

### Pick a reference

```{r}
# Bulk reference covering 37 main cell types from purified human samples
ref <- celldex::HumanPrimaryCellAtlasData()

# Run SingleR on the (log-)normalized counts
sce  <- as.SingleCellExperiment(seu)
pred <- SingleR(test = sce, ref = ref,
                labels = ref$label.main,
                assay.type.test = "logcounts",
                BPPARAM = BiocParallel::SerialParam())
table(pred$labels)
```

Attach the calls back to the Seurat object:

```{r}
seu$singler_label <- pred$labels
seu$singler_score <- apply(pred$scores, 1, max)

DimPlot(seu, group.by = "singler_label", label = TRUE,
        repel = TRUE) + NoLegend()
```

### Score-distribution plot --- a built-in confidence check

```{r}
plotScoreHeatmap(pred)
plotDeltaDistribution(pred)
```

::: {.callout-important title="Think about it"}
Some cells will get a label like `"Macrophage"` with `delta` close to zero. What does that mean and what should you do about it?

<details>

<summary>Click for answer</summary>

`delta` is the gap between the cell's top-label score and the median score across all labels. A small `delta` means the reference can't decide --- the cell either looks ambiguous (multiple types fit equally well) or is a type the reference doesn't contain.

In practice: prune low-confidence calls (`pruneScores()` from SingleR), and trust your manual marker-based label more for those cells.

</details>
:::

```{r}
seu$singler_label_pruned <- ifelse(
  is.na(SingleR::pruneScores(pred)), NA, pred$labels
)
```

## Path 2 --- Azimuth (Seurat reference mapping)

Azimuth wraps Seurat's anchor-based reference mapping with curated atlases. The PBMC reference is the most widely tested.

```{r}
# install if needed
# remotes::install_github("satijalab/azimuth")
library(Azimuth)

# Will download the PBMC reference on first use (~1 GB)
seu <- RunAzimuth(seu, reference = "pbmcref")

# New columns: predicted.celltype.l1 / l2 / l3 + scores
head(seu@meta.data[, grep("predicted", colnames(seu@meta.data))])

DimPlot(seu, group.by = "predicted.celltype.l2",
        label = TRUE, repel = TRUE) + NoLegend()
```

::: {.callout-important title="Think about it"}
The PBMC reference is for **peripheral blood**. The workshop dataset is **NSCLC tumor cells**. What does Azimuth do with cells (e.g. epithelial / tumor) that aren't in the reference? How would you spot that?

<details>

<summary>Click for answer</summary>

Azimuth still assigns the best-matching label, but the **mapping score** (`mapping.score`) and per-label **prediction score** drop sharply for cells outside the reference's support. Plot `FeaturePlot(seu, "mapping.score")` --- low-score regions (often corresponding to tumor / non-immune cells) flag clusters where you should *not* trust Azimuth's calls.

For a tumor dataset like this, the right move is to use Azimuth on the immune compartment and rely on manual markers (or a tumor-specific reference) for the rest.

</details>
:::

## Compare the three annotations

```{r}
ann <- seu@meta.data |>
  select(celltype,                         # manual (from Tutorial 03)
         singler_label_pruned,             # SingleR (pruned)
         predicted.celltype.l2)            # Azimuth

table(ann$celltype, ann$singler_label_pruned, useNA = "ifany")
table(ann$celltype, ann$predicted.celltype.l2, useNA = "ifany")
```

A cleaner cross-tabulation as a heatmap:

```{r}
ann |>
  count(celltype, predicted.celltype.l2) |>
  ggplot(aes(predicted.celltype.l2, celltype, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Manual (rows) vs. Azimuth (cols)",
       x = NULL, y = NULL)
```

## Reconcile disagreements

A simple "agree-or-flag" rule: keep the manual label if at least one automatic method agrees; otherwise mark for review.

```{r}
seu$celltype_final <- with(seu@meta.data, dplyr::case_when(
  celltype == singler_label_pruned     ~ celltype,
  celltype == predicted.celltype.l2    ~ celltype,
  TRUE                                 ~ paste0("REVIEW:", celltype)
))

table(seu$celltype_final)
```

::: {.callout-important title="Think about it"}
A reviewer asks you to "use Azimuth labels instead of your manual ones because they're more reproducible". How do you respond?

<details>

<summary>Click for answer</summary>

Reproducibility is necessary but not sufficient. Azimuth is reproducible *and* biased by its reference: any cell type absent from the reference will be silently re-labeled. The right answer is to use both, as the tutorial does, and document where they disagree --- not to pick one because it's mechanical.

</details>
:::

## Save the final annotated object

```{r}
saveRDS(seu, "../data/nsclc_annotated_final.rds")
```

## Wrap-up

You now have:

-   A `SingleR` label per cell (with confidence pruning)
-   An Azimuth label per cell (with a mapping score)
-   A reconciled `celltype_final` you can carry into Tutorial 04 (pseudobulk DE) or downstream analyses

::: {.callout-tip title="Going further"}
-   Try a tissue-appropriate reference: `celldex::BlueprintEncodeData()`, the [Tabula Sapiens](https://tabula-sapiens-portal.ds.czbiohub.org/) atlas, or the [HCA Lung Cell Atlas](https://data.humancellatlas.org/).
-   Repeat with **CellTypist** (Python) for a third opinion.
-   For tumour data, look at **TumorDecon** or **ProjecTILs** to handle the tumour/immune mix more carefully.
:::

## See also

-   [Lecture 04 --- Annotation](../Lecture_Folder/Lecture_04_Annotation.html)
-   [Glossary --- Cell type annotation](../Resources_Folder/Glossary.html#c)
-   `SingleR` book: [bioconductor.org/books/release/SingleRBook](https://bioconductor.org/books/release/SingleRBook/)
-   Azimuth web tool: [azimuth.hubmapconsortium.org](https://azimuth.hubmapconsortium.org/)
