---
title: "Tutorial 16 — Spatial Transcriptomics with Visium"
subtitle: "Load a 10x Visium tissue, find spatially variable genes, integrate two sections, and sketch cell-type deconvolution"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

::: {.callout-note title="Running the code in this tutorial"}
Every code chunk here is tagged with `#| eval: false`, so the published page shows the code **without running it**. To run it yourself, open the downloaded `.qmd` in **RStudio**:

- **Run one chunk:** click the green ▶ (*Run Current Chunk*) at the chunk's top-right, or press *Ctrl/Cmd + Enter*. This runs the chunk **regardless of its `eval` setting** — the easiest way to work through the tutorial interactively.
- **Run a chunk on render:** change that chunk's `#| eval: false` to `#| eval: true` (or delete the line) so it executes when the document is rendered.
- **Render the whole tutorial:** click the **Render** button in RStudio (or run `quarto render` in a terminal) to execute the chunks top-to-bottom and knit a finished HTML. To run everything on render, set `eval: true` once in the YAML header at the top.
:::


## About this tutorial

A laptop-friendly **introduction to spatial transcriptomics** using 10x **Visium** — the most accessible spatial platform today. We use the `stxBrain` dataset from `SeuratData`: matched **anterior** and **posterior** sagittal sections of an adult mouse brain, \~50 MB total, installed with one R command (no manual download, no 10x click-through).


::: callout-note
**Companion lecture:** [Lecture 16 — Spatial Transcriptomics: Gene Expression in Tissue Context](../Lecture_Folder/Lecture_16_Spatial.html) · **Companion book chapter:** [Chapter 16 — Spatial Transcriptomics](../Resources_Folder/Chapter_16_Spatial.html) — the long-form prose treatment of this tutorial's material, with cross-references to the prerequisite appendices.
:::


You'll learn to:

-   Load Visium data into Seurat (counts + tissue image + spot coordinates)
-   Run the spatial-aware QC, normalization (`SCTransform`), and clustering workflow
-   Plot expression on top of the histology image
-   Find **spatially variable features** — genes whose expression has spatial structure beyond cluster membership
-   Integrate two tissue sections (the same anterior + posterior idea as `ifnb`'s CTRL + STIM)

::: callout-note
**Why mouse brain for a primer?** It has dramatic, well-known anatomy (cortex, hippocampus, thalamus, hypothalamus, cerebellum) — easy to recognize on the histology image, easy to validate with marker genes (`Hpca` in hippocampus, `Plp1` in white matter, `Pcp4` in cerebellum). A clean teaching dataset; no laptop will struggle with it.
:::

::: {.callout-tip title="How to use this page"}
1.  Download the `.qmd` source: <a href="Tutorial_16_Spatial_Transcriptomics.qmd.txt" download="Tutorial_16_Spatial_Transcriptomics.qmd">Tutorial_16_Spatial_Transcriptomics.qmd</a>. If your browser saves the file as `Tutorial_16_Spatial_Transcriptomics.qmd.txt`, **drop the trailing `.txt`** so the filename ends in `.qmd`, then open it in RStudio.
2.  Install the `stxBrain` dataset once (one R command — see [Setup](#setup)).
3.  Work through the chunks, flipping `eval: true` as you go.
:::

## Dataset — `stxBrain` (10x Visium adult mouse brain)

**What it is.** Two 10x Visium sections from the same adult mouse brain — one **anterior** (\~2,700 spots), one **posterior** (\~3,300 spots). Each spot is \~55 µm in diameter and captures the transcripts from \~1–10 cells underneath. The "spots" are the unit of analysis here, not single cells.

-   Distribution: `SeuratData::InstallData("stxBrain")` (no manual download)
-   Source: 10x Genomics public Visium dataset
-   Used in: this tutorial; the [Seurat spatial vignette](https://satijalab.org/seurat/articles/spatial_vignette.html)

::: callout-warning
A Visium "spot" is **not a single cell.** Most analyses borrow the language of single-cell (clusters, marker genes, integration) but interpret each unit as a small *neighbourhood* of cells. For true single-cell resolution use higher-resolution spatial platforms — Xenium, MERFISH, or Visium HD.
:::

## Learning goals

-   Load a Visium dataset and understand the four objects in a `SpatialAssay` (counts, image, coordinates, scale factors)
-   Run `SCTransform` on spatial data and explain why it's preferred over `LogNormalize` here
-   Use `SpatialFeaturePlot` and `SpatialDimPlot` to overlay expression / cluster on tissue
-   Find spatially variable features with Moran's I
-   Merge two tissue sections and run a joint clustering


::: {.callout-warning title="Common errors / things that bite"}
**`SeuratData::InstallData("stxBrain")` is slow** — \~50 MB download. First run takes a few minutes; subsequent runs are instant.

**`SCTransform` warns "iteration limit reached"** — usually safe to ignore. Visium spots have a wider count distribution than single cells, so SCT's NB-regression sometimes hits its iteration limit on outlier spots. The result is still usable.

**`SpatialFeaturePlot` shows nothing on the tissue** — the `images` slot wasn't loaded. After `SaveH5Seurat` / `Convert`, the histology image can be lost; `Read10X_Image(dirname(seu@tools$Staffli@imgs[1]))` re-attaches it.

**Anterior + posterior integration is slow / OOMs** — the SCT-anchor path is RAM-hungry on Visium. If you have <16 GB RAM, downsample spots first (`subset(brain, downsample = 1500)`) or use Harmony on the SCT residuals instead of `IntegrateData`.
:::


::: {.callout-tip title="Solutions / instructor copy"}
The *Think about it* prompts on this page have inline answers (click "Show answer" / "Click for answer" to expand). For the full **runnable instructor copy** with every chunk pre-evaluated and outputs shown:

- **For students:** the same `.qmd` you downloaded already contains every solution inline — just expand the collapsed answers
- **For instructors:** render with `quarto render --profile solutions` from the `Exercise_Folder/` directory. The fully-evaluated HTML is written to `docs/Exercise_Folder/_solutions/Tutorial_16_Spatial_Transcriptomics.html`. See [Exercise_Folder/_quarto-solutions.yml](https://github.com/wcresko/scRNAseq_tutorial/blob/main/Exercise_Folder/_quarto-solutions.yml) for the build profile

If you're running a graded course, link the rendered solutions page from your LMS (Canvas / Moodle / etc.) — it's behind the same `_solutions/` path on the workshop's deployed site.
:::

## Setup {#setup}

```{r}
#| label: setup

# Packages are installed in Tutorial 00 (Setup → bonus modules); the stxBrain
# dataset install is on the Datasets page. Here we just load the libraries.
library(Seurat)
library(SeuratData)
library(tidyverse)
library(patchwork)

# ---------------------------------------------------------------------------
# Output directory for this module's figures and tables.
# Every figure/table chunk below writes a file named  Mod16_C<chunk>_<name>
# into ../output/Mod16/ so it can be cross-referenced from the rest of the site.
#   Mod16 = Module 16 (this tutorial);  C<n> = the nth code chunk.
# ---------------------------------------------------------------------------
out_dir <- "../output/Mod16"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

set.seed(2026)
```

## Step 1 — Load both sections

```{r}
#| label: load

ant  <- LoadData("stxBrain", type = "anterior1")
post <- LoadData("stxBrain", type = "posterior1")

ant
post

# Check the assays — Visium objects have a "Spatial" assay (not "RNA")
DefaultAssay(ant)

# --- Table out: per-section dimensions (features x spots) -------------------
tibble::tibble(
  section  = c("anterior1", "posterior1"),
  features = c(nrow(ant), nrow(post)),
  spots    = c(ncol(ant), ncol(post))
) |>
  readr::write_csv(file.path(out_dir, "Mod16_C2_section_dimensions.csv"))
```

::: {.callout-important title="Think about it"}
1.  How many features (genes) are in each section? How many spots?
2.  What's stored in `ant@images`?

<details>

<summary>Show answers</summary>

1.  About 32,000 genes × \~2,700 spots (anterior) and \~3,300 spots (posterior). Each spot is bigger than a single cell, so the absolute spot count is much smaller than a typical scRNA-seq experiment.
2.  An object of class `VisiumV1` containing the **histology image** (low-res JPEG), the **per-spot tissue-coordinate table** (which pixel each spot maps to), and the **scale factors** that translate between Visium's three coordinate systems (full-res, low-res, hi-res). `SpatialFeaturePlot` uses all of these to draw expression on top of the H&E.

</details>
:::

## Step 2 — Spatial QC

The same per-spot QC metrics as scRNA-seq apply: `nCount_Spatial`, `nFeature_Spatial`, percent mitochondrial. **But also visualize them spatially** — a localized QC outlier (e.g. one corner of the tissue with low UMI counts) is a tissue artefact (bubble, fold, edge cells), not random noise.

```{r}
#| label: qc
#| fig-cap: "Per-spot QC for the anterior section: violin distributions and the same metrics overlaid on the tissue."

ant[["percent.mt"]] <- PercentageFeatureSet(ant, pattern = "^mt-")
post[["percent.mt"]] <- PercentageFeatureSet(post, pattern = "^mt-")

# Standard violin
p1 <- VlnPlot(ant,
              features = c("nCount_Spatial", "nFeature_Spatial", "percent.mt"),
              pt.size = 0, ncol = 3) &
  xlab("Anterior section") &
  theme(plot.title = element_text(size = 11))
p1 <- p1 +
  patchwork::plot_annotation(
    title    = "Per-spot QC metrics — stxBrain anterior section",
    subtitle = "UMIs/spot, genes/spot and mitochondrial % distributions",
    caption  = "Module 16 · Spatial Transcriptomics"
  )

# Spatial overlay — same metrics, on the tissue
p2 <- SpatialFeaturePlot(ant,
                         features = c("nCount_Spatial", "percent.mt"),
                         pt.size.factor = 1.6) +
        plot_layout(ncol = 2)
p2 <- p2 +
  patchwork::plot_annotation(
    title    = "Spatial QC overlay — stxBrain anterior section",
    subtitle = "UMIs/spot and mitochondrial % on the histology image",
    caption  = "Module 16 · Spatial Transcriptomics"
  )

p1
p2

# --- Figures out -----------------------------------------------------------
ggsave(file.path(out_dir, "Mod16_C3_qc_violins.png"), p1,
       width = 10, height = 4, dpi = 300)
ggsave(file.path(out_dir, "Mod16_C3_qc_spatial.png"), p2,
       width = 10, height = 5, dpi = 300)
```

::: callout-tip
`pattern = "^mt-"` (lowercase) for mouse mitochondrial genes. For human use `"^MT-"`.
:::

::: {.callout-important title="Think about it"}
On the spatial `nCount_Spatial` overlay, regions of consistently low UMI counts often correspond to specific tissue features. What might they be?

<details>

<summary>Show answers</summary>

White matter is famously low-yield on Visium because the tissue is mostly axons (myelin), with relatively few neuronal cell bodies — fewer cells per spot, lower transcript counts per spot. Tissue **edges** also drop off because spots there are only partially covered. Don't filter these spots reflexively — for the brain dataset they are real biology.

</details>
:::

## Step 3 — Normalize with SCTransform

`SCTransform` is the recommended normalization for Visium because spot-level counts have a **wider dynamic range** than single-cell counts (each spot pools 1–10 cells), and `SCTransform`'s regularized negative binomial model handles that better than `LogNormalize`.

```{r}
#| label: norm

ant  <- SCTransform(ant,  assay = "Spatial", verbose = FALSE)
post <- SCTransform(post, assay = "Spatial", verbose = FALSE)

DefaultAssay(ant)    # Should now be "SCT"
```

## Step 4 — Visualize known marker genes on the tissue

Sanity-check that the data captures known anatomy. For the mouse brain:

-   `Hpca` — hippocampus
-   `Pcp4` — cerebellum, parts of cortex
-   `Plp1` — myelin / white matter
-   `Sst` — somatostatin interneurons
-   `Th`  — dopaminergic / noradrenergic neurons (substantia nigra)
-   `Cck` — broadly cortical inhibitory neurons

```{r}
#| label: markers
#| fig-cap: "Known anatomical marker genes overlaid on the anterior and posterior tissue sections."

p_markers_ant <- SpatialFeaturePlot(ant, features = c("Hpca", "Plp1", "Sst"),
                   ncol = 3, alpha = c(0.1, 1)) +
  patchwork::plot_annotation(
    title    = "Anatomical markers on tissue — anterior section",
    subtitle = "Hpca (hippocampus), Plp1 (white matter), Sst (interneurons)",
    caption  = "Module 16 · Spatial Transcriptomics"
  )

p_markers_post <- SpatialFeaturePlot(post, features = c("Hpca", "Plp1", "Pcp4"),
                   ncol = 3, alpha = c(0.1, 1)) +
  patchwork::plot_annotation(
    title    = "Anatomical markers on tissue — posterior section",
    subtitle = "Hpca (hippocampus), Plp1 (white matter), Pcp4 (cerebellum)",
    caption  = "Module 16 · Spatial Transcriptomics"
  )

p_markers_ant
p_markers_post

# --- Figures out -----------------------------------------------------------
ggsave(file.path(out_dir, "Mod16_C5_markers_anterior.png"),  p_markers_ant,
       width = 12, height = 4, dpi = 300)
ggsave(file.path(out_dir, "Mod16_C5_markers_posterior.png"), p_markers_post,
       width = 12, height = 4, dpi = 300)
```

::: {.callout-important title="Think about it"}
`Hpca` should be highest in the hippocampus (a banana-shaped structure in the medial part of each section). Can you see the hippocampus light up? If not, you may have the section orientation flipped — SeuratData's `stxBrain` is stored with the cortical surface roughly **at the top** of the image.

<details>

<summary>Show answers</summary>

If you don't recognize the hippocampus, open the [Allen Mouse Brain Atlas](https://mouse.brain-map.org/static/atlas) for a side-by-side reference. The anterior section catches the *front* of the hippocampus and not much else of it; the posterior section gets the bulk. `Hpca` will appear stronger on `posterior1` than on `anterior1`.

</details>
:::

## Step 5 — Cluster spots

Same workflow as scRNA-seq, but on the SCT assay:

```{r}
#| label: cluster
#| fig-cap: "Anterior-section spot clusters shown as a UMAP and overlaid on the tissue."

ant <- ant |>
  RunPCA(assay = "SCT", verbose = FALSE) |>
  FindNeighbors(reduction = "pca", dims = 1:30) |>
  FindClusters(resolution = 0.5, verbose = FALSE) |>
  RunUMAP(reduction = "pca", dims = 1:30)

# UMAP and tissue side-by-side
p_umap <- DimPlot(ant, reduction = "umap", label = TRUE) +
            labs(title = "Anterior — UMAP", x = "UMAP 1", y = "UMAP 2",
                 colour = "Cluster")
p_tis  <- SpatialDimPlot(ant, label = TRUE, label.size = 3) +
            labs(title = "Anterior — clusters on tissue", fill = "Cluster")

p_cluster <- (p_umap + p_tis) +
  patchwork::plot_annotation(
    title    = "Spot clustering — stxBrain anterior section",
    subtitle = "SNN clusters on the SCT assay, in UMAP space and on the histology image",
    caption  = "Module 16 · Spatial Transcriptomics"
  )
p_cluster

# --- Figure out ------------------------------------------------------------
ggsave(file.path(out_dir, "Mod16_C6_anterior_clusters.png"), p_cluster,
       width = 12, height = 5, dpi = 300)
```

::: {.callout-important title="Think about it"}
A spatial cluster that occupies a single contiguous anatomical region (e.g. one cluster covers the entire hippocampus) is biologically meaningful. A cluster that's scattered randomly across the section in single spots is more likely a technical / mixed-cell artefact. Why?

<details>

<summary>Show answers</summary>

Real cell types are spatially organized in tissue — neurons in cortex, oligodendrocytes in white matter, etc. A cluster whose spots all share *spatial* proximity is recovering an anatomical structure. A cluster whose spots are scattered is recovering either (a) very rare cells distributed throughout (immune cells, vasculature — sometimes legitimate) or (b) a technical signature (depth, ambient RNA, batch). The spatial layout is itself a quality check.

</details>
:::

## Step 6 — Spatially variable features (Moran's I)

Some genes are variable *and* their variation has spatial structure (a clear pattern on the tissue), some are variable but spatially random. The first kind is what you usually want to study in a spatial dataset.

```{r}
#| label: svf
#| fig-cap: "Top spatially variable genes (Moran's I) in the anterior section, overlaid on the tissue."

ant <- FindSpatiallyVariableFeatures(
  ant,
  assay         = "SCT",
  features      = VariableFeatures(ant)[1:1000],   # restrict to top-1000 HVGs for speed
  selection.method = "moransi"
)

# Top spatially variable features
top_sv <- SpatiallyVariableFeatures(ant, selection.method = "moransi")[1:6]
top_sv

p_svf <- SpatialFeaturePlot(ant, features = top_sv, ncol = 3, alpha = c(0.1, 1)) +
  patchwork::plot_annotation(
    title    = "Top spatially variable features (Moran's I) — anterior section",
    subtitle = "Genes whose expression has the strongest spatial structure",
    caption  = "Module 16 · Spatial Transcriptomics"
  )
p_svf

# --- Outputs: top spatially variable genes (table) + tissue overlay --------
tibble::tibble(rank = seq_along(top_sv), gene = top_sv) |>
  readr::write_csv(file.path(out_dir, "Mod16_C7_spatially_variable_features.csv"))

ggsave(file.path(out_dir, "Mod16_C7_spatially_variable_features.png"), p_svf,
       width = 12, height = 7, dpi = 300)
```

::: {.callout-important title="Think about it"}
A gene with high variance but **low** Moran's I is variable spot-to-spot but with no spatial pattern — what does that suggest?

<details>

<summary>Show answers</summary>

Either (a) genuine cell-cell heterogeneity within an anatomical region (rare cell types scattered across a region), or more often (b) a technical signature like ambient-RNA bleed, mitochondrial percent, or sequencing depth. Spatial variability is a stronger biological signal than overall variability.

</details>
:::

## Step 7 — Merge anterior + posterior, joint cluster

The same two-sample integration logic as `ifnb`, but on Visium. We use `SCTransform` integration:

```{r}
#| label: merge-integrate
#| fig-cap: "Joint embedding of both sections after SCT integration: UMAP coloured by section and cluster, plus per-section cluster maps on tissue."

# Each section is its own "batch"
ant$slice  <- "anterior"
post$slice <- "posterior"

brain.list <- list(ant = ant, post = post)
features   <- SelectIntegrationFeatures(object.list = brain.list,
                                        nfeatures   = 3000,
                                        verbose     = FALSE)
brain.list <- PrepSCTIntegration(object.list = brain.list,
                                 anchor.features = features,
                                 verbose = FALSE)
anchors    <- FindIntegrationAnchors(object.list = brain.list,
                                     normalization.method = "SCT",
                                     anchor.features = features,
                                     verbose = FALSE)
brain      <- IntegrateData(anchorset = anchors,
                            normalization.method = "SCT",
                            verbose = FALSE)

DefaultAssay(brain) <- "integrated"

brain <- brain |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:30) |>
  FindClusters(resolution = 0.5, verbose = FALSE) |>
  RunUMAP(dims = 1:30)

# Joint UMAP
p_int_umap <- DimPlot(brain, group.by = c("slice", "seurat_clusters")) +
  patchwork::plot_annotation(
    title    = "Integrated UMAP — anterior + posterior sections",
    subtitle = "Coloured by section of origin and by joint cluster",
    caption  = "Module 16 · Spatial Transcriptomics"
  )
p_int_umap

# And per-slice spatial cluster maps
p_int_spatial <- SpatialDimPlot(brain, label = TRUE) +
  patchwork::plot_annotation(
    title    = "Integrated clusters on tissue — both sections",
    subtitle = "Joint cluster assignments mapped back onto each section",
    caption  = "Module 16 · Spatial Transcriptomics"
  )
p_int_spatial

# --- Figures out -----------------------------------------------------------
ggsave(file.path(out_dir, "Mod16_C8_integrated_umap.png"), p_int_umap,
       width = 12, height = 5, dpi = 300)
ggsave(file.path(out_dir, "Mod16_C8_integrated_spatial_clusters.png"), p_int_spatial,
       width = 10, height = 5, dpi = 300)
```

::: {.callout-important title="Think about it"}
Why integrate sections from the same brain at all? The animal is one animal — what could differ?

<details>

<summary>Show answers</summary>

Even matched sections can drift: minor differences in tissue handling (RNA quality, embedding, cryostat thickness), Visium chip-to-chip variation, different lots of capture probes, different sequencing depth. Integration aligns the technical axes so the *anatomical* differences (anterior catches striatum, posterior catches cerebellum) are what dominates the joint embedding rather than chip-to-chip noise.

</details>
:::

## Step 8 — (Optional) cell-type deconvolution from a single-cell reference

A Visium spot covers \~1–10 cells, so each spot's transcriptome is a *mixture*. Tools like **`spacexr`** (the RCTD method), **`CARD`**, **`cell2location`** (Python), and **`SPOTlight`** decompose each spot into estimated cell-type proportions, using a single-cell reference of the same tissue.

This is too heavy to install and run end-to-end here, but the recipe is:

*Optional reference recipe — left `#| eval: false`; it's heavy to run and is here as a pattern for your own data, so you don't need to change anything.*

```{r}
#| label: deconv-recipe
#| eval: false

# Reference recipe for an actual deconvolution with spacexr (RCTD).
library(spacexr)

# 1. Build a reference from a single-cell brain dataset (e.g. Allen Mouse Brain
#    Atlas or a Tabula Muris brain subset) — a Reference() object with cell-type labels.
# 2. Build a SpatialRNA() object from your Visium counts + spot coordinates.

# 3. Run RCTD:
rctd <- create.RCTD(spatial_rna, reference, max_cores = 4)
rctd <- run.RCTD(rctd, doublet_mode = "doublet")

# 4. Each spot now has weights for the contributing cell types; visualize as a
#    per-cell-type SpatialFeaturePlot.
```

For the laptop workflow here, **a fast proxy** is to compute a **module score** for each cell type using its top markers from a single-cell reference. We borrow markers from the `ifnb` PBMC reference for one quick example (mouse brain doesn't have PBMCs, so this is purely illustrative — for a real run, use a brain-tissue scRNA-seq reference):

*Optional illustrative sketch — left `#| eval: false`; you don't need to change it.*

```{r}
#| label: module-score
#| eval: false

# Illustrative sketch only (cross-species gene-name conversion is non-trivial):
brain <- AddModuleScore(brain,
                        features = list(Microglia = c("Cx3cr1", "P2ry12", "Tmem119")),
                        name     = "Microglia")
SpatialFeaturePlot(brain, features = "Microglia1")
```

## Step 9 — Save

```{r}
#| label: save

dir.create("../data", showWarnings = FALSE)
saveRDS(brain, "../data/stxBrain_integrated.rds")
```

## Wrap-up

You've done the canonical Visium workflow on a real two-section dataset:

1.  Loaded Visium data including the histology image
2.  Visualized QC and marker expression on the tissue
3.  Normalized with `SCTransform` and clustered the spots
4.  Found spatially variable features (Moran's I)
5.  Integrated two sections and re-clustered jointly
6.  Sketched the recipe for cell-type deconvolution

::: {.callout-tip title="Going further"}
-   **Higher-resolution platforms.** Visium HD (subcellular bins), 10x Xenium, MERFISH, NanoString CosMx — same conceptual workflow, much higher resolution. Seurat 5 supports Visium HD natively; Xenium requires Vitessce / Seurat 5 / `xeniumtools`.
-   **Cell–cell communication in space.** Tools like `LIANA+` (Python) extend the LIANA framework to spatial data — they ask which ligand-receptor pairs are *spatially co-located*, not just co-expressed.
-   **Spatial domains via graph methods.** `BANKSY` and `STAGATE` produce spatially-aware clusters that explicitly use coordinates as features, often cleaner than the SNN approach above.
-   **Image-aware normalization.** `STUtility` and `semla` add tissue-image-based normalization that handles section-edge effects better than off-the-shelf SCT.
:::

## See also

-   [Seurat spatial vignette](https://satijalab.org/seurat/articles/spatial_vignette.html) — the canonical reference workflow this tutorial follows
-   [scNotebooks Module 10](https://integrativebioinformatics.github.io/scNotebooks/modules/en/module10/module10.html) — covers Visium + 3D visualization + deconvolution on 10x breast cancer
-   `spacexr` (RCTD) — [github.com/dmcable/spacexr](https://github.com/dmcable/spacexr)
-   Moses & Pachter 2022 *Nat Methods* — landscape review of spatial methods, [doi:10.1038/s41592-022-01409-2](https://doi.org/10.1038/s41592-022-01409-2)
