---
title: "Tutorial 05 — DESeq2 Primer on `airway`"
subtitle: "A bulk-RNA-seq DE walkthrough before we tackle pseudobulk (Patel-style)"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

A primer that gets the DESeq2 mechanics out of the way **before** the pseudobulk DE tutorial (Tutorial 06). The dataset is `airway` (a Bioconductor-shipped 8-sample bulk RNA-seq experiment of dexamethasone-treated airway smooth muscle cells), so there is **no manual download** — installing the package gives you the data.

This tutorial follows [Patel's `runDESeq2.R`](https://github.com/kpatel427/YouTubeTutorials/blob/main/runDESeq2.R) and [`getData.R`](https://github.com/kpatel427/YouTubeTutorials/blob/main/getData.R), with workshop-specific call-out boxes.

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it. To run it:

1.  Download the `.qmd` source: <a href="Tutorial_05_Primer_DESeq2_Basics.qmd.txt" download="Tutorial_05_Primer_DESeq2_Basics.qmd">Tutorial_05_Primer_DESeq2_Basics.qmd</a>. If your browser saves the file as `Tutorial_05_Primer_DESeq2_Basics.qmd.txt`, **drop the trailing `.txt`** so the filename ends in `.qmd`, then open it in RStudio.
2.  Install the packages (block below) — no other download needed.
3.  Work through the chunks, flipping `eval: true` as you go.
:::

::: callout-note
**Companion lecture:** [Lecture 04 — Pseudobulk DE & DESeq2](../Lecture_Folder/Lecture_04_Pseudobulk_DE.html)
:::

## Dataset — `airway`

**What it is.** RNA-seq of four primary human airway smooth muscle cell lines, treated **untreated** vs. **dexamethasone**, two technical replicates each → **8 samples**. Standard reference dataset for teaching DESeq2.

-   Bioconductor landing page: <https://bioconductor.org/packages/airway/>
-   Original paper: Himes et al. (2014) *PLoS ONE* 9: e99625

There's no separate download — the counts and metadata ride along with the package.

```{r}
#| label: install
# One-time install (uncomment if needed):
# install.packages("BiocManager")
# BiocManager::install(c("DESeq2", "airway", "apeglm", "tidyverse"))
```

## Setup

```{r}
#| label: setup
library(DESeq2)
library(airway)
library(tidyverse)
set.seed(2026)
```

## Step 1 — Load the data and pull out counts + metadata

```{r}
#| label: load
data(airway)
airway

counts_data <- assay(airway)         # genes × samples integer matrix
colData     <- as.data.frame(colData(airway))

# Tidy the metadata to mimic Patel's getData.R: keep cell line + treatment,
# rename "trt"/"untrt" to something readable.
colData <- colData[, c("cell", "dex")]
colData$dex <- factor(gsub("trt",   "treated",   colData$dex))
colData$dex <- factor(gsub("untrt", "untreated", colData$dex))

head(counts_data[, 1:4])
colData
```

::: {.callout-important title="Think about it"}
1.  What's in the rows of `counts_data`? In the columns?
2.  Why do we want `dex` as a `factor` rather than a `character`?

<details>
<summary>Show answers</summary>

1.  Rows are **Ensembl gene IDs**; columns are **sample IDs** (`SRR…` accessions). Values are integer read counts.
2.  DESeq2 builds a model matrix from factors and uses **factor levels** to decide which group is the *reference* (denominator) in `log2FoldChange`. A character column would be coerced silently and might pick the wrong reference.
</details>
:::

## Step 2 — Sanity-check the metadata

DESeq2 fails fast and unhelpfully if `colData` rows don't line up with `countData` columns. Always check.

```{r}
#| label: sanity
all(colnames(counts_data) %in% rownames(colData))
all(colnames(counts_data) == rownames(colData))
```

::: callout-warning
Both should return `TRUE`. If the second is `FALSE`, **reorder `colData`** to match — don't reorder the counts (you'll forget which orientation you fixed). `colData <- colData[colnames(counts_data), ]`.
:::

## Step 3 — Build the `DESeqDataSet`

```{r}
#| label: build-dds
dds <- DESeqDataSetFromMatrix(
  countData = counts_data,
  colData   = colData,
  design    = ~ cell + dex
)
dds
```

::: {.callout-important title="Think about it"}
The design is `~ cell + dex`. Why include `cell` if we only care about `dex`?

<details>
<summary>Show answers</summary>

The eight samples come from **four different cell lines**. Each cell line is a paired (treated, untreated) block. Adding `cell` as a covariate **soaks up the cell-line effect** so the test for `dex` is within-line — the bulk equivalent of a paired analysis.
</details>
:::

## Step 4 — Pre-filter low-count genes

```{r}
#| label: prefilter
keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]
nrow(dds)
```

::: callout-tip
**Pre-filtering is for speed only**, not for FDR control. DESeq2 already does *independent filtering* at the `results()` step, which protects multiple testing. Dropping rows with `< 10` total counts just makes `DESeq()` faster.
:::

## Step 5 — Set the reference level

```{r}
#| label: relevel
dds$dex <- relevel(dds$dex, ref = "untreated")
```

::: callout-warning
**This step is easy to forget and easy to misread.** With `ref = "untreated"`, a positive `log2FoldChange` means the gene is **higher in treated**. Without `relevel()`, R's alphabetical default would flip the sign on you.
:::

## Step 6 — Run `DESeq()` and pull results

```{r}
#| label: run-deseq
dds <- DESeq(dds)
res <- results(dds, contrast = c("dex", "treated", "untreated"))
summary(res)
```

::: {.callout-important title="Think about it"}
`summary(res)` reports counts of "LFC > 0 (up)" and "outliers" and "low counts". What do "outliers" and "low counts" mean here?

<details>
<summary>Show answers</summary>

-   **Outliers** are genes flagged by Cook's distance — a single sample with a wildly different count from the rest. DESeq2 sets `padj = NA` for these so they don't drive results.
-   **Low counts** are genes filtered by *independent filtering*: their mean expression is so low they have no power to be detected at the chosen FDR. DESeq2 estimates the optimal cut adaptively.
</details>
:::

## Step 7 — Shrink the LFCs before ranking / plotting

```{r}
#| label: lfcshrink
res_shrunk <- lfcShrink(dds,
                        coef = "dex_treated_vs_untreated",
                        type = "apeglm")
plotMA(res_shrunk, ylim = c(-3, 3))
```

::: callout-tip
**Always rank by shrunk LFCs.** Genes with `baseMean = 2` can show wild raw LFCs purely by sampling noise; `apeglm` shrinks those toward zero and de-clutters your top-hits list.
:::

## Step 8 — Inspect a top hit

```{r}
#| label: top-hit
top_genes <- as.data.frame(res_shrunk) |>
  rownames_to_column("gene") |>
  arrange(padj) |>
  head(5)
top_genes

plotCounts(dds, gene = top_genes$gene[1], intgroup = "dex")
```

::: {.callout-important title="Think about it"}
Why is `plotCounts()` a more honest way to confirm a hit than just trusting the p-value?

<details>
<summary>Show answers</summary>

It shows the **per-sample** values. A single outlier sample can drag the model into "significance" that disappears under inspection. Eyeballing the per-sample dots beats a single p-value every time.
</details>
:::

## Step 9 — Sanity check by transformation + PCA

Quality-control your samples *post-hoc* with a variance-stabilizing transform (`vst`) and a PCA:

```{r}
#| label: vst-pca
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex", "cell"))
```

::: callout-tip
**What you want to see.** Treated and untreated separated on PC1 or PC2 *within* each cell line. If samples cluster by cell line only and not by `dex`, your treatment effect is small relative to between-line variation — interpret the DE results accordingly.
:::

## Stretch goals

-   **Interaction terms.** Try `design = ~ cell + dex + cell:dex` and compare results. (See [Patel's `interactionTerms.R`](https://github.com/kpatel427/YouTubeTutorials/blob/main/interactionTerms.R).)
-   **Volcano plot.** Plot `-log10(padj)` vs. `log2FoldChange` from `res_shrunk` and label your top 20 hits.
-   **Pathway enrichment.** Feed the significant genes into `clusterProfiler::enrichGO()` for a downstream view.

## What you've practiced

-   Building a `DESeqDataSet` from a counts matrix + sample table
-   Adding a covariate to a design
-   Reading DESeq2 output (LFC, lfcSE, padj, baseMean)
-   Shrinking LFCs for honest ranking
-   QCing samples with `vst` + PCA

You're now ready for [Tutorial 06 — Pseudobulk DE](Tutorial_06_Pseudobulk_DE.html), where the same DESeq2 mechanics apply to a per-cell-type aggregated count matrix.

## Credits

-   Patel: [`runDESeq2.R`](https://github.com/kpatel427/YouTubeTutorials/blob/main/runDESeq2.R) and [`getData.R`](https://github.com/kpatel427/YouTubeTutorials/blob/main/getData.R)
-   Data: [`airway`](https://bioconductor.org/packages/airway/) (Himes et al., 2014, *PLoS ONE*)
