---
title: "Tutorial 03 — scRNA-seq: Markers & Cell Type Annotation"
subtitle: "Find cluster markers, visualize, label by hand and with SingleR"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

Part three of the series. We start from the **clustered object saved in Tutorial 02** and walk through:

-   Finding marker genes per cluster
-   The canonical marker visualizations (Vln / Feature / Heatmap / Ridge / Dot)
-   Manual cell-type annotation with `RenameIdents`
-   Automatic annotation with `SingleR`
-   Saving the final, annotated object

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it. To run it:

1.  [Download the `.qmd` source](Tutorial_03_Markers_Annotation.qmd) and open it in RStudio.
2.  Make sure you have completed **[Tutorial 02](Tutorial_02_DimReduction_Clustering.html)** — it writes the `nsclc_clustered.rds` that this tutorial loads.
3.  Work through the chunks, flipping `eval: true` as you go.
:::

## Dataset — resuming from Tutorial 02

This tutorial loads the clustered `.rds` written at the end of Tutorial 02:

| File | Produced by | What it is |
|---|---|---|
| `../data/nsclc_clustered.rds` | Tutorial 02 | Seurat object with PCA, SNN graph, clusters at five resolutions (`RNA_snn_res.0.1/0.3/0.5/0.7/1.0`), and a UMAP embedding. |

### Data dictionary — the Seurat object at this stage

| Slot | Class | Contents |
|---|---|---|
| `seu@meta.data$RNA_snn_res.0.1 … 1.0` | factor | Cluster assignments at each resolution from `FindClusters()`. Active `Idents()` is set to `RNA_snn_res.0.5` here. |
| `seu@reductions$pca` | `DimReduc` | Cell embeddings in PC space (`seu@reductions$pca@cell.embeddings`, dims = 1:50). |
| `seu@reductions$umap` | `DimReduc` | 2-D UMAP embedding for visualization. |
| `seu@graphs$RNA_nn`, `RNA_snn` | `Graph` | k-NN and SNN graphs from `FindNeighbors()`. |

### Optional — external reference used by `SingleR`

If you run the automatic-annotation chunk, `celldex::HumanPrimaryCellAtlasData()` downloads a bulk-RNA-seq reference compiled from [Mabbott *et al.* 2013](https://doi.org/10.1186/1471-2164-14-632). It is fetched from the `ExperimentHub` cache, not from the `data/` folder.

| Label | Meaning |
|---|---|
| `$label.main` | 37 coarse cell types (e.g. `T_cells`, `B_cell`, `Monocyte`, `Epithelial_cells`). |
| `$label.fine` | 157 finer subtypes (e.g. `T_cells:CD4+_Th1`, `Monocyte:CD14+`). |

The upstream dataset is the **10x 20k NSCLC DTC** sample — see [Tutorial 01](Tutorial_01_QC_Preprocessing.html) for its description and data dictionary.

## Learning goals

-   Use `FindMarkers` / `FindAllMarkers` appropriately
-   Read each of the five standard marker-visualization plots and know when to pick which
-   Assign cell-type labels manually and with a reference-based classifier
-   Cross-check manual and automatic labels to catch ambiguity

## Setup

```{r}
#| label: setup

library(Seurat)
library(tidyverse)

# For automatic annotation
# BiocManager::install(c("SingleR", "celldex"))
# library(SingleR)
# library(celldex)

set.seed(2026)
```

```{r}
#| label: resume

nsclc.seurat.obj <- readRDS("../data/nsclc_clustered.rds")
Idents(nsclc.seurat.obj) <- "RNA_snn_res.0.5"
nsclc.seurat.obj
```

------------------------------------------------------------------------

## Step 1 — Find markers for one cluster

```{r}
#| label: markers-one

cluster1.markers <- FindMarkers(nsclc.seurat.obj, ident.1 = 1, min.pct = 0.25)
head(cluster1.markers, 10)
```

::: {.callout-important title="Think about it"}
1.  What does `min.pct = 0.25` do, and why is that a useful default?
2.  The output has `avg_log2FC`, `pct.1`, `pct.2`. Which gene would you pick as a *specific* marker for this cluster?

<details>

<summary>Show answers</summary>

1.  Only tests genes detected in ≥ 25% of cells in at least one of the two groups. Prunes low-expression noise and speeds the test up significantly.
2.  A specific marker has **high `pct.1` and low `pct.2`** (expressed in almost all cluster-1 cells, almost none elsewhere) *and* high positive `avg_log2FC`. Markers that are broadly expressed (high `pct.2`) may be real but aren't *specific*.

</details>
:::

## Step 2 — Markers distinguishing one cluster from a pair

```{r}
#| label: markers-vs

cluster5.markers <- FindMarkers(
  nsclc.seurat.obj,
  ident.1 = 5,
  ident.2 = c(0, 3),
  min.pct = 0.25
)
head(cluster5.markers, 10)
```

::: {.callout-important title="Think about it"}
1.  Why might you compare cluster 5 *only* to clusters 0 and 3, rather than to all other cells?
2.  What's a situation where you'd want `ident.1` vs `ident.2` to be two groups of clusters (e.g. `ident.1 = c(5, 7)`)?

<details>

<summary>Show answers</summary>

1.  If 0 and 3 look similar on UMAP or markers to cluster 5, you want genes that **specifically** separate 5 from *those neighbors* — genome-wide vs-all tests can be dominated by genes separating cluster 5 from totally unrelated populations.
2.  When you have **biological groups** of clusters, e.g. merging two T-cell subtypes (5 + 7) to compare against NK cells (3 + 8). Grouping lets you test at the lineage level rather than the cluster level.

</details>
:::

## Step 3 — Markers for every cluster

```{r}
#| label: markers-all

nsclc.markers <- FindAllMarkers(
  nsclc.seurat.obj,
  only.pos        = TRUE,
  min.pct         = 0.25,
  logfc.threshold = 0.25
)

top_markers <- nsclc.markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 5)
top_markers
```

::: {.callout-important title="Think about it"}
1.  `only.pos = TRUE` — what do you lose?
2.  For condition-level DE (treated vs control), is `FindAllMarkers` the right tool?

<details>

<summary>Show answers</summary>

1.  You only see genes **up** in the cluster of interest. Sometimes a gene is a strong *negative* marker (consistently *lower* than elsewhere) and also diagnostic. Rerun without `only.pos` if you want both directions.
2.  No. `FindAllMarkers` / `FindMarkers` treat each cell as an independent observation, which is anti-conservative for condition-level comparisons across biological samples. For that, use **pseudobulk** + DESeq2/edgeR/limma (see the Pseudobulk extra lecture).

</details>
:::

## Step 4 — Visualizing markers

### 4a. Violin plots — expression distribution per cluster

```{r}
#| label: viz-vln

top_genes <- top_markers |> ungroup() |> slice_head(n = 6) |> pull(gene)
VlnPlot(nsclc.seurat.obj, features = top_genes, pt.size = 0, ncol = 3)
```

### 4b. Feature plots — expression on UMAP

```{r}
#| label: viz-feature

FeaturePlot(nsclc.seurat.obj, features = top_genes[1:4])
```

### 4c. Heatmap — top markers across clusters

```{r}
#| label: viz-heatmap

top10 <- nsclc.markers |> group_by(cluster) |> slice_max(avg_log2FC, n = 10)
DoHeatmap(nsclc.seurat.obj, features = top10$gene) + NoLegend()
```

### 4d. Ridge plots — expression density per cluster

```{r}
#| label: viz-ridge

RidgePlot(nsclc.seurat.obj, features = top_genes[1:4], ncol = 2)
```

### 4e. Dot plot — % expressing × mean expression

```{r}
#| label: viz-dot

DotPlot(nsclc.seurat.obj, features = top_genes) + RotatedAxis()
```

::: {.callout-important title="Think about it"}
1.  Which of the five plot types is best for **showing a 20-gene marker panel across 10 clusters on one page**?
2.  Which is best for **checking whether a single gene has spatial structure on the UMAP**?
3.  Which is most misleading if your cluster sizes are very unequal, and why?

<details>

<summary>Show answers</summary>

1.  **Dot plot** — compact, shows both percent-expressing (dot size) and mean expression (color) in a genes × clusters grid.
2.  **Feature plot** — overlays expression on the 2-D embedding so you can see geography.
3.  **Heatmap** can be misleading: it shows cells, not clusters, so a dominant cluster (say 40% of cells) visually monopolizes the plot. Dot plot normalizes per cluster and avoids this.

</details>
:::

## Step 5 — Manual annotation

```{r}
#| label: rename

# Example only — replace these with the labels appropriate for YOUR markers.
# Work through the top markers per cluster, look them up in CellMarker 2.0
# or PanglaoDB, and decide the biology.

# new.cluster.ids <- c("T cells", "Macrophages", "Epithelial",
#                      "Fibroblasts", "B cells", ...)
# names(new.cluster.ids) <- levels(nsclc.seurat.obj)
# nsclc.seurat.obj <- RenameIdents(nsclc.seurat.obj, new.cluster.ids)
# nsclc.seurat.obj$celltype_manual <- Idents(nsclc.seurat.obj)

# DimPlot(nsclc.seurat.obj, reduction = "umap", label = TRUE, repel = TRUE) +
#   NoLegend()
```

::: {.callout-important title="Think about it"}
1.  Two clusters share KRT8/KRT18 as top markers. Are they the same cell type?
2.  What's the risk of assigning a label from a single marker gene?

<details>

<summary>Show answers</summary>

1.  Not necessarily — KRT8/18 are broad epithelial markers. Both clusters are likely epithelial, but could be distinct subtypes (tumor vs. normal epithelium, alveolar type I vs. II, etc.). You need *discriminating* markers on top of the shared ones.
2.  Many markers are non-specific, and biological "marker genes" are often graded rather than binary. Label from **combinations** (a marker *plus* absence of another marker) and always cross-check with an automatic method.

</details>
:::

## Step 6 — Automatic annotation with SingleR

```{r}
#| label: singler

# library(SingleR)
# library(celldex)
#
# # Pick a reference appropriate for your tissue
# ref <- celldex::HumanPrimaryCellAtlasData()
#
# # SingleR expects log-normalized expression
# expr <- GetAssayData(nsclc.seurat.obj, slot = "data")
#
# singler.labels <- SingleR(
#   test    = expr,
#   ref     = ref,
#   labels  = ref$label.main
# )
#
# nsclc.seurat.obj$celltype_singler <- singler.labels$labels
#
# DimPlot(nsclc.seurat.obj, group.by = "celltype_singler", label = TRUE, repel = TRUE) +
#   NoLegend()
```

::: {.callout-important title="Think about it"}
1.  How would you check whether SingleR is *confident* about a label vs just picking the best of a bad set?
2.  What would you do when SingleR and your manual annotation disagree on a cluster?

<details>

<summary>Show answers</summary>

1.  Inspect `singler.labels$scores` and the `pruned.labels` column (labels it declines to assign when score gaps are small). You can also plot `plotScoreHeatmap(singler.labels)` to see per-cell confidence.
2.  Disagreements often flag **real ambiguity** — rare subtypes, transitional states, or references that don't include your biology (e.g. a tumor reference vs a healthy-tissue reference). Re-check markers, pick a more tissue-appropriate reference, and document the uncertainty.

</details>
:::

## Step 7 — Save the final object

```{r}
#| label: save

saveRDS(nsclc.seurat.obj, file = "../data/nsclc_annotated.rds")
```

## Next steps

-   **Condition-level DE** with pseudobulk (see the Pseudobulk extra lecture)
-   **Integration** across samples or batches (Harmony / CCA / scVI)
-   **Trajectory inference** with Monocle3 / Slingshot
-   **Cell–cell communication** with CellChat / LIANA

## Credits

-   Dataset: 10x Genomics — 20k NSCLC DTC, 3' v3.1 Multiplex
-   Workflow source: Khushbu Patel, single-cell RNA-seq standard workflow
-   Reference: [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html) · [SingleR](https://bioconductor.org/packages/release/bioc/html/SingleR.html)
