---
title: "Tutorial 01 — scRNA-seq: QC & Preprocessing"
subtitle: "Load a 10x dataset, filter cells, normalize, pick HVGs, scale"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

The first of a three-part series walking the standard **Seurat** scRNA-seq workflow, using the public **10x Genomics NSCLC DTC** dataset (20k cells, 7 donors, 3' v3.1 Multiplex). This part stops after **ScaleData** — the object you save here is the input to Tutorial 02.

-   Data: [20k NSCLC DTC (10x)](https://www.10xgenomics.com/resources/datasets/20-k-mixture-of-nsclc-dt-cs-from-7-donors-3-v-3-1-multiplex-3-1-standard-6-0-0)
-   Upstream script: *Khushbu Patel, single-cell RNA-seq standard workflow*
-   Reference vignette: [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)

::: callout-note
Chunks default to `eval: false`. Place the downloaded `.h5` in `../data/` (or adjust the path) and flip chunks to `eval: true` as you work through them.
:::

## Learning goals

By the end of this tutorial you will be able to:

-   Read a 10x `.h5` file into Seurat
-   Add per-cell QC metrics and visualize them
-   Choose filtering thresholds that are defensible for *your* dataset
-   Run `NormalizeData` → `FindVariableFeatures` → `ScaleData` and understand what each does

## Setup

```{r}
#| label: setup
#| eval: true

library(Seurat)
library(tidyverse)

set.seed(2026)
```

------------------------------------------------------------------------

## Step 1 — Read the 10x `.h5`

```{r}
#| label: load

nsclc.sparse.m <- Read10X_h5(filename = "../data/20k_NSCLC_DTC_3p_nextgem_Multiplex_count_raw_feature_bc_matrix.h5")

# 10x multi-modal h5 files return a list; keep the Gene Expression slice
str(nsclc.sparse.m)
cts <- nsclc.sparse.m$`Gene Expression`
dim(cts)
```

::: {.callout-important title="Think about it"}
1.  Why does `Read10X_h5` sometimes return a **list** instead of a matrix?
2.  What does each dimension of `cts` represent?

<details>

<summary>Show answers</summary>

1.  10x Multiplex / CellPlex / CITE-seq / multiome runs produce multiple measurement types in one file (e.g. `Gene Expression`, `Antibody Capture`, `Peaks`). The reader returns a named list so you can pull the slice you want.
2.  Rows = genes (features), columns = cell barcodes. `dim(cts)[1]` is the gene count, `dim(cts)[2]` is the cell count.

</details>
:::

## Step 2 — Create the Seurat object

```{r}
#| label: create-object

nsclc.seurat.obj <- CreateSeuratObject(
  counts       = cts,
  project      = "NSCLC",
  min.cells    = 3,
  min.features = 200
)
nsclc.seurat.obj
```

::: {.callout-important title="Think about it"}
1.  What exactly do `min.cells = 3` and `min.features = 200` remove?
2.  The `nsclc.sparse.m` list still sits in memory. What would `rm(nsclc.sparse.m); gc()` do, and why might you want to?

<details>

<summary>Show answers</summary>

1.  `min.cells = 3` drops genes detected in fewer than 3 cells (likely noise). `min.features = 200` drops barcodes with fewer than 200 detected genes — mostly empty droplets and broken cells. These are *permissive* first-pass filters.
2.  It releases the raw list and its memory. The full 10x raw feature matrix is large (often several GB); once counts are inside the Seurat object, the raw list is redundant.

</details>
:::

## Step 3 — Add QC metrics

```{r}
#| label: qc-metrics

nsclc.seurat.obj[["percent.mt"]] <- PercentageFeatureSet(nsclc.seurat.obj, pattern = "^MT-")
head(nsclc.seurat.obj@meta.data)
```

::: {.callout-important title="Think about it"}
1.  What biological signal does `percent.mt` capture?
2.  Why `"^MT-"` specifically? What would you use for mouse?

<details>

<summary>Show answers</summary>

1.  Dying or stressed cells leak cytoplasmic RNA while retaining mitochondrial RNA, so their mitochondrial fraction rises. High `percent.mt` flags apoptotic / compromised cells.
2.  `"^MT-"` matches human mitochondrial gene symbols (MT-ND1, MT-CO1, …). For mouse the convention is lowercase: `"^mt-"`.

</details>
:::

## Step 4 — Visualize QC

```{r}
#| label: qc-plots

VlnPlot(nsclc.seurat.obj,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3)

FeatureScatter(nsclc.seurat.obj,
               feature1 = "nCount_RNA",
               feature2 = "nFeature_RNA") +
  geom_smooth(method = "lm")
```

::: {.callout-important title="Think about it"}
1.  What does a **fat tail** on the `nFeature_RNA` violin suggest?
2.  On the `nCount_RNA` vs `nFeature_RNA` scatter, what would a doublet look like? A dying cell?

<details>

<summary>Show answers</summary>

1.  A long upper tail often means **doublets** — barcodes where two cells got trapped in one droplet, inflating both UMI count and unique gene count.
2.  Doublets sit **high on both axes**, often above the main trend line. Dying cells are **low** on both and tend to have **high `percent.mt`** (check by coloring or faceting).

</details>
:::

## Step 5 — Filter low-quality cells

```{r}
#| label: filter

nsclc.seurat.obj <- subset(
  nsclc.seurat.obj,
  subset = nFeature_RNA > 200 &
           nFeature_RNA < 2500 &
           percent.mt   < 5
)
nsclc.seurat.obj
```

::: {.callout-important title="Think about it"}
1.  The default cutoff here is `percent.mt < 5`. Is that appropriate for tumor tissue?
2.  How would you set thresholds *empirically* rather than by default?

<details>

<summary>Show answers</summary>

1.  Not always — tumor and certain primary tissues routinely have higher baseline `percent.mt` (\~10–20%). A 5% cutoff can strip out viable tumor cells. **Inspect the distribution first**; set the knee.
2.  Look at the violin/histogram shape; choose thresholds that trim the clear **tails** (not the mode). Some workshops use median ± *k* × MAD, or the `scater::isOutlier()` function for an automated rule.

</details>
:::

## Step 6 — Normalize

```{r}
#| label: normalize

nsclc.seurat.obj <- NormalizeData(nsclc.seurat.obj)
# Equivalent explicit form:
# NormalizeData(nsclc.seurat.obj,
#               normalization.method = "LogNormalize",
#               scale.factor = 10000)
```

::: {.callout-important title="Think about it"}
1.  What does `LogNormalize` do algebraically?
2.  When would you prefer `SCTransform()` instead?

<details>

<summary>Show answers</summary>

1.  For each cell: divide each gene's count by the cell's total count, multiply by `scale.factor` (default 10 000), then take `log1p`. In formula form: `log(1 + (count_gc / total_c) * 10000)`.
2.  `SCTransform` models the mean–variance relationship per gene with a regularized negative binomial, which is usually better for **low-count** datasets and more faithful downstream variance. It also replaces the NormalizeData → FindVariableFeatures → ScaleData chain in a single call.

</details>
:::

## Step 7 — Highly variable features

```{r}
#| label: hvgs

nsclc.seurat.obj <- FindVariableFeatures(
  nsclc.seurat.obj,
  selection.method = "vst",
  nfeatures = 2000
)

top10 <- head(VariableFeatures(nsclc.seurat.obj), 10)
plot1 <- VariableFeaturePlot(nsclc.seurat.obj)
LabelPoints(plot = plot1, points = top10, repel = TRUE)
```

::: {.callout-important title="Think about it"}
1.  Why do we want to focus on *variable* genes instead of using all genes?
2.  What happens if you set `nfeatures = 500`? Or `nfeatures = 10000`?

<details>

<summary>Show answers</summary>

1.  Most genes are housekeeping / ubiquitously expressed and carry little cell-type information. Selecting HVGs raises signal-to-noise for PCA and clustering and speeds everything up.
2.  Too few (500) risks missing subtle cell-type signals and is brittle to noisy gene choices. Too many (10 000) pulls in low-information genes and dilutes the signal. 1 500–3 000 is a common sweet spot; 2 000 is a good default.

</details>
:::

## Step 8 — Scale

```{r}
#| label: scale

all.genes <- rownames(nsclc.seurat.obj)
nsclc.seurat.obj <- ScaleData(nsclc.seurat.obj, features = all.genes)
```

::: {.callout-important title="Think about it"}
1.  What does `ScaleData` do to each gene numerically?
2.  Why is scaling necessary *before* PCA?
3.  How would you regress out the cell cycle or `percent.mt` here?

<details>

<summary>Show answers</summary>

1.  Centers each gene's expression to mean 0 and scales to unit variance across cells (z-scoring).
2.  PCA finds directions of maximum variance. Without scaling, highly-expressed genes (which inherently have large raw variance) would dominate PCs, regardless of whether they are biologically informative.
3.  Pass `vars.to.regress = c("percent.mt", "S.Score", "G2M.Score")` (after running `CellCycleScoring()` for the latter two). ScaleData regresses those out of each gene's expression before z-scoring.

</details>
:::

## Save and continue to Tutorial 02

```{r}
#| label: save

saveRDS(nsclc.seurat.obj, file = "../data/nsclc_preprocessed.rds")
```

::: callout-tip
The saved `.rds` is the starting point for **Tutorial 02 — Dimensionality reduction & clustering**.
:::

## Credits

-   Dataset: [10x Genomics — 20k NSCLC DTC, 3' v3.1 Multiplex](https://www.10xgenomics.com/resources/datasets/20-k-mixture-of-nsclc-dt-cs-from-7-donors-3-v-3-1-multiplex-3-1-standard-6-0-0)
-   Workflow source: Khushbu Patel, single-cell RNA-seq standard workflow
-   Reference: [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)
