---
title: "Tutorial 02 — scRNA-seq: Dimensionality Reduction & Clustering"
subtitle: "PCA, how many PCs, neighbor graph, graph clustering at multiple resolutions, UMAP — and seeing the batch effect"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

Part two of the series. We start from the **preprocessed object saved in Tutorial 01** (the `ifnb` dataset — CTRL + STIM PBMCs joined together) and walk through:


::: callout-note
**Companion book chapter:** [Chapter 2 — Dimensionality Reduction & Clustering](../Resources_Folder/Chapter_02_DimReduction.html) — the long-form prose treatment of this tutorial's material, with cross-references to the prerequisite appendices. · **HPC version (Talapas):** [Talapas 04 — DimRed & Clustering](Talapas/Tutorial_Talapas_04_DimReduction_Clustering.html)
:::


-   Linear dimensionality reduction with PCA
-   Choosing how many PCs to keep
-   Building the shared nearest-neighbor graph
-   Clustering at multiple resolutions and comparing them
-   UMAP embedding — and an explicit look at the **batch effect** between samples

::: callout-important
We deliberately do **NOT** run integration in this tutorial. Running PCA + clustering + UMAP on the unintegrated joint matrix lets you **see** what a batch effect looks like — cell types splitting into per-sample islands instead of co-clustering. Tutorial 05 will fix it. Until then, every UMAP we produce is a teaching artifact, not a final figure.
:::

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it. To run it:

1.  Download the `.qmd` source: <a href="Tutorial_02_DimReduction_Clustering.qmd.txt" download="Tutorial_02_DimReduction_Clustering.qmd">Tutorial_02_DimReduction_Clustering.qmd</a>. If your browser saves the file as `Tutorial_02_DimReduction_Clustering.qmd.txt`, **drop the trailing `.txt`** so the filename ends in `.qmd`, then open it in RStudio.
2.  Make sure you have completed **[Tutorial 01](Tutorial_01_QC_Preprocessing.html)** — it writes the `ifnb_preprocessed.rds` that this tutorial loads.
3.  Work through the chunks, flipping `eval: true` as you go.
:::

## Dataset — resuming from Tutorial 01

| File | Produced by | What it is |
|---|---|---|
| `../data/ifnb_preprocessed.rds` | Tutorial 01 | A `Seurat` object with raw + log-normalized counts for both samples (`stim ∈ {"CTRL","STIM"}`), 2,000 HVGs, scaled data, and the `seurat_annotations` ground-truth labels. |

### Data dictionary — the Seurat object at this stage

| Slot | Class | Contents |
|---|---|---|
| `seu@assays$RNA@counts` | `dgCMatrix` | Raw UMI counts (cells that survived QC only). |
| `seu@assays$RNA@data` | `dgCMatrix` | `log1p((counts / total_per_cell) * 10000)` values from `NormalizeData()`. |
| `seu@assays$RNA@scale.data` | `matrix` | z-scored expression from `ScaleData()`. Input to `RunPCA()`. |
| `seu@assays$RNA@var.features` | character | \~2,000 highly variable gene symbols. |
| `seu@meta.data$stim` | factor | `"CTRL"` or `"STIM"`. The covariate we expect to see drive PC1/PC2. |
| `seu@meta.data$seurat_annotations` | factor | Author-curated cell-type labels. |

## Learning goals

-   Run `RunPCA()` and read PCA diagnostic plots
-   Use `ElbowPlot()` to pick a defensible number of PCs
-   Run `FindNeighbors()` → `FindClusters()` at several resolutions and inspect the difference
-   Switch the active identity between clusterings with `Idents()`
-   Generate a UMAP and overlay cluster labels
-   **Diagnose a batch effect** by coloring the same UMAP by sample, by cluster, and by cell-type label


::: {.callout-warning title="Common errors / things that bite"}
**`readRDS("../data/ifnb_preprocessed.rds")` fails** — you didn't complete Tutorial 01, or your working directory isn't `Exercise_Folder/`. Run `getwd()` to check; the `..` in the path is relative to where the `.qmd` file sits. Re-run Tutorial 01 first.

**`FindClusters` is slow** — install `presto` (`install.packages("presto")`); Seurat will detect it automatically and use a much faster Wilcoxon implementation throughout.

**`DimPlot` shows the wrong cluster column** — `Idents()` defaults to `seurat_clusters`, which is the *most recent* `FindClusters()` resolution. To plot a different resolution, use `DimPlot(seu, group.by = "RNA_snn_res.0.3")` instead of relying on `Idents()`.
:::


::: {.callout-tip title="Solutions / instructor copy"}
The *Think about it* prompts on this page have inline answers (click "Show answer" / "Click for answer" to expand). For the full **runnable instructor copy** with every chunk pre-evaluated and outputs shown:

- **For students:** the same `.qmd` you downloaded already contains every solution inline — just expand the collapsed answers
- **For instructors:** render with `quarto render --profile solutions` from the `Exercise_Folder/` directory. The fully-evaluated HTML is written to `docs/Exercise_Folder/_solutions/Tutorial_02_DimReduction_Clustering.html`. See [Exercise_Folder/_quarto-solutions.yml](https://github.com/wcresko/scRNAseq_tutorial/blob/main/Exercise_Folder/_quarto-solutions.yml) for the build profile

If you're running a graded course, link the rendered solutions page from your LMS (Canvas / Moodle / etc.) — it's behind the same `_solutions/` path on the workshop's deployed site.
:::

## Setup

```{r}
#| label: setup

library(Seurat)
library(tidyverse)
set.seed(2026)
```

```{r}
#| label: resume

ifnb <- readRDS("../data/ifnb_preprocessed.rds")
ifnb
```

------------------------------------------------------------------------

## Step 1 — Linear dimensionality reduction (PCA)

```{r}
#| label: pca

ifnb <- RunPCA(
  ifnb,
  features = VariableFeatures(object = ifnb)
)

print(ifnb[["pca"]], dims = 1:5, nfeatures = 5)
DimHeatmap(ifnb, dims = 1, cells = 500, balanced = TRUE)
```

::: {.callout-important title="Think about it"}
1.  Look at the top genes loading PC1. Do you recognize any **interferon-stimulated genes** (ISG15, IFI6, IFIT1, IFIT3, ISG20, MX1, OAS1)? What does that suggest about what PC1 is encoding?
2.  PCA here runs on HVGs, not all genes. What would change if we included all genes?

<details>

<summary>Show answers</summary>

1.  PC1 in the unintegrated `ifnb` object is heavily loaded with ISGs because the IFN-β response is the largest source of variance in this dataset. **PC1 is essentially `stim` (CTRL vs STIM)** — i.e. the batch / condition axis, not a cell-type axis. That's the batch effect we'll fix in Tutorial 05.
2.  PCs would be dominated by highly-expressed housekeeping genes whose variance simply reflects sequencing depth. HVG-restricted PCA isolates biologically-variable signal.

</details>
:::

## Step 2 — Choose how many PCs to keep

```{r}
#| label: elbow

ElbowPlot(ifnb, ndims = 40)
```

::: {.callout-important title="Think about it"}
1.  Where is "the elbow" on an ElbowPlot, and why does it suggest a cutoff?
2.  Is it risky to choose too *few* PCs? Too *many*?

<details>

<summary>Show answers</summary>

1.  The elbow is where the variance explained per PC flattens out — additional PCs stop adding much signal. Picking the elbow captures most of the biological structure without the noise-heavy tail.
2.  **Too few** loses cell-type resolution (rare populations blur together). **Too many** pulls in noisy dimensions and inflates distances in the neighbor graph. Being slightly generous (e.g. 15–30) is usually safe; for `ifnb` the elbow sits around PC15–PC20.

</details>
:::

```{r}
#| label: n-pcs

n_pcs <- 20
```

## Step 3 — Build the neighbor graph

```{r}
#| label: neighbors

ifnb <- FindNeighbors(ifnb, dims = 1:n_pcs)
```

::: {.callout-important title="Think about it"}
1.  What graph does `FindNeighbors()` build?
2.  Why operate on PC space rather than raw expression?

<details>

<summary>Show answers</summary>

1.  It builds a **k-nearest-neighbor** graph (default `k = 20`) and then a **shared-nearest-neighbor** graph: two cells are connected if they share many nearest neighbors. The Jaccard-weighted SNN graph is what clustering runs on.
2.  PC space is denoised, low-dimensional, and has well-behaved Euclidean distance. Raw expression is sparse, high-dimensional, and distance-distorted by the *curse of dimensionality*.

</details>
:::

## Step 4 — Cluster at several resolutions

```{r}
#| label: cluster

ifnb <- FindClusters(
  ifnb,
  resolution = c(0.1, 0.3, 0.5, 0.7, 1.0)
)

head(ifnb@meta.data)
```

::: callout-note
`FindClusters()` here uses Seurat's default **Louvain** algorithm (`algorithm = 1`). Lecture 02 introduces **Leiden** (`algorithm = 4`) as the modern preferred method; we use the Louvain default for portability because Leiden needs the `leidenalg` Python package via `reticulate`. To switch, add `algorithm = 4` to the call above once that dependency is installed.
:::

::: {.callout-important title="Think about it"}
1.  What does `resolution` control intuitively?
2.  Each resolution creates its own `RNA_snn_res.XYZ` column in the metadata. How do you make one of them the *active* identity?

<details>

<summary>Show answers</summary>

1.  The granularity knob for the graph-based community detection. Higher resolution ⇒ smaller, more numerous clusters.
2.  `Idents(ifnb) <- "RNA_snn_res.0.5"` — or any of the resolution columns. `DimPlot(..., group.by = "RNA_snn_res.0.5")` also works without changing active idents.

</details>
:::

## Step 5 — UMAP

```{r}
#| label: umap

ifnb <- RunUMAP(ifnb, dims = 1:n_pcs)

Idents(ifnb) <- "RNA_snn_res.0.5"
DimPlot(ifnb, reduction = "umap", label = TRUE) + ggtitle("Unintegrated — clusters")
```

## Step 6 — Diagnose the batch effect

This is the key exercise. The UMAP above is colored by cluster. Now color the *same UMAP* by **sample** and by **author cell-type label** and compare.

```{r}
#| label: batch-diagnostic

p_by_sample <- DimPlot(ifnb, reduction = "umap", group.by = "stim") +
  ggtitle("Unintegrated — by sample (stim)")

p_by_celltype <- DimPlot(ifnb, reduction = "umap", group.by = "seurat_annotations",
                         label = TRUE, repel = TRUE) +
  NoLegend() +
  ggtitle("Unintegrated — by author cell-type")

p_by_sample
p_by_celltype
```

::: {.callout-important title="Think about it"}
1.  In the **by-sample** plot, do the CTRL and STIM cells overlap, or do they form parallel islands?
2.  In the **by-cell-type** plot, does each cell type form one cluster or several? If a cell type splits, how does its split correlate with sample?
3.  Cross-tabulate: `table(ifnb$RNA_snn_res.0.5, ifnb$stim)`. Are any clusters mostly CTRL? Mostly STIM?
4.  Based on these three diagnostics, do you need to integrate?

<details>

<summary>Show answers</summary>

1.  CTRL and STIM cells form **largely parallel structures**. Many cell-type clusters appear twice — once for CTRL, once for STIM — because the IFN-β response shifts every cell along PC1.
2.  Most cell types **split into two sister clusters**, with the split tracking `stim`. T cells appear once on the CTRL side and once on the STIM side; same for monocytes, B cells, NK cells, etc. This is the classic "per-sample islands" pattern — the batch / condition axis dominates the embedding.
3.  Most clusters at `res = 0.5` will be \>80% one sample. Some are \>95% one sample.
4.  Yes. The batch axis is dominating PC1 and is splitting biologically identical cell types across two clusters. **Tutorial 05** will integrate the data with Harmony and Seurat anchors, and you'll see the same cells re-arrange into shared, sample-mixed clusters.

</details>
:::

## Step 7 — Cross-tab clusters vs cell types vs samples

```{r}
#| label: crosstab

# Clusters × samples — how mixed is each cluster across CTRL / STIM?
prop.table(table(cluster = ifnb$RNA_snn_res.0.5, sample = ifnb$stim), margin = 1) |>
  round(2)

# Clusters × cell types — how clean is each cluster's biological identity?
table(cluster = ifnb$RNA_snn_res.0.5, celltype = ifnb$seurat_annotations)
```

::: {.callout-important title="Think about it"}
1.  A "good" cluster on integrated data would be \~50/50 CTRL/STIM (because both samples have all cell types) and dominated by a single cell-type label. Are any of your unintegrated clusters like that already? Why might some be more sample-mixed than others?
2.  What's the lowest-abundance cell-type label in `seurat_annotations`? Does it split across two clusters or stay together?

<details>

<summary>Show answers</summary>

1.  A few clusters — typically very small populations like dendritic cells (DC) or megakaryocytes (Mk) — may already mix CTRL/STIM, because their absolute interferon response is small enough that other biology dominates their distance to PC1. Most clusters won't.
2.  Often DC, Mk, or eryth (\<2% each). Whether they split tells you something about how strong the IFN response is in those cells; rare cell types with weak ISG responses sometimes survive the unintegrated UMAP intact, while big subsets like CD14 monocytes split cleanly along the CTRL/STIM axis.

</details>
:::

## Save and continue to Tutorial 03

```{r}
#| label: save

saveRDS(ifnb, file = "../data/ifnb_clustered.rds")
```

::: callout-tip
Continue with **Tutorial 03 — Markers & Cell Type Annotation**. We'll do annotation on the unintegrated clusters first (it's still possible — markers are cell-type-driven even when the embedding is sample-driven), then in **Tutorial 05** we'll redo PCA + clustering on the integrated representation and watch the embedding reorganize.
:::

## Credits

-   Dataset: `ifnb` from the `SeuratData` package — Kang *et al.* (2017) *Nat Biotechnol* 36: 89–94
-   Reference vignettes: [Seurat integration tutorial](https://satijalab.org/seurat/articles/integration_introduction.html) · [clustree](https://cran.r-project.org/package=clustree)
-   Inspiration: [scNotebooks](https://integrativebioinformatics.github.io/scNotebooks/)
