---
title: "Tutorial 04 — Pseudobulk DE with DESeq2"
subtitle: "Aggregate to (sample × cell type), then run a proper bulk-style DE test"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

Part four of the series, paired with **[Lecture 05 --- Downstream analyses](../Lecture_Folder/Lecture_05_Downstream.html)**. We start from the **annotated object saved at the end of Tutorial 03** and learn to:

-   Aggregate single-cell counts to a *(sample × cell type)* pseudobulk matrix
-   Run a proper, well-calibrated DE test with `DESeq2`, one cell type at a time
-   Apply LFC shrinkage for ranking and visualize results
-   Avoid the most common pseudobulk pitfalls

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it. To run it:

1.  [Download the `.qmd` source](Tutorial_04_Pseudobulk_DE.qmd) and open it in RStudio.
2.  Make sure you've completed **[Tutorial 03](Tutorial_03_Markers_Annotation.html)** --- it writes `nsclc_annotated.rds`, which this tutorial loads.
3.  Work through the chunks, flipping `eval: true` as you go.
:::

## Dataset --- resuming from Tutorial 03

| File | Produced by | What it is |
|---|---|---|
| `../data/nsclc_annotated.rds` | Tutorial 03 | Seurat object with cluster labels, manual cell-type assignments, and (optionally) `SingleR` predictions in `seu$celltype`. |

### A note on the contrast we'll test

The 20k NSCLC dataset comes from **seven donors**. There is no built-in treatment vs. control contrast, so for teaching purposes we'll define a **synthetic contrast** by splitting donors into two groups. *In your own work this will be your real biological covariate (treatment / genotype / time-point).*

## Setup

```{r}
#| label: setup
library(Seurat)
library(DESeq2)
library(tidyverse)
library(patchwork)

set.seed(2026)

seu <- readRDS("../data/nsclc_annotated.rds")
seu
```

Inspect the relevant metadata before going further:

```{r}
table(seu$celltype, seu$sample_id)
```

If your column names are different, adjust the rest of the tutorial accordingly. The two columns we need are:

-   `seu$celltype` --- the per-cell label from Tutorial 03
-   `seu$sample_id` --- the donor / library / sample of origin (this is what makes pseudobulk **biologically replicated**)

## Step 1 --- assign a synthetic condition

Split the seven donors into a "groupA" / "groupB" assignment so we have a two-level contrast.

```{r}
sample_meta <- tibble(
  sample_id = unique(seu$sample_id),
  condition = rep(c("groupA","groupB"), length.out = length(unique(seu$sample_id)))
)
seu$condition <- sample_meta$condition[match(seu$sample_id, sample_meta$sample_id)]
table(seu$condition, seu$sample_id)
```

::: {.callout-important title="Think about it"}
Why is it not enough to just run `FindMarkers(seu, ident.1 = "groupA", ident.2 = "groupB")` on the per-cell expression matrix?

<details>

<summary>Click for answer</summary>

Because cells from one donor are **not statistically independent**. Running a per-cell test treats every cell as a separate replicate, which inflates the effective sample size and produces anti-conservative p-values. The number of independent biological replicates is the **number of donors**, not the number of cells. Pseudobulk fixes this by aggregating to one observation per (sample × cell type).

</details>
:::

## Step 2 --- aggregate to pseudobulk

```{r}
pb <- AggregateExpression(
  seu,
  assays    = "RNA",
  slot      = "counts",
  group.by  = c("sample_id", "celltype"),
  return.seurat = FALSE
)$RNA

dim(pb)               # genes × (sample × celltype) replicates
head(colnames(pb), 10)
```

The column names look like `<sample_id>_<celltype>`. Build the matching `colData`:

```{r}
meta_pb <- tibble(
  group_id  = colnames(pb),
  sample_id = sub("_.*",  "", colnames(pb)),
  celltype  = sub("^[^_]+_", "", colnames(pb))
) |>
  left_join(sample_meta, by = "sample_id")

head(meta_pb)
```

## Step 3 --- filter low-count groups

A pseudobulk column built from very few cells is noise. Drop them.

```{r}
# Count cells in each (sample × celltype) combo
cells_per_group <- seu@meta.data |>
  count(sample_id, celltype) |>
  unite(group_id, sample_id, celltype, sep = "_")

meta_pb <- meta_pb |>
  left_join(cells_per_group, by = "group_id") |>
  rename(n_cells = n)

# Keep groups with at least 10 cells
keep <- meta_pb$n_cells >= 10
pb       <- pb[, keep]
meta_pb  <- meta_pb[keep, ]

table(meta_pb$celltype)
```

::: {.callout-important title="Think about it"}
Why is `n_cells >= 10` reasonable? Why might you raise or lower it?

<details>

<summary>Click for answer</summary>

The threshold balances signal against noise: pseudobulk built from 1--2 cells is dominated by stochastic dropout and is unlikely to represent the cell type. A floor of \~10 cells is the de-facto default in the field (see Squair et al. 2021 and Crowell et al. 2020). Raise it (e.g. 20--50) for very sparse cell types or shallow libraries; lower it only if you also lower your statistical-significance expectations.

</details>
:::

## Step 4 --- run DESeq2 per cell type

Wrap the test in a function and `map_dfr` over cell types:

```{r}
run_de <- function(ct) {
  keep <- meta_pb$celltype == ct
  if (sum(keep) < 4 || length(unique(meta_pb$condition[keep])) < 2) {
    return(NULL)   # not enough replicates for a contrast
  }
  dds <- DESeqDataSetFromMatrix(
    countData = pb[, keep],
    colData   = meta_pb[keep, ],
    design    = ~ condition
  )
  dds <- DESeq(dds, quiet = TRUE)
  res <- results(dds, contrast = c("condition","groupB","groupA"))
  res <- lfcShrink(dds, coef = "condition_groupB_vs_groupA",
                   res = res, type = "apeglm")
  as.data.frame(res) |>
    rownames_to_column("gene") |>
    mutate(celltype = ct)
}

de_all <- map_dfr(unique(meta_pb$celltype), run_de) |>
  filter(!is.na(padj))
glimpse(de_all)
```

## Step 5 --- inspect the top hits

```{r}
de_top <- de_all |>
  group_by(celltype) |>
  slice_min(padj, n = 10) |>
  ungroup()
de_top
```

## Step 6 --- visualize

A simple per-cell-type volcano plot:

```{r}
de_all |>
  filter(celltype %in% c("T_CD4","B_cell","Macrophage")) |>   # adjust to your labels
  ggplot(aes(log2FoldChange, -log10(padj))) +
  geom_point(alpha = 0.4, size = 1) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  facet_wrap(~ celltype) +
  labs(title = "Pseudobulk DE: groupB vs groupA",
       x = "log2 FC (shrunken)", y = "-log10(padj)")
```

::: {.callout-important title="Think about it"}
Some cell types will show many DE genes; others almost none. What can drive this pattern *that has nothing to do with biology*?

<details>

<summary>Click for answer</summary>

Three usual suspects:

1.  **Cell counts per group.** Cell types that are abundant in every donor have well-estimated dispersions; rare types do not.
2.  **Sample size.** With 7 donors split 4-vs-3, you have very little power even on a clean signal --- and almost none if a cell type is missing in 1--2 donors.
3.  **Library composition.** A cell type whose library is dominated by a few hyper-expressed genes will have noisier dispersion estimates and fewer detected hits.

Always inspect `n_cells` per group and the dispersion plot before reporting "X genes are significantly DE".

</details>
:::

## Step 7 --- save results

```{r}
write_csv(de_all,  "../data/nsclc_pseudobulk_de.csv")
write_csv(de_top,  "../data/nsclc_pseudobulk_de_top.csv")
```

## Wrap-up

You now have:

-   A pseudobulk matrix per *(sample, cell type)*.
-   Per-cell-type DE results from `DESeq2` with shrunken LFCs.
-   A workflow you can re-point at *any* real biological contrast in your own data.

::: {.callout-tip title="Going further"}
-   Add covariates (batch, sex, age) to the design: `~ batch + condition`.
-   Use `IHW` or `ashr` for adaptive multiple-testing.
-   Hand the gene lists to `clusterProfiler::compareCluster` for cross-cell-type enrichment.
-   Compare against `muscat::pbDS` (formal pseudobulk wrapper) and `glmGamPoi`.
:::

## See also

-   [Lecture 05 --- Downstream analyses](../Lecture_Folder/Lecture_05_Downstream.html)
-   [Lecture (Extra) --- Pseudobulk DE deep dive](../Lecture_Folder/Lecture_Extra_Pseudobulk.html)
-   Squair et al. 2021 --- the paper that made pseudobulk the default ([doi](https://doi.org/10.1038/s41467-021-25960-2))
-   Crowell et al. 2020 --- `muscat` package paper ([doi](https://doi.org/10.1038/s41467-020-19894-4))
