---
title: "Tutorial: Seurat Basic scRNA-seq Workflow"
subtitle: "From raw counts to annotated clusters (GSE158130 neuroblastoma SK-N-SH)"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

This walk-through follows the standard **Seurat** basic workflow, step by step. Each section below is a self-contained code chunk you can run in order in RStudio. The workflow mirrors the [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html), applied here to the neuroblastoma SK-N-SH dataset from **GEO accession [GSE158130](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE158130)**.

It is adapted from the pipeline in [kpatel427/scRNAseq](https://github.com/kpatel427/scRNAseq) (Khushbu Patel, `scRNASeq_pipeline.R`), which itself follows the Seurat PBMC tutorial.

::: callout-note
**Chunks are set to `eval: false`** so the document renders without running the analysis. Set chunk-level `eval: true` (or flip the document default) as you work through each step interactively.
:::

## Learning goals

By the end of this tutorial you will be able to:

-   Create a **Seurat object** from a counts matrix and filter genes/cells
-   Run **QC** and remove low-quality cells
-   **Normalize**, select **variable features**, and **scale** the data
-   Run **PCA** and decide how many components to keep
-   Build a **neighbor graph** and run graph-based **clustering**
-   Visualize clusters with **UMAP**
-   Find **cluster marker genes** and make the canonical plots
-   **Save** the processed object for downstream work

## Setup

```{r}
#| label: setup
#| eval: true

library(Seurat)
library(ggplot2)
library(patchwork)
library(tidyverse)

# Reproducible seeds for any stochastic steps (UMAP, JackStraw)
set.seed(2026)
```

------------------------------------------------------------------------

## Step 1 — Load raw counts

Download the counts matrix from GEO ([GSE158130 supplementary files](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE158130)) and place it in a local `data/` folder before running this chunk. The original pipeline stores it as either an `.RDS` or a space-separated `.txt` file.

```{r}
#| label: load-counts

# Option A: pre-serialized RDS
# counts <- readRDS("data/GSE158130_SK-N-SH_counts.RDS")

# Option B: plain-text counts matrix (rows = genes, columns = cells)
counts <- read.delim(
  "data/GSE158130_SK-N-SH_counts.txt",
  header = TRUE, sep = " "
)

dim(counts)
counts[1:5, 1:5]
```

## Step 2 — Create the Seurat object

Include only genes detected in **≥3 cells**, and only cells with **≥200 detected genes**. This is a permissive first pass; the stricter QC comes next.

```{r}
#| label: create-seurat

gse <- CreateSeuratObject(
  counts      = counts,
  project     = "gse158130",
  min.cells   = 3,
  min.features = 200
)
gse
```

## Step 3 — Per-cell quality control

### 3a. Compute percent mitochondrial

Dying or stressed cells tend to have high mitochondrial read fractions.

```{r}
#| label: qc-mito
gse[["percent.mt"]] <- PercentageFeatureSet(gse, pattern = "^MT-")
head(gse@meta.data, 5)
```

### 3b. Visualize QC metrics

```{r}
#| label: qc-violin
VlnPlot(gse,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3)
```

### 3c. Look at metric relationships

The `nCount_RNA` vs `nFeature_RNA` scatter should be roughly monotonic — deviations can flag doublets or empty droplets.

```{r}
#| label: qc-scatter
p1 <- FeatureScatter(gse, feature1 = "nCount_RNA", feature2 = "percent.mt")
p2 <- FeatureScatter(gse, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
p1 + p2
```

### 3d. Apply filters

Keep cells with 200 < `nFeature_RNA` < 2500 and `percent.mt` < 5. Always look at distributions first and adjust thresholds for your own dataset.

```{r}
#| label: qc-filter
gse <- subset(gse,
              subset = nFeature_RNA > 200 &
                       nFeature_RNA < 2500 &
                       percent.mt   < 5)
gse
```

## Step 4 — Normalization

Log-normalize with a scale factor of 10,000 (the Seurat default).

```{r}
#| label: normalize
gse <- NormalizeData(gse,
                     normalization.method = "LogNormalize",
                     scale.factor = 10000)
```

::: callout-tip
**Alternative:** `SCTransform()` models the mean-variance relationship and is often a stronger choice, especially for low-count cells. It replaces steps 4–6 in a single call.
:::

## Step 5 — Identify highly variable features (HVGs)

```{r}
#| label: hvgs
gse <- FindVariableFeatures(gse,
                            selection.method = "vst",
                            nfeatures = 2000)

top10 <- head(VariableFeatures(gse), 10)

p3 <- VariableFeaturePlot(gse)
p4 <- LabelPoints(plot = p3, points = top10, repel = TRUE)
p3 + p4
```

## Step 6 — Scale the data

Z-scores each gene across cells. Required before PCA.

```{r}
#| label: scale
all.genes <- rownames(gse)
gse <- ScaleData(gse, features = all.genes)
```

::: callout-note
To also regress out unwanted sources of variation:

``` r
gse <- ScaleData(gse, features = all.genes,
                 vars.to.regress = c("percent.mt", "nCount_RNA"))
```
:::

## Step 7 — Linear dimensionality reduction (PCA)

```{r}
#| label: pca
gse <- RunPCA(gse, features = VariableFeatures(gse))

# Inspect loadings on the first few PCs
print(gse[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(gse, dims = 1:2, reduction = "pca")
DimPlot(gse, reduction = "pca")
DimHeatmap(gse, dims = 1:5, cells = 500, balanced = TRUE)
```

## Step 8 — Choose how many PCs to keep

Two complementary tools:

-   **JackStraw**: permutation test, slow but principled
-   **ElbowPlot**: fast visual heuristic; look for the point where additional PCs stop adding signal

```{r}
#| label: pc-selection

# JackStraw (can be slow; subset dims to speed it up)
gse <- JackStraw(gse, num.replicate = 100)
gse <- ScoreJackStraw(gse, dims = 1:20)
JackStrawPlot(gse, dims = 1:15)

ElbowPlot(gse, ndims = 40)
```

For this dataset, both plots suggest ~**15 PCs** capture the bulk of the signal. Use that downstream.

```{r}
#| label: pc-chosen
n_pcs <- 15
```

## Step 9 — Cluster the cells

Build the shared-nearest-neighbor graph from the chosen PCs, then run graph-based community detection.

```{r}
#| label: cluster
gse <- FindNeighbors(gse, dims = 1:n_pcs)
gse <- FindClusters(gse, resolution = 0.5)

head(Idents(gse), 5)
```

::: callout-tip
`resolution` controls granularity. Try 0.4–1.2 and inspect the result. Higher = more clusters.
:::

## Step 10 — Non-linear embedding (UMAP)

```{r}
#| label: umap
gse <- RunUMAP(gse, dims = 1:n_pcs)
DimPlot(gse, reduction = "umap", label = TRUE)
```

## Step 11 — Find cluster marker genes

### 11a. Markers for a single cluster

```{r}
#| label: markers-single
cluster1.markers <- FindMarkers(gse, ident.1 = 2, min.pct = 0.25)
head(cluster1.markers, 5)
```

### 11b. Markers distinguishing one cluster from a specific pair

```{r}
#| label: markers-vs
cluster5.markers <- FindMarkers(gse,
                                ident.1 = 5,
                                ident.2 = c(0, 3),
                                min.pct = 0.25)
head(cluster5.markers, 5)
```

### 11c. Markers for every cluster (positive only)

```{r}
#| label: markers-all
gse.markers <- FindAllMarkers(gse,
                              only.pos = TRUE,
                              min.pct = 0.25,
                              logfc.threshold = 0.25)

top2_per_cluster <- gse.markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 2)
top2_per_cluster
```

## Step 12 — Visualize marker expression

### Violin plot — distribution per cluster

```{r}
#| label: viz-violin
VlnPlot(gse, features = c("HMGB2", "TK1"), slot = "counts", log = TRUE)
```

### Feature plot — expression on UMAP

```{r}
#| label: viz-feature
top_genes <- top2_per_cluster |> pull(gene) |> head(5)
FeaturePlot(gse, features = top_genes)

# Co-expression of two markers
FeaturePlot(gse, features = top_genes[1:2], blend = TRUE)
```

### Heatmap — top markers across clusters

```{r}
#| label: viz-heatmap
top10 <- gse.markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 10)

DoHeatmap(gse, features = top10$gene) + NoLegend()
```

### Ridge plot — expression distributions

```{r}
#| label: viz-ridge
RidgePlot(gse, features = top_genes, ncol = 2)
```

### Dot plot — per-cluster mean expression + % expressing

```{r}
#| label: viz-dot
DotPlot(gse, features = top_genes) + RotatedAxis()
```

## Step 13 — Assign cell-type identities (optional)

Once you have markers, you can re-label clusters. The example below shows the pattern — **replace the labels with the cell types appropriate for your dataset.**

```{r}
#| label: rename-clusters

# new.cluster.ids <- c("ClusterA", "ClusterB", "ClusterC", ...)
# names(new.cluster.ids) <- levels(gse)
# gse <- RenameIdents(gse, new.cluster.ids)
# DimPlot(gse, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

::: callout-tip
See the Annotation section of Lecture 01 for **automatic** alternatives: `SingleR`, `Azimuth`, `CellTypist`, `scType`.
:::

## Step 14 — Save the processed object

```{r}
#| label: save
saveRDS(gse, file = "data/gse158130_final.rds")
```

------------------------------------------------------------------------

## Where to go next

-   **Integration** across samples/batches — see Lecture 01 (Harmony, CCA/RPCA, scVI)
-   **Condition-level DE** — see the Pseudobulk extra lecture
-   **Enrichment / pathway analysis** — `clusterProfiler`, `fgsea`, `enrichR`
-   **Gene co-expression modules** — `hdWGCNA`
-   **Trajectory / pseudotime** — `Monocle3`, `Slingshot`

## Credits &amp; references

-   [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html) — the canonical walkthrough this workflow mirrors
-   [kpatel427/scRNAseq](https://github.com/kpatel427/scRNAseq) — pipeline adapted here to GSE158130
-   GEO [GSE158130](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE158130) — neuroblastoma SK-N-SH cells
