---
title: "Tutorial 09 — Reference-based Annotation (SingleR + Azimuth)"
subtitle: "Map cells against a labeled reference for fast, reproducible cell-type calls"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

A short hands-on companion to **[Lecture 07 — Reference-based Annotation](../Lecture_Folder/Lecture_09_Reference_Annotation.html)**. We start from the **integrated, annotated object saved at the end of Tutorial 04** (`ifnb_integrated.rds`) and:

-   Run `SingleR` against a built-in `celldex` reference
-   Run `Azimuth` against the PBMC reference (a perfect fit for this dataset)
-   Compare the two automatic calls against the manual labels you assigned in Tutorial 03 *and* against the author-curated `seurat_annotations` ground truth
-   Reconcile disagreements and save a final, reference-validated annotation

::: callout-note
**Why this works particularly cleanly for `ifnb`.** The dataset is human PBMCs — exactly what the Azimuth `pbmcref` and `celldex::HumanPrimaryCellAtlasData()` references are built for. You should expect **high agreement** between methods. That's a feature, not a coincidence: it's the easy regime, where reference-based annotation shines. In Tutorial 04 of the **NSCLC scaling exercise** (`Exercise_Folder/NSCLC_FullDataset/`) you'll see what happens when the dataset includes tumor / non-immune cells outside the reference's support — the harder regime.
:::

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it. To run it:

1.  Download the `.qmd` source: <a href="Tutorial_09_Reference_Annotation.qmd.txt" download="Tutorial_09_Reference_Annotation.qmd">Tutorial_09_Reference_Annotation.qmd</a>. If your browser saves the file as `Tutorial_09_Reference_Annotation.qmd.txt`, **drop the trailing `.txt`** so the filename ends in `.qmd`, then open it in RStudio.
2.  Make sure you've completed **[Tutorial 04](Tutorial_04_Integration.html)** first.
3.  Work through the chunks, flipping `eval: true` as you go.
:::


::: {.callout-warning title="Common errors / things that bite"}
**`celldex::HumanPrimaryCellAtlasData()` fails or hangs** — the underlying data is in `ExperimentHub`. First-time download is \~500 MB and can hit firewalls. Run `setExperimentHubOption("PROXY", Sys.getenv("HTTP_PROXY"))` if behind a corporate proxy. The cached version is reused on subsequent runs.

**`RunAzimuth` errors during reference download** — the PBMC reference is \~1 GB and downloads on first call. Confirm internet, sufficient disk, and no `tar` corruption. If it fails midway, delete the partial cache directory (`~/.cache/Azimuth/` or similar) and retry.

**SingleR labels are mostly "T_cells", "B_cell"** — the `HumanPrimaryCellAtlasData` reference is broad. For finer resolution, use `label.fine` instead of `label.main`, or switch to a more specific reference like `MonacoImmuneData` for PBMCs.
:::


::: {.callout-tip title="Solutions / instructor copy"}
The *Think about it* prompts on this page have inline answers (click "Show answer" / "Click for answer" to expand). For the full **runnable instructor copy** with every chunk pre-evaluated and outputs shown:

- **For students:** the same `.qmd` you downloaded already contains every solution inline — just expand the collapsed answers
- **For instructors:** render with `quarto render --profile solutions` from the `Exercise_Folder/` directory. The fully-evaluated HTML is written to `docs/Exercise_Folder/_solutions/Tutorial_09_Reference_Annotation.html`. See [Exercise_Folder/_quarto-solutions.yml](https://github.com/wcresko/scRNAseq_tutorial/blob/main/Exercise_Folder/_quarto-solutions.yml) for the build profile

If you're running a graded course, link the rendered solutions page from your LMS (Canvas / Moodle / etc.) — it's behind the same `_solutions/` path on the workshop's deployed site.
:::

## Setup

```{r}
#| label: setup
library(Seurat)
library(SingleR)
library(celldex)
library(SingleCellExperiment)
library(tidyverse)
library(patchwork)

set.seed(2026)

ifnb <- readRDS("../data/ifnb_integrated.rds")
DefaultAssay(ifnb) <- "RNA"

DimPlot(ifnb, group.by = "seurat_annotations", reduction = "umap_harmony",
        label = TRUE, repel = TRUE) + NoLegend()
```

## Path 1 — `SingleR` against a `celldex` reference

`SingleR` correlates each query cell's expression against per-label reference profiles and returns the best-matching label.

### Pick a reference

```{r}
#| label: singler-ref
# Bulk reference covering 37 main cell types from purified human samples
ref <- celldex::HumanPrimaryCellAtlasData()

# Run SingleR on the (log-)normalized counts
sce  <- as.SingleCellExperiment(ifnb)
pred <- SingleR(test = sce, ref = ref,
                labels = ref$label.main,
                assay.type.test = "logcounts",
                BPPARAM = BiocParallel::SerialParam())
table(pred$labels)
```

Attach the calls back to the Seurat object:

```{r}
#| label: singler-attach

ifnb$singler_label <- pred$labels
ifnb$singler_score <- apply(pred$scores, 1, max)

DimPlot(ifnb, group.by = "singler_label", reduction = "umap_harmony",
        label = TRUE, repel = TRUE) + NoLegend()
```

### Score-distribution plot — a built-in confidence check

```{r}
#| label: singler-confidence

plotScoreHeatmap(pred)
plotDeltaDistribution(pred)
```

::: {.callout-important title="Think about it"}
Some cells will get a label like `"T_cells"` with `delta` close to zero. What does that mean and what should you do about it?

<details>

<summary>Click for answer</summary>

`delta` is the gap between the cell's top-label score and the median score across all labels. A small `delta` means the reference can't decide — the cell either looks ambiguous (multiple types fit equally well) or is a type the reference doesn't contain.

In practice: prune low-confidence calls (`pruneScores()` from SingleR), and trust your manual marker-based label more for those cells.

</details>
:::

```{r}
#| label: singler-prune

ifnb$singler_label_pruned <- ifelse(
  is.na(SingleR::pruneScores(pred)), NA, pred$labels
)
```

## Path 2 — Azimuth (Seurat reference mapping)

Azimuth wraps Seurat's anchor-based reference mapping with curated atlases. The PBMC reference is the most widely tested.

```{r}
#| label: azimuth
# install if needed
# remotes::install_github("satijalab/azimuth")
library(Azimuth)

# Will download the PBMC reference on first use (~1 GB)
ifnb <- RunAzimuth(ifnb, reference = "pbmcref")

# New columns: predicted.celltype.l1 / l2 / l3 + scores
head(ifnb@meta.data[, grep("predicted", colnames(ifnb@meta.data))])

DimPlot(ifnb, group.by = "predicted.celltype.l2", reduction = "umap_harmony",
        label = TRUE, repel = TRUE) + NoLegend()
```

::: {.callout-important title="Think about it"}
Azimuth's PBMC reference is built for **resting** PBMCs. The `ifnb` STIM cells are **interferon-β stimulated**. Will Azimuth confidently place STIM cells, or might it struggle? How would you spot a problem?

<details>

<summary>Click for answer</summary>

Azimuth still assigns the best-matching label, but the **mapping score** (`mapping.score`) reflects how well the cell's neighborhood projects into the reference space. Plot `FeaturePlot(ifnb, "mapping.score")` — if STIM cells have **systematically lower mapping scores** than CTRL cells, that's evidence that the IFN response is shifting STIM cells off the reference manifold even though their *cell-type* identity is unchanged.

In practice the PBMC reference is robust enough that this effect is mild — but it's a great teaching example of how a "good" reference can still under-represent a perturbation state.

</details>
:::

## Compare the three annotations

```{r}
#| label: compare

ann <- ifnb@meta.data |>
  select(seurat_annotations,                  # ground truth (Kang authors)
         singler_label_pruned,                # SingleR (pruned)
         predicted.celltype.l2)               # Azimuth

table(ann$seurat_annotations, ann$singler_label_pruned, useNA = "ifany")
table(ann$seurat_annotations, ann$predicted.celltype.l2, useNA = "ifany")
```

A cleaner cross-tabulation as a heatmap:

```{r}
#| label: compare-heatmap

ann |>
  count(seurat_annotations, predicted.celltype.l2) |>
  ggplot(aes(predicted.celltype.l2, seurat_annotations, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Ground truth (rows) vs. Azimuth (cols)",
       x = NULL, y = NULL)
```

## Reconcile disagreements

A simple "agree-or-flag" rule: keep the ground-truth label if at least one automatic method agrees; otherwise mark for review.

```{r}
#| label: reconcile

ifnb$celltype_final <- with(ifnb@meta.data, dplyr::case_when(
  as.character(seurat_annotations) == singler_label_pruned     ~ as.character(seurat_annotations),
  as.character(seurat_annotations) == predicted.celltype.l2    ~ as.character(seurat_annotations),
  TRUE                                                         ~ paste0("REVIEW:", seurat_annotations)
))

table(ifnb$celltype_final)
```

::: {.callout-important title="Think about it"}
A reviewer asks you to "use Azimuth labels instead of your manual ones because they're more reproducible". How do you respond?

<details>

<summary>Click for answer</summary>

Reproducibility is necessary but not sufficient. Azimuth is reproducible *and* biased by its reference: any cell type absent from the reference will be silently re-labeled. The right answer is to use both, as the tutorial does, and document where they disagree — not to pick one because it's mechanical.

</details>
:::

## Save the final annotated object

```{r}
#| label: save

saveRDS(ifnb, "../data/ifnb_annotated_final.rds")
```

## Wrap-up

You now have:

-   A `SingleR` label per cell (with confidence pruning)
-   An Azimuth label per cell (with a mapping score)
-   A reconciled `celltype_final` you can carry into Tutorial 06 (pseudobulk DE) or downstream analyses

::: {.callout-tip title="Going further"}
-   Try a tissue-appropriate reference: `celldex::BlueprintEncodeData()`, the [Tabula Sapiens](https://tabula-sapiens-portal.ds.czbiohub.org/) atlas, or the [HCA Lung Cell Atlas](https://data.humancellatlas.org/).
-   Repeat with **CellTypist** (Python) for a third opinion.
-   For tumor data, look at **TumorDecon** or **ProjecTILs** to handle the tumor/immune mix more carefully — see also the parallel reference-annotation example in the **NSCLC scaling exercise**.
:::

## See also

-   [Lecture 07 — Reference-based Annotation](../Lecture_Folder/Lecture_09_Reference_Annotation.html)
-   [Lecture 03 — Markers & Manual Annotation](../Lecture_Folder/Lecture_03_Markers_Annotation.html)
-   [Glossary — Cell type annotation](../Resources_Folder/Glossary.html#c)
-   `SingleR` book: [bioconductor.org/books/release/SingleRBook](https://bioconductor.org/books/release/SingleRBook/)
-   Azimuth web tool: [azimuth.hubmapconsortium.org](https://azimuth.hubmapconsortium.org/)
