---
title: "Tutorial 02 — scRNA-seq: Dimensionality Reduction & Clustering"
subtitle: "PCA, how many PCs, neighbor graph, Leiden at multiple resolutions, UMAP"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

Part two of the series. We start from the **preprocessed object saved in Tutorial 01** and walk through:

-   Linear dimensionality reduction with PCA
-   Choosing how many PCs to keep
-   Building the shared nearest-neighbor graph
-   Clustering at multiple resolutions and comparing them
-   UMAP embedding

## Learning goals

-   Run `RunPCA()` and read PCA diagnostic plots
-   Use `ElbowPlot()` to pick a defensible number of PCs
-   Run `FindNeighbors()` → `FindClusters()` at several resolutions and inspect the difference
-   Switch the active identity between clusterings with `Idents()`
-   Generate a UMAP and overlay cluster labels

## Setup

```{r}
#| label: setup
#| eval: true

library(Seurat)
library(tidyverse)
set.seed(2026)
```

```{r}
#| label: resume

nsclc.seurat.obj <- readRDS("../data/nsclc_preprocessed.rds")
nsclc.seurat.obj
```

------------------------------------------------------------------------

## Step 1 — Linear dimensionality reduction (PCA)

```{r}
#| label: pca

nsclc.seurat.obj <- RunPCA(
  nsclc.seurat.obj,
  features = VariableFeatures(object = nsclc.seurat.obj)
)

print(nsclc.seurat.obj[["pca"]], dims = 1:5, nfeatures = 5)
DimHeatmap(nsclc.seurat.obj, dims = 1, cells = 500, balanced = TRUE)
```

::: {.callout-important title="Think about it"}
1.  PCA here runs on HVGs, not all genes. What would change if we included all genes?
2.  `DimHeatmap(dims = 1, cells = 500)` — why show only 500 cells per PC?
3.  What does a PC whose top loadings are all mitochondrial genes usually mean?

<details>

<summary>Show answers</summary>

1.  PCs would be dominated by highly-expressed housekeeping genes whose variance simply reflects sequencing depth. HVG-restricted PCA isolates biologically-variable signal.
2.  The heatmap orders cells along the PC and shows gene expression for the extreme low and high cells. 500 is enough to see the gradient; more would just smear the display and slow rendering.
3.  It's a **technical PC** tracking cell quality / stress rather than biology. If it dominates, consider regressing `percent.mt` in `ScaleData`, or excluding it from downstream clustering dims.

</details>
:::

## Step 2 — Choose how many PCs to keep

```{r}
#| label: elbow

ElbowPlot(nsclc.seurat.obj, ndims = 40)
```

::: {.callout-important title="Think about it"}
1.  Where is "the elbow" on an ElbowPlot, and why does it suggest a cutoff?
2.  What's a principled alternative to eyeballing the elbow?
3.  Is it risky to choose too *few* PCs? Too *many*?

<details>

<summary>Show answers</summary>

1.  The elbow is where the variance explained per PC flattens out — additional PCs stop adding much signal. Picking the elbow captures most of the biological structure without the noise-heavy tail.
2.  `JackStraw()` + `JackStrawPlot()` (permutation-based significance) is more principled but slow on large datasets. `Stuart et al. 2019` also showed that results are fairly robust across reasonable PC counts, so the heuristic is fine in practice.
3.  **Too few** loses cell-type resolution (rare populations blur together). **Too many** pulls in noisy dimensions and inflates distances in the neighbor graph. Being slightly generous (e.g. 15–30) is usually safe.

</details>
:::

```{r}
#| label: n-pcs

n_pcs <- 15
```

## Step 3 — Build the neighbor graph

```{r}
#| label: neighbors

nsclc.seurat.obj <- FindNeighbors(nsclc.seurat.obj, dims = 1:n_pcs)
```

::: {.callout-important title="Think about it"}
1.  What graph does `FindNeighbors()` build?
2.  Why operate on PC space rather than raw expression?

<details>

<summary>Show answers</summary>

1.  It builds a **k-nearest-neighbor** graph (default `k = 20`) and then a **shared-nearest-neighbor** graph: two cells are connected if they share many nearest neighbors. The Jaccard-weighted SNN graph is what clustering runs on.
2.  PC space is denoised, low-dimensional, and has well-behaved Euclidean distance. Raw expression is sparse, high-dimensional, and distance-distorted by the *curse of dimensionality*.

</details>
:::

## Step 4 — Cluster at several resolutions

```{r}
#| label: cluster

nsclc.seurat.obj <- FindClusters(
  nsclc.seurat.obj,
  resolution = c(0.1, 0.3, 0.5, 0.7, 1.0)
)

head(nsclc.seurat.obj@meta.data)
```

::: {.callout-important title="Think about it"}
1.  What does `resolution` control intuitively?
2.  Each resolution creates its own `RNA_snn_res.XYZ` column in the metadata. How do you make one of them the *active* identity?
3.  Try `clustree::clustree(nsclc.seurat.obj, prefix = "RNA_snn_res.")`. What does it show?

<details>

<summary>Show answers</summary>

1.  The granularity knob for the graph-based community detection. Higher resolution ⇒ smaller, more numerous clusters.
2.  `Idents(nsclc.seurat.obj) <- "RNA_snn_res.0.3"` — or any of the resolution columns. `DimPlot(..., group.by = "RNA_snn_res.0.3")` also works without changing active idents.
3.  A **clustering tree**: each row is a resolution, each node is a cluster, and edges show how cells flow from one resolution to the next. Stable clusters stay as one node across levels; unstable ones fragment. It's a great way to pick a resolution visually.

</details>
:::

## Step 5 — Visualize cluster counts across resolutions

```{r}
#| label: cluster-counts

# quick summary of cluster counts per resolution
nsclc.seurat.obj@meta.data |>
  select(starts_with("RNA_snn_res.")) |>
  map_int(~ length(unique(.x)))
```

::: {.callout-important title="Think about it"}
1.  At `res = 0.1` you might see 6 clusters; at `res = 1.0` you might see 20+. Which is "right"?
2.  Does changing resolution change the *underlying biology* or just how the same biology is partitioned?

<details>

<summary>Show answers</summary>

1.  Neither is objectively right. Resolution is a **knob** — choose the one that matches the biological question. For a broad cell-type atlas, lower resolution is fine; to split subtypes, you need higher. Confirm with markers.
2.  Just the partitioning. The underlying graph and cell embeddings are unchanged — resolution decides where to draw the cluster boundaries.

</details>
:::

## Step 6 — Pick a resolution, set identities, and plot

```{r}
#| label: pick-res

Idents(nsclc.seurat.obj) <- "RNA_snn_res.0.5"
DimPlot(
  nsclc.seurat.obj,
  group.by = "RNA_snn_res.0.5",
  label    = TRUE
)
```

## Step 7 — Non-linear embedding (UMAP)

```{r}
#| label: umap

nsclc.seurat.obj <- RunUMAP(nsclc.seurat.obj, dims = 1:n_pcs)
DimPlot(nsclc.seurat.obj, reduction = "umap", label = TRUE)
```

::: {.callout-important title="Think about it"}
1.  Clusters are defined on the graph, *not* on UMAP. How can clustering and UMAP disagree visually?
2.  Why shouldn't you compare UMAP distances quantitatively between clusters?

<details>

<summary>Show answers</summary>

1.  UMAP is a separate non-linear projection. A cluster may occasionally split visually on UMAP while still being one graph community — and vice versa. Trust the graph-based cluster call; use UMAP as a sanity check.
2.  UMAP preserves *local* neighborhoods but distorts *global* distances to fit in 2D. A cluster that looks far away on UMAP isn't necessarily biologically more distant than one nearby.

</details>
:::

## Save and continue to Tutorial 03

```{r}
#| label: save

saveRDS(nsclc.seurat.obj, file = "../data/nsclc_clustered.rds")
```

::: callout-tip
Continue with **Tutorial 03 — Markers & Cell Type Annotation**.
:::

## Credits

-   Workflow source: Khushbu Patel, single-cell RNA-seq standard workflow
-   References: [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html) · [clustree](https://cran.r-project.org/package=clustree)
