---
title: "Talapas 07 — Pseudobulk DE with DESeq2"
subtitle: "Aggregate to (donor × condition × cell type), then DESeq2 per cell type — on 20k NSCLC"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

::: callout-important
**This is the Talapas Advanced path.** Laptop counterpart: **[Tutorial 06 — DESeq2: Bulk Primer + Pseudobulk DE](../Tutorial_06_DESeq2_DE.html)**, which is the place to read the DESeq2 mechanics in detail. This page just runs the same pseudobulk workflow on the 20k NSCLC object as a SLURM job.
:::

## About this tutorial

Starting from `nsclc_integrated.rds` ([Talapas 06](Tutorial_Talapas_06_Integration.html)), aggregate single-cell counts to one observation per `(donor, condition, cell_type)` and run DESeq2 per cell type.

The NSCLC dataset has 7 donors but **no built-in condition variable** (all 7 are NSCLC primary tumors). The script demonstrates the workflow against a **synthetic two-group split** (donors d1–d3 vs d4–d7). For real biology, replace `condition` with your own metadata (treatment, timepoint, response group, etc.).

The executable script is **[`07_pseudobulk_de.R`](scripts/07_pseudobulk_de.R)**. Submit with:

```bash
sbatch --job-name=pb_de --time=04:00:00 --mem=96G run_rscript.sbatch 07_pseudobulk_de.R
```

## What the script does

1. Reads `nsclc_integrated.rds`, assigns a synthetic `condition` (or uses your own).
2. `AggregateExpression(group.by = c("donor", "condition", "celltype"))` → pseudobulk matrix.
3. Aligns pseudobulk colnames with a metadata table parsed from the Seurat object.
4. Drops `(donor × condition × celltype)` groups with fewer than 10 cells.
5. Loops over cell types: builds `DESeqDataSet`, runs `DESeq()`, shrinks LFCs with `apeglm`.
6. Concatenates results, drops `NA`-padj rows, writes `nsclc_pseudobulk_de.csv`.

## Required resources

| Resource | Recommendation |
|---|---|
| Partition | `compute` |
| Time | 2–4 h |
| CPUs | 8 |
| Memory | 96 GB (DESeq2 + dense pseudobulk matrices) |

## Code (mirror of the script)

```{r}
#| label: setup
library(Seurat)
library(DESeq2)
library(tidyverse)
set.seed(2026)

seu <- readRDS("../objects/nsclc_integrated.rds")
DefaultAssay(seu) <- "RNA"
```

```{r}
#| label: condition
# Replace with your real metadata for real DE
if (!"condition" %in% colnames(seu@meta.data))
  seu$condition <- ifelse(seu$donor %in% c("d1","d2","d3"), "GroupA", "GroupB")
seu$celltype <- seu$celltype_manual
```

```{r}
#| label: aggregate
pb <- AggregateExpression(
  seu, assays = "RNA", slot = "counts",
  group.by = c("donor", "condition", "celltype"),
  return.seurat = FALSE)$RNA
```

```{r}
#| label: de-loop
# See the script for the full per-cell-type loop and filter logic.
# Output: nsclc_pseudobulk_de.csv with columns gene, baseMean, log2FoldChange,
# lfcSE, pvalue, padj, celltype.
```

::: {.callout-important title="Think about it"}
Why pseudobulk and not `FindMarkers(group.by = "condition")` on the cells directly?

<details>
<summary>Show answer</summary>

Cells from the same donor are **not statistically independent**. Per-cell tests inflate the effective sample size, producing **anti-conservative** p-values. Pseudobulk respects the donor as the biological unit. See Squair *et al.* 2021 (Nat Commun) for the full demonstration on the `ifnb` dataset — same logic applies here.

</details>
:::

## Next

-   **[Talapas 08 — Functional Analysis](Tutorial_Talapas_08_FunctionalAnalysis.html)** — feed `nsclc_pseudobulk_de.csv` into clusterProfiler
-   **[Talapas 09 — Differential Abundance](Tutorial_Talapas_09_DifferentialAbundance.html)** — the *complementary* question: do cell-type proportions change?
