---
title: "Talapas 06 — Multi-Sample Integration with Harmony"
subtitle: "Correct donor-level batch effects on the 20k NSCLC integrated PCA embedding"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

::: callout-important
**This is the Talapas Advanced path, not a laptop tutorial.** Submit each step as a SLURM job — see [Talapas 01](Tutorial_Talapas_01_SLURM_Basics.html). The laptop counterpart is **[Tutorial 05 — Two-Sample Integration](../Tutorial_05_Integration.html)**.
:::

## About this tutorial

Pick up the annotated NSCLC object (`nsclc_annotated.rds`, written by [Talapas 05](Tutorial_Talapas_05_Markers_Annotation.html)) and integrate across donors using **harmony**. The 10x 20k NSCLC dataset pools dissociated tumor cells from **seven donors** via CMO multiplexing, so each donor is a natural batch — donor-of-origin captures most of the technical and inter-individual variation we want to remove before condition-level analyses.

The executable script is **[`06_integration.R`](scripts/06_integration.R)**. Submit with:

```bash
sbatch --job-name=integrate run_rscript.sbatch 06_integration.R
```

## What the script does

1. Loads `nsclc_annotated.rds` and sets `RNA` as the default assay.
2. If the object lacks a `donor` column, synthesizes seven pseudo-donors so the workflow demonstrates correctly (replace with real CMO assignments for a real analysis — see the 10x **CellPlex** demultiplexing guidance in the dataset's web summary).
3. Re-runs PCA on the joint matrix.
4. Calls **`harmony::RunHarmony(group.by.vars = "donor")`** to produce a donor-corrected embedding stored in the `harmony` slot.
5. Re-runs `FindNeighbors` / `FindClusters` / `RunUMAP` on the harmony embedding.
6. Writes `nsclc_integrated.rds` for downstream pseudobulk DE / DA.

## Required resources

| Resource | Recommendation |
|---|---|
| Partition | `compute` |
| Time | 1–2 h for 20k cells |
| CPUs | 8 |
| Memory | 64 GB |

## Code (mirror of the script)

```{r}
#| label: setup
library(Seurat)
library(harmony)
library(tidyverse)
set.seed(2026)

seu <- readRDS("../objects/nsclc_annotated.rds")
```

```{r}
#| label: harmony
DefaultAssay(seu) <- "RNA"
seu <- RunPCA(seu, npcs = 30, verbose = FALSE)
seu <- RunHarmony(seu, group.by.vars = "donor",
                  reduction.use = "pca", reduction.save = "harmony")
```

```{r}
#| label: cluster-umap
seu <- FindNeighbors(seu, reduction = "harmony", dims = 1:30)
seu <- FindClusters(seu, resolution = 0.5)
seu <- RunUMAP(seu, reduction = "harmony", dims = 1:30,
               reduction.name = "umap_harmony")
saveRDS(seu, "../objects/nsclc_integrated.rds")
```

::: {.callout-important title="Think about it"}
After harmony, you typically see cells from different donors co-mingle in each cluster on `umap_harmony`. What would tell you that integration has **over-corrected** (i.e. removed real biology, not just batch)?

<details>
<summary>Show answer</summary>

A cell-type marker that should be cell-type-specific (`CD3D` for T cells, `EPCAM` for epithelial) becoming **scattered** across multiple clusters after integration, or two cell types that should be distinct (T vs B) being merged. Always re-check `DotPlot()` of your canonical markers on the harmony clusters before continuing.

</details>
:::

## Next

-   **[Talapas 07 — Pseudobulk DE](Tutorial_Talapas_07_Pseudobulk_DE.html)** — DESeq2 on (donor × condition × cell type) counts
-   **[Talapas 09 — Differential Abundance with miloR](Tutorial_Talapas_09_DifferentialAbundance.html)** — uses the harmony embedding directly
