---
title: "Talapas 09 — Differential Abundance with miloR"
subtitle: "Neighbourhood-level abundance testing on the 20k NSCLC harmony graph"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

::: callout-important
**This is the Talapas Advanced path.** Laptop counterpart: **[Tutorial 08 — Differential Abundance](../Tutorial_08_DifferentialAbundance.html)**, which discusses the miloR workflow in detail (refining neighborhoods, sample design, multiple-testing on the graph). This page runs the same workflow on the 20k NSCLC integrated object.
:::

## About this tutorial

Pseudobulk DE ([Talapas 07](Tutorial_Talapas_07_Pseudobulk_DE.html)) asks: **"for cells of the same type, which genes change?"** Differential abundance asks the complementary question: **"do the cell-type proportions themselves change?"**

`miloR` partitions the integrated kNN graph into overlapping neighbourhoods and tests each one for a condition-dependent change in cell count — a finer resolution than aggregating by cluster label.

The executable script is **[`09_differential_abundance.R`](scripts/09_differential_abundance.R)**. Submit with:

```bash
sbatch --job-name=da --time=04:00:00 --mem=96G run_rscript.sbatch 09_differential_abundance.R
```

## What the script does

1. Loads `nsclc_integrated.rds` (from [Talapas 06](Tutorial_Talapas_06_Integration.html)).
2. Synthesizes a binary `condition` if none present (same split as Talapas 07).
3. Converts to `SingleCellExperiment` → `Milo`.
4. `buildGraph(k = 30, d = 30, reduced.dim = "HARMONY")` — uses the harmony embedding.
5. `makeNhoods()`, `countCells()` with `sample = "donor"`.
6. `calcNhoodDistance()`, `testNhoods(design = ~ condition)`.
7. Writes `nsclc_milo_da.csv` with per-neighbourhood logFC, SpatialFDR, and the neighbourhood index.

## Required resources

| Resource | Recommendation |
|---|---|
| Partition | `compute` |
| Time | 2–4 h |
| CPUs | 8 |
| Memory | 96 GB (Milo graph + distance matrices) |

## Code (mirror of the script)

```{r}
#| label: setup
library(Seurat)
library(miloR)
library(SingleCellExperiment)
library(scater)
library(tidyverse)
set.seed(2026)

seu <- readRDS("../objects/nsclc_integrated.rds")
sce <- as.SingleCellExperiment(seu)
milo <- Milo(sce)
```

```{r}
#| label: milo
milo <- buildGraph(milo, k = 30, d = 30, reduced.dim = "HARMONY")
milo <- makeNhoods(milo, prop = 0.1, k = 30, d = 30,
                   refined = TRUE, reduced_dims = "HARMONY")
milo <- countCells(milo, meta.data = as.data.frame(colData(milo)),
                   sample = "donor")
milo <- calcNhoodDistance(milo, d = 30, reduced.dim = "HARMONY")

design_df <- as.data.frame(colData(milo)) |>
  distinct(donor, condition) |>
  column_to_rownames("donor")

res <- testNhoods(milo, design = ~ condition, design.df = design_df)
write_csv(as_tibble(res), "../objects/nsclc_milo_da.csv")
```

::: {.callout-important title="Think about it"}
Pseudobulk DE (Talapas 07) and differential abundance (this tutorial) sometimes give **opposite** answers for the same cell type. How can both be right?

<details>
<summary>Show answer</summary>

They answer different questions:

-   Pseudobulk DE: "**within** cells annotated as cell type X, which genes shift between conditions?"
-   Differential abundance: "do **proportions** of cell type X change between conditions?"

A cell type can be unchanged in proportion but every cell in it shifts its expression (DE-positive, DA-negative), **or** it can shift dramatically in proportion without changing per-cell expression (DA-positive, DE-negative). They're complementary, not redundant — and reporting both is the modern best practice.

</details>
:::

## What you've completed

You've now run the **full Talapas Advanced path**: raw FASTQs → Cell Ranger counts → QC → dim-reduction → markers → integration → pseudobulk DE → functional enrichment → differential abundance. Each step is a SLURM job; each output flows into the next. This is the production scRNA-seq pipeline at the scale of real datasets.

For the *bulk* RNA-seq + DESeq2 fundamentals that underpin Talapas 07, revisit the **laptop** [Tutorial 06 — DESeq2: Bulk Primer + Pseudobulk DE](../Tutorial_06_DESeq2_DE.html).
