---
title: "Talapas 03 — QC & Preprocessing on 20k NSCLC DTC"
subtitle: "Load a 10x dataset, filter cells, normalize, pick HVGs, scale"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

::: callout-important
**This is the Talapas scaling exercise, not the laptop tutorial.**
The main workshop sequence (Tutorials 01 → 11 in [Exercises](../../Tutorials.html)) runs on the small **`ifnb`** dataset (\~25 MB) and is designed for a laptop. **Do those first.** This NSCLC scaling exercise re-runs the QC / dim-reduction / annotation steps on a 10× larger dataset (20k cells, 7 donors, \~250 MB compressed) so you can see how the same workflow scales — typically run as R + SLURM scripts on Talapas. See the [SLURM Basics tutorial](Tutorial_Talapas_01_SLURM_Basics.html).
:::

## About this tutorial

The first of a three-part **NSCLC scaling exercise** walking the standard **Seurat** scRNA-seq workflow on the public **10x Genomics 20k NSCLC DTC** dataset (20k cells, 7 donors, 3' v3.1 Multiplex). This part stops after **ScaleData** — the object you save here is the input to Talapas 04.

-   Upstream script: *Khushbu Patel, single-cell RNA-seq standard workflow*
-   Reference vignette: [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)
-   Laptop counterpart: [Tutorial 01 — QC & Preprocessing on `ifnb`](../Tutorial_01_QC_Preprocessing.html)

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it (chunks default to `eval: false`). To actually run the analysis:

1.  Download the `.qmd` source: <a href="Tutorial_Talapas_03_QC_Preprocessing.qmd.txt" download="Tutorial_Talapas_03_QC_Preprocessing.qmd">Tutorial_Talapas_03_QC_Preprocessing.qmd</a>. If your browser saves the file as `Tutorial_Talapas_03_QC_Preprocessing.qmd.txt`, **drop the trailing `.txt`** so the filename ends in `.qmd`, then open it in RStudio.
2.  Download the dataset using the browser instructions [below](#downloading-the-data) and place the files in a `data/` folder at the repository root.
3.  Work through the chunks, flipping `eval: true` as you go.
:::

## Dataset — 20k NSCLC DTC (10x Genomics)

**What it is.** Twenty-thousand dissociated tumor cells (DTCs) pooled from **seven non-small-cell lung cancer (NSCLC) donors**, run on the 10x Chromium **3' v3.1** assay with **CellPlex multiplexing** (CMO hashing) so that every cell carries both a gene-expression profile and a sample-of-origin barcode.

-   Landing page: [10x Genomics — 20k NSCLC DTC, 3' v3.1 Multiplex](https://www.10xgenomics.com/datasets/20-k-mixture-of-nsclc-dt-cs-from-7-donors-3-v-3-1-multiplex-3-1-standard-6-0-0)
-   Reference genome: GRCh38 (human)
-   Chemistry: 10x Chromium Single Cell 3' v3.1 (dual index, Next GEM), CellPlex multiplexing
-   Cell count (target recovery): \~20,000 post-filter cells across 7 donors

This is the **same dataset Patel uses** in [`singleCell_standard_workflow.R`](https://github.com/kpatel427/YouTubeTutorials/blob/main/singleCell_standard_workflow.R) — but we use the smaller **filtered** matrix instead of the 1.5 GB raw matrix because Cell Ranger has already trimmed the empty-droplet barcodes for you.

### Downloading the data {#downloading-the-data}

10x Genomics gates dataset downloads behind a click-through terms-of-service prompt, so programmatic transfers with `wget` / `curl` return **HTTP 403** on a fresh machine. The reliable path — and the one we use in this workshop — is to download the data manually through your web browser:

1.  Open the [landing page](https://www.10xgenomics.com/datasets/20-k-mixture-of-nsclc-dt-cs-from-7-donors-3-v-3-1-multiplex-3-1-standard-6-0-0) in your browser.
2.  Scroll to the **Output and supplemental files** section.
3.  Find the row labelled *Feature / cell matrix (filtered)* and click **Download**. Accept the *Sign in or share data with us* / TOS pop-up if it appears.
4.  Save the file `20k_NSCLC_DTC_3p_nextgem_Multiplex_count_filtered_feature_bc_matrix.tar.gz` (≈250 MB) to your usual downloads folder.

Now make a `data/` directory at the **root of the tutorial repository** (i.e. one level up from `Exercise_Folder/`), move the downloaded archive into it, and unpack it. From a terminal:

```bash
# From the repository root (the folder that contains Exercise_Folder/)
mkdir -p data
mv ~/Downloads/20k_NSCLC_DTC_3p_nextgem_Multiplex_count_filtered_feature_bc_matrix.tar.gz data/
cd data
tar -xzf 20k_NSCLC_DTC_3p_nextgem_Multiplex_count_filtered_feature_bc_matrix.tar.gz
```

::: callout-tip
If you prefer not to use a terminal, double-clicking the `.tar.gz` on macOS, Windows, or most Linux desktops will extract it in place. The result is the same.
:::

After unpacking you should have a `filtered_feature_bc_matrix/` sub-directory inside `data/`, containing the **three sparse-matrix files** Cell Ranger emits:

```
data/
└── filtered_feature_bc_matrix/
    ├── barcodes.tsv.gz
    ├── features.tsv.gz
    └── matrix.mtx.gz
```

Those three files together encode the counts matrix that this tutorial uses; `Read10X()` in Seurat reads the directory and stitches them into a sparse matrix.

### The three files inside the bundle

| File | Format | Contents |
|---|---|---|
| `barcodes.tsv.gz` | gzipped TSV, one column | One row per cell barcode, e.g. `AAACCTGAGAAACCAT-1` — the 16-bp 10x barcode plus a `-N` suffix indicating the GEM well. These become the **column names** of the counts matrix. |
| `features.tsv.gz` | gzipped TSV, three columns | One row per measured feature: `Ensembl ID`, `gene symbol`, and feature **type** (`Gene Expression` or `Multiplexing Capture`). The `gene symbol` column becomes the **row names** of the counts matrix. |
| `matrix.mtx.gz` | gzipped Matrix Market | Sparse integer **UMI counts** in coordinate (`i  j  value`) format — one entry per non-zero (feature, cell) pair. The header records total features × cells × non-zero entries. |

::: callout-note
Cell Ranger has already dropped empty-droplet barcodes from this **filtered** matrix, which is why we use it instead of the larger raw matrix. The `min.cells = 3` and `min.features = 200` arguments we apply in Step 2 are *additional* per-feature and per-cell sanity filters layered on top of Cell Ranger's filter.
:::

### Data dictionary — `Read10X()` return value

`Read10X()` on this multi-feature directory returns a **named list** of sparse matrices (one per feature type). Each sparse matrix has:

| Slot | Type | Description |
|---|---|---|
| `dimnames[[1]]` (rows) | character | Gene **symbols** (e.g. `MT-ND1`, `CD3D`, `KRT8`), taken from column 2 of `features.tsv.gz`. |
| `dimnames[[2]]` (cols) | character | 16-bp 10x **cell barcodes** (e.g. `AAACCTGAGAAACCAT-1`), taken from `barcodes.tsv.gz`. The `-1` suffix is the GEM-well index. |
| values | integer | **UMI counts** from `matrix.mtx.gz` — number of unique transcript molecules observed for that (gene, cell) pair. |

| Named slot in the returned list | Matrix shape | Meaning |
|---|---|---|
| `` $`Gene Expression` `` | \~36,600 genes × \~42,000 barcodes | The gene-expression counts. **This is what the tutorial uses.** |
| `` $`Multiplexing Capture` `` | \~12 CMOs × \~42,000 barcodes | Per-barcode CellPlex hashtag counts, used to demultiplex donors. Not used here. |

## Learning goals

By the end of this tutorial you will be able to:

-   Read a 10x sparse-matrix directory into Seurat
-   Add per-cell QC metrics and visualize them
-   Choose filtering thresholds that are defensible for *your* dataset
-   Run `NormalizeData` → `FindVariableFeatures` → `ScaleData` and understand what each does
-   Track how each preprocessing step changes the dimensions of the counts matrix

## Setup

```{r}
#| label: setup

library(Seurat)
library(tidyverse)

set.seed(2026)
```

------------------------------------------------------------------------

## Step 1 — Read the 10x filtered matrix

```{r}
#| label: load

# After unpacking the .tar.gz, ../data/filtered_feature_bc_matrix/ contains
#   barcodes.tsv.gz, features.tsv.gz, matrix.mtx.gz
nsclc.sparse.m <- Read10X(data.dir = "../data/filtered_feature_bc_matrix/")

# 10x multi-feature outputs return a NAMED LIST — keep the Gene Expression slice
str(nsclc.sparse.m)
cts <- nsclc.sparse.m$`Gene Expression`
dim(cts)

# Record the starting matrix dimensions for the summary table later
dim_raw <- dim(cts)
```

::: {.callout-important title="Think about it"}
1.  Why does `Read10X` sometimes return a **list** instead of a matrix?
2.  What does each dimension of `cts` represent?

<details>

<summary>Show answers</summary>

1.  10x Multiplex / CellPlex / CITE-seq / multiome runs produce multiple measurement types in one Cell Ranger output (e.g. `Gene Expression`, `Antibody Capture`, `Peaks`, `Multiplexing Capture`). The reader returns a named list so you can pull the slice you want.
2.  Rows = genes (features), columns = cell barcodes. `dim(cts)[1]` is the gene count, `dim(cts)[2]` is the cell count.

</details>
:::

## Step 2 — Create the Seurat object

```{r}
#| label: create-object

nsclc.seurat.obj <- CreateSeuratObject(
  counts       = cts,
  project      = "NSCLC",
  min.cells    = 3,
  min.features = 200
)
nsclc.seurat.obj

# Record dimensions after the CreateSeuratObject filters
dim_after_create <- dim(nsclc.seurat.obj)
```

::: {.callout-important title="Think about it"}
1.  What exactly do `min.cells = 3` and `min.features = 200` remove?
2.  The `nsclc.sparse.m` list still sits in memory. What would `rm(nsclc.sparse.m); gc()` do, and why might you want to?

<details>

<summary>Show answers</summary>

1.  `min.cells = 3` drops genes detected in fewer than 3 cells (likely noise). `min.features = 200` drops barcodes with fewer than 200 detected genes — mostly empty droplets and broken cells. These are *permissive* first-pass filters.
2.  It releases the raw list and its memory. The full 10x raw feature matrix is large (often several GB); once counts are inside the Seurat object, the raw list is redundant.

</details>
:::

## Step 3 — Add QC metrics

```{r}
#| label: qc-metrics

nsclc.seurat.obj[["percent.mt"]] <- PercentageFeatureSet(nsclc.seurat.obj, pattern = "^MT-")
head(nsclc.seurat.obj@meta.data)
```

::: {.callout-important title="Think about it"}
1.  What biological signal does `percent.mt` capture?
2.  Why `"^MT-"` specifically? What would you use for mouse?

<details>

<summary>Show answers</summary>

1.  Dying or stressed cells leak cytoplasmic RNA while retaining mitochondrial RNA, so their mitochondrial fraction rises. High `percent.mt` flags apoptotic / compromised cells.
2.  `"^MT-"` matches human mitochondrial gene symbols (MT-ND1, MT-CO1, …). For mouse the convention is lowercase: `"^mt-"`.

</details>
:::

## Step 4 — Visualize QC

```{r}
#| label: qc-plots

VlnPlot(nsclc.seurat.obj,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3)

FeatureScatter(nsclc.seurat.obj,
               feature1 = "nCount_RNA",
               feature2 = "nFeature_RNA") +
  geom_smooth(method = "lm")
```

::: {.callout-important title="Think about it"}
1.  What does a **fat tail** on the `nFeature_RNA` violin suggest?
2.  On the `nCount_RNA` vs `nFeature_RNA` scatter, what would a doublet look like? A dying cell?

<details>

<summary>Show answers</summary>

1.  A long upper tail often means **doublets** — barcodes where two cells got trapped in one droplet, inflating both UMI count and unique gene count.
2.  Doublets sit **high on both axes**, often above the main trend line. Dying cells are **low** on both and tend to have **high `percent.mt`** (check by coloring or faceting).

</details>
:::

## Step 5 — Doublet detection with `scDblFinder`

`scDblFinder` simulates artificial doublets, trains a classifier, and returns a per-cell score and class. NSCLC's CellPlex multiplexing means we have CMO-derived sample labels — but the doublets we care about here are **droplet-level** doublets (two cells in one bead), which form *within* each 10x channel. For the simplest case (all cells loaded in one library) we run scDblFinder on the entire object; if you have CMO-demultiplexed sample assignments and want per-sample doublet rates, split first.

```{r}
#| label: doublet
#| eval: false

# install if needed:
# BiocManager::install("scDblFinder")
library(scDblFinder)

set.seed(2026)

sce <- as.SingleCellExperiment(nsclc.seurat.obj)
sce <- scDblFinder(sce, returnType = "sce")

nsclc.seurat.obj$scDblFinder.score <- colData(sce)$scDblFinder.score
nsclc.seurat.obj$scDblFinder.class <- colData(sce)$scDblFinder.class

table(nsclc.seurat.obj$scDblFinder.class)

VlnPlot(nsclc.seurat.obj,
        features = c("nFeature_RNA", "nCount_RNA"),
        group.by = "scDblFinder.class", pt.size = 0)
```

::: {.callout-important title="Think about it"}
A 20k-cell NSCLC run will typically flag \~5–8% of cells as doublets. Why might the doublet rate be higher in tumor dissociations than in PBMCs?

<details>

<summary>Show answers</summary>

Tumor dissociations produce **larger, stickier cells** (epithelial / mesenchymal) that are more likely to clump than free-floating immune cells. Some "doublets" are actually un-dissociated cell pairs that survive the microfluidic pinch. CellPlex / hash-tag demultiplexing helps catch *cross-sample* doublets (one barcode mapping to two CMOs) but does **not** catch within-sample doublets — that's what scDblFinder is for.

</details>
:::

## Step 6 — Filter low-quality cells

```{r}
#| label: filter

# Optionally include the scDblFinder.class filter (uncomment after running Step 4b):
nsclc.seurat.obj <- subset(
  nsclc.seurat.obj,
  subset = nFeature_RNA > 200 &
           nFeature_RNA < 2500 &
           percent.mt   < 5
           # & scDblFinder.class == "singlet"
)
nsclc.seurat.obj

# Record dimensions after the QC subset() filter
dim_after_filter <- dim(nsclc.seurat.obj)
```

::: {.callout-important title="Think about it"}
1.  The default cutoff here is `percent.mt < 5`. Is that appropriate for tumor tissue?
2.  How would you set thresholds *empirically* rather than by default?

<details>

<summary>Show answers</summary>

1.  Not always — tumor and certain primary tissues routinely have higher baseline `percent.mt` (\~10–20%). A 5% cutoff can strip out viable tumor cells. **Inspect the distribution first**; set the knee.
2.  Look at the violin/histogram shape; choose thresholds that trim the clear **tails** (not the mode). Some workshops use median ± *k* × MAD, or the `scater::isOutlier()` function for an automated rule.

</details>
:::

## Step 7 — Normalize

```{r}
#| label: normalize

nsclc.seurat.obj <- NormalizeData(nsclc.seurat.obj)
# Equivalent explicit form:
# NormalizeData(nsclc.seurat.obj,
#               normalization.method = "LogNormalize",
#               scale.factor = 10000)

# Record dimensions after normalization (these should be identical to dim_after_filter)
dim_after_normalize <- dim(nsclc.seurat.obj)
```

::: {.callout-important title="Think about it"}
1.  What does `LogNormalize` do algebraically?
2.  When would you prefer `SCTransform()` instead?

<details>

<summary>Show answers</summary>

1.  For each cell: divide each gene's count by the cell's total count, multiply by `scale.factor` (default 10 000), then take `log1p`. In formula form: `log(1 + (count_gc / total_c) * 10000)`.
2.  `SCTransform` models the mean–variance relationship per gene with a regularized negative binomial, which is usually better for **low-count** datasets and more faithful downstream variance. It also replaces the NormalizeData → FindVariableFeatures → ScaleData chain in a single call.

</details>
:::

## Counts matrix progression — summary

So far we've taken a freshly loaded 10x counts matrix and pushed it through two filtering steps. `NormalizeData()` rescales values but **does not** add or drop rows or columns, so the matrix shape is unchanged after Step 6. The table below makes the progression explicit so you can see exactly how many features (genes) and barcodes (cells) each step removed.

```{r}
#| label: dim-summary

dim_summary <- tibble::tribble(
  ~step,                                                        ~features_before,    ~features_after,        ~barcodes_before,    ~barcodes_after,
  "Step 1 — Read10X (raw load)",                                 NA_integer_,         dim_raw[1],             NA_integer_,         dim_raw[2],
  "Step 2 — CreateSeuratObject (min.cells=3, min.features=200)", dim_raw[1],          dim_after_create[1],    dim_raw[2],          dim_after_create[2],
  "Step 6 — subset() QC filter (nFeature_RNA, percent.mt)",      dim_after_create[1], dim_after_filter[1],    dim_after_create[2], dim_after_filter[2],
  "Step 7 — NormalizeData (no dimension change)",                dim_after_filter[1], dim_after_normalize[1], dim_after_filter[2], dim_after_normalize[2]
)
dim_summary
```

Your printed numbers will be similar to (not necessarily identical to) the values below — small differences in package versions or `Read10X` behaviour can shift the exact counts:

| Step | Features before | Features after | Barcodes before | Barcodes after |
|---|---:|---:|---:|---:|
| Step 1 — Read10X (raw load)                                    | —      | ~36,601 | —      | ~42,386 |
| Step 2 — CreateSeuratObject (`min.cells=3`, `min.features=200`) | ~36,601 | ~24,000 | ~42,386 | ~22,000 |
| Step 6 — `subset()` QC filter (`nFeature_RNA`, `percent.mt`)    | ~24,000 | ~24,000 | ~22,000 | ~17,000 |
| Step 7 — `NormalizeData` (no dimension change)                  | ~24,000 | ~24,000 | ~17,000 | ~17,000 |

Steps 7 (`FindVariableFeatures`) and 8 (`ScaleData`) likewise leave the matrix shape intact — they only attach extra metadata or transform values in place — so this table captures the entire dimensional story of the tutorial.

::: {.callout-important title="Think about it"}
1.  Which step removes the largest number of features? Which removes the largest number of barcodes? Are those expected?
2.  If you relaxed `percent.mt < 5` to `percent.mt < 10`, which row of this table would change, and in which direction?

<details>

<summary>Show answers</summary>

1.  Most **features** are removed by `CreateSeuratObject`'s `min.cells = 3` rule, because many genes are detected in essentially zero cells of this filtered matrix (and the `Multiplexing Capture` rows do not appear here at all — we kept only the `Gene Expression` slice). Most **barcodes** are removed by `subset()` — that's where we apply biological QC thresholds on top of Cell Ranger's empty-droplet filter.
2.  Only the Step 5 row would change. `barcodes_after` would go up because a more permissive `percent.mt` threshold retains more cells; the feature count is unaffected.

</details>
:::

## Step 8 — Highly variable features

```{r}
#| label: hvgs

nsclc.seurat.obj <- FindVariableFeatures(
  nsclc.seurat.obj,
  selection.method = "vst",
  nfeatures = 2000
)

top10 <- head(VariableFeatures(nsclc.seurat.obj), 10)
plot1 <- VariableFeaturePlot(nsclc.seurat.obj)
LabelPoints(plot = plot1, points = top10, repel = TRUE)
```

::: {.callout-important title="Think about it"}
1.  Why do we want to focus on *variable* genes instead of using all genes?
2.  What happens if you set `nfeatures = 500`? Or `nfeatures = 10000`?

<details>

<summary>Show answers</summary>

1.  Most genes are housekeeping / ubiquitously expressed and carry little cell-type information. Selecting HVGs raises signal-to-noise for PCA and clustering and speeds everything up.
2.  Too few (500) risks missing subtle cell-type signals and is brittle to noisy gene choices. Too many (10 000) pulls in low-information genes and dilutes the signal. 1 500–3 000 is a common sweet spot; 2 000 is a good default.

</details>
:::

## Step 9 — Scale

```{r}
#| label: scale

all.genes <- rownames(nsclc.seurat.obj)
nsclc.seurat.obj <- ScaleData(nsclc.seurat.obj, features = all.genes)
```

::: {.callout-important title="Think about it"}
1.  What does `ScaleData` do to each gene numerically?
2.  Why is scaling necessary *before* PCA?
3.  How would you regress out the cell cycle or `percent.mt` here?

<details>

<summary>Show answers</summary>

1.  Centers each gene's expression to mean 0 and scales to unit variance across cells (z-scoring).
2.  PCA finds directions of maximum variance. Without scaling, highly-expressed genes (which inherently have large raw variance) would dominate PCs, regardless of whether they are biologically informative.
3.  Pass `vars.to.regress = c("percent.mt", "S.Score", "G2M.Score")` (after running `CellCycleScoring()` for the latter two). ScaleData regresses those out of each gene's expression before z-scoring.

</details>
:::

## Save and continue to Tutorial 02

```{r}
#| label: save

saveRDS(nsclc.seurat.obj, file = "../data/nsclc_preprocessed.rds")
```

::: callout-tip
The saved `.rds` is the starting point for **Tutorial 02 — Dimensionality reduction & clustering**.
:::

## Credits

-   Dataset: [10x Genomics — 20k NSCLC DTC, 3' v3.1 Multiplex](https://www.10xgenomics.com/datasets/20-k-mixture-of-nsclc-dt-cs-from-7-donors-3-v-3-1-multiplex-3-1-standard-6-0-0)
-   Workflow source: Khushbu Patel, single-cell RNA-seq standard workflow
-   Reference: [Seurat PBMC 3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)
