---
title: "Tutorial 07 — WGCNA on COVID-19 PBMCs (GSE152418)"
subtitle: "Weighted gene co-expression networks, Patel-style"
author: "Single Cell RNA-seq Workshop"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

## About this tutorial

A worked **WGCNA** pipeline that mirrors [Patel's `WGCNA.R`](https://github.com/kpatel427/YouTubeTutorials/blob/main/WGCNA.R), with workshop-specific commentary and call-out boxes. The dataset is **GEO accession GSE152418** — bulk RNA-seq of PBMCs from COVID-19 patients across a severity gradient. We'll build a signed co-expression network and correlate the resulting modules with disease severity.

::: {.callout-tip title="How to use this page"}
The rendered HTML shows the code but does **not** execute it. To run it:

1.  [Download the `.qmd` source](Tutorial_07_WGCNA.qmd) and open it in RStudio.
2.  Download the GEO data (block below) into `../data/`.
3.  Work through the chunks, flipping `eval: true` as you go.
:::

::: callout-note
**Companion lecture:** [Lecture 07 — WGCNA](../Lecture_Folder/Lecture_07_WGCNA.html)
:::

## Dataset — GSE152418 (COVID-19 PBMC bulk RNA-seq)

**What it is.** PBMC bulk RNA-seq from a published cross-sectional study of COVID-19 patients (Arunachalam et al. 2020, *Science* 369: 1210–1220). Roughly 34 samples spanning healthy controls, convalescent, moderate, severe, and ICU patients.

-   GEO landing page: <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE152418>
-   Companion paper: Arunachalam et al. (2020) *Science* — <https://doi.org/10.1126/science.abc6261>
-   Files we'll use:
    -   `GSE152418_p20047_Study1_RawCounts.txt.gz` — gene × sample raw count table
    -   GEO **soft / matrix** file for sample metadata (loaded via `GEOquery`)

### Download with `wget` or `curl`

Direct GEO supplementary FTP links (NCBI mirror). Place into `../data/`.

With `wget`:

```bash
mkdir -p ../data && cd ../data

# Raw counts (supplementary file)
wget https://ftp.ncbi.nlm.nih.gov/geo/series/GSE152nnn/GSE152418/suppl/GSE152418_p20047_Study1_RawCounts.txt.gz
gunzip GSE152418_p20047_Study1_RawCounts.txt.gz
```

With `curl`:

```bash
mkdir -p ../data && cd ../data

curl -L -O https://ftp.ncbi.nlm.nih.gov/geo/series/GSE152nnn/GSE152418/suppl/GSE152418_p20047_Study1_RawCounts.txt.gz
gunzip GSE152418_p20047_Study1_RawCounts.txt.gz
```

From R via `GEOquery`:

```{r}
#| eval: false
library(GEOquery)
# This pulls the series matrix (metadata) and supplementary files into ../data/
getGEOSuppFiles("GSE152418", baseDir = "../data")
```

::: callout-tip
GEO link rot: if the supplementary file URL changes, get the current one from the *Supplementary file* table at the bottom of the GSE152418 GEO landing page.
:::

## Setup

```{r}
#| label: setup
# One-time install:
# install.packages("BiocManager")
# BiocManager::install(c("WGCNA", "DESeq2", "GEOquery", "CorLevelPlot",
#                        "gridExtra", "tidyverse"))

library(WGCNA)
library(DESeq2)
library(GEOquery)
library(tidyverse)
library(CorLevelPlot)
library(gridExtra)

allowWGCNAThreads()    # multi-threading

set.seed(2026)
```

## Step 1 — Load data and metadata

```{r}
#| label: load
data <- read.delim("../data/GSE152418_p20047_Study1_RawCounts.txt", header = TRUE)

# Pull GEO metadata
gse       <- getGEO("GSE152418", GSEMatrix = TRUE)
phenoData <- pData(phenoData(gse[[1]]))
phenoData <- phenoData[, c(1, 2, 46:50)]
head(phenoData)
```

```{r}
#| label: tidy-counts
# Reshape to a tidy long form, attach phenoData (matching `samples` to `title`),
# then spread back to a counts matrix indexed by Ensembl ID.
data <- data |>
  gather(key = "samples", value = "counts", -ENSEMBLID) |>
  mutate(samples = gsub("\\.", "-", samples)) |>
  inner_join(phenoData, by = c("samples" = "title")) |>
  select(1, 3, 4) |>
  spread(key = "geo_accession", value = "counts") |>
  column_to_rownames(var = "ENSEMBLID")

dim(data)
data[1:5, 1:5]
```

## Step 2 — Outlier detection

::: callout-note
**Why before normalization?** Outlying samples bias the variance-stabilizing transform and inflate the dispersion estimates downstream. Drop them first.
:::

```{r}
#| label: gsg
gsg <- goodSamplesGenes(t(data))
summary(gsg)
gsg$allOK

# Drop outlier genes
data <- data[gsg$goodGenes == TRUE, ]
```

```{r}
#| label: hclust
htree <- hclust(dist(t(data)), method = "average")
plot(htree)
```

```{r}
#| label: pca-outlier
pca       <- prcomp(t(data))
pca.dat   <- pca$x
pca.var   <- pca$sdev^2
pca.var.p <- round(pca.var / sum(pca.var) * 100, digits = 2)

ggplot(as.data.frame(pca.dat), aes(PC1, PC2)) +
  geom_point() +
  geom_text(label = rownames(pca.dat), nudge_y = 5) +
  labs(x = paste0("PC1: ", pca.var.p[1], "%"),
       y = paste0("PC2: ", pca.var.p[2], "%")) +
  theme_minimal()
```

::: {.callout-important title="Think about it"}
The hierarchical-clustering tree and the PCA both expose outliers. Which would you trust more on a small dataset?

<details>
<summary>Show answers</summary>

PCA — it gives you a continuous view of how far each sample is from the rest, in the same space the model will use. `hclust` only tells you the *order* of merging, which can hide a sample that's mildly aberrant but never gets clustered alone.
</details>
:::

```{r}
#| label: drop-outliers
# Patel's example drops a couple of obvious outliers; modify to fit your eyeballed picks.
samples.to.exclude <- c("GSM4614993", "GSM4614994", "GSM4614995")
data.subset <- data[, !(colnames(data) %in% samples.to.exclude)]

# Subset metadata accordingly
colData <- phenoData |>
  filter(!(geo_accession %in% samples.to.exclude)) |>
  column_to_rownames(var = "geo_accession")

# Make sure rows of colData match cols of counts
all(rownames(colData) %in% colnames(data.subset))
all(rownames(colData) == colnames(data.subset))
```

## Step 3 — Normalize with DESeq2 VST

WGCNA expects expression values that are **roughly homoscedastic** across the mean range. Variance-stabilizing transformation does that for count data:

```{r}
#| label: vst
dds <- DESeqDataSetFromMatrix(countData = data.subset,
                              colData   = colData,
                              design    = ~ 1)        # WGCNA is unsupervised

# Drop genes with < 15 reads in 3/4 of samples
dds75 <- dds[ rowSums(counts(dds) >= 15) >= 24, ]
nrow(dds75)

dds_norm <- vst(dds75)
norm.counts <- assay(dds_norm) |> t()      # WGCNA wants samples × genes
```

::: callout-tip
**Why `~ 1`?** WGCNA's network is unsupervised — the design should not encode the trait you're going to test against, otherwise you're "double dipping". Use `~ 1` for an unmodelled VST.
:::

## Step 4 — Pick the soft-thresholding power

```{r}
#| label: pick-power
power <- c(1:10, seq(12, 30, by = 2))
sft <- pickSoftThreshold(norm.counts,
                         powerVector = power,
                         networkType = "signed",
                         verbose     = 5)
sft.data <- sft$fitIndices

a1 <- ggplot(sft.data, aes(Power, SFT.R.sq, label = Power)) +
        geom_point() + geom_text(nudge_y = 0.05) +
        geom_hline(yintercept = 0.8, color = "red") +
        labs(x = "Power", y = "Scale-free topology fit, signed R²") +
        theme_classic()
a2 <- ggplot(sft.data, aes(Power, mean.k., label = Power)) +
        geom_point() + geom_text(nudge_y = 0.1) +
        labs(x = "Power", y = "Mean connectivity") +
        theme_classic()
grid.arrange(a1, a2, nrow = 2)
```

::: callout-warning
**Pick the smallest power that crosses R² ≈ 0.8** *and* gives a reasonable mean connectivity. Patel's example lands at **β = 18** for this dataset; yours will vary.
:::

## Step 5 — Build the network

```{r}
#| label: build-network
soft_power     <- 18
temp_cor       <- cor                    # save base::cor before WGCNA shadows it
cor            <- WGCNA::cor

bwnet <- blockwiseModules(norm.counts,
                          maxBlockSize     = 14000,
                          TOMType          = "signed",
                          power            = soft_power,
                          mergeCutHeight   = 0.25,
                          numericLabels    = FALSE,
                          randomSeed       = 1234,
                          verbose          = 3)

cor <- temp_cor                          # restore base::cor
```

::: callout-warning
**`cor` shadowing.** WGCNA replaces base R's `cor()`. If you forget to restore it (`cor <- temp_cor`), unrelated downstream code can break in confusing ways.
:::

```{r}
#| label: dendro
plotDendroAndColors(bwnet$dendrograms[[1]],
                    cbind(bwnet$unmergedColors, bwnet$colors),
                    c("unmerged", "merged"),
                    dendroLabels = FALSE,
                    addGuide     = TRUE,
                    hang         = 0.03,
                    guideHang    = 0.05)
```

## Step 6 — Module eigengenes & module–trait correlations

```{r}
#| label: MEs
module_eigengenes <- bwnet$MEs
head(module_eigengenes)
table(bwnet$colors)        # genes per module
```

```{r}
#| label: trait-binarize
# Build a numeric trait matrix: severity (Healthy / Convalescent / ICU / Moderate / Severe).
traits <- colData |>
  mutate(disease_state_bin = ifelse(grepl("severe", `disease state:ch1`,
                                          ignore.case = TRUE), 1, 0)) |>
  select(disease_state_bin)

severity.out <- binarizeCategoricalColumns(colData$`severity:ch1`,
                                           includePairwise = FALSE,
                                           includeLevelVsAll = TRUE,
                                           minCount = 1)
traits <- cbind(traits, severity.out)
```

```{r}
#| label: cor-trait
nSamples <- nrow(norm.counts)
nGenes   <- ncol(norm.counts)

module.trait.corr   <- cor(module_eigengenes, traits, use = "p")
module.trait.corr.p <- corPvalueStudent(module.trait.corr, nSamples)

heatmap.data <- merge(module_eigengenes, traits, by = "row.names")
heatmap.data <- column_to_rownames(heatmap.data, var = "Row.names")

CorLevelPlot(heatmap.data,
             x      = names(traits),
             y      = names(module_eigengenes),
             col    = c("blue1", "skyblue", "white", "pink", "red"))
```

::: {.callout-important title="Think about it"}
Modules with strong *|r|* and small p against the severity column are candidates for follow-up. What's the next thing you'd do with such a module?

<details>
<summary>Show answers</summary>

Two things in parallel:

1.  **GO / pathway enrichment** of the module's gene list (e.g. `clusterProfiler::enrichGO`) to give the module a biological label.
2.  **Hub gene identification**: rank module members by **module membership (kME)** = correlation of each gene's expression with the module eigengene. The top-kME genes are your most representative — and the most worth following up at the bench.
</details>
:::

## Step 7 — Hub genes within a module

```{r}
#| label: kme
module.gene.mapping <- as.data.frame(bwnet$colors)
module.gene.mapping |>
  rownames_to_column("gene") |>
  filter(`bwnet$colors` == "turquoise") |>     # change to your module of interest
  head(20)

# kME (gene–module correlation) — high values = hubs
mm <- as.data.frame(cor(norm.counts, module_eigengenes, use = "p"))
mm |>
  rownames_to_column("gene") |>
  arrange(desc(MEturquoise)) |>
  head(15)
```

## Stretch goals

-   **GO / KEGG enrichment** per module via `clusterProfiler` to label modules biologically.
-   **Validation in a held-out cohort** — re-project the module eigengenes onto a second study (e.g. another COVID-19 PBMC GSE) and check whether the trait correlation reproduces.
-   **Single-cell extension** — try [hdWGCNA](https://smorabit.github.io/hdWGCNA/) on a scRNA-seq object (metacells per cell type, signed network).

## What you've practiced

-   Outlier detection with `goodSamplesGenes` / hclust / PCA
-   VST normalization for downstream WGCNA
-   Picking a soft-thresholding power so the network is approximately scale-free
-   `blockwiseModules` end-to-end and reading the dendrogram
-   Module eigengenes, module–trait correlation, and hub genes via kME

## Credits

-   Patel: [`WGCNA.R`](https://github.com/kpatel427/YouTubeTutorials/blob/main/WGCNA.R)
-   Data: GEO **GSE152418** (Arunachalam et al. 2020, *Science*)
