---
title: "Appendix D - R Programming Fundamentals"
author: "Bill Cresko"
format:
  revealjs:
    footer: BioE_Stats_2026 - Knight Campus
    transition: fade
    transition-speed: slow
    theme: default
    slide-number: true
    code-fold: false
    code-overflow: wrap
    code-line-numbers: true
    highlight-style: github
    embed-resources: true
    css: cresko_style.css
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

theme_set(theme_minimal(base_size = 14))
set.seed(2026)
```

# Appendix D: R Programming Fundamentals {background-color="#2c3e50"}

## Overview

::: {.callout-tip title="Pre-requisite Material"}
This appendix provides a comprehensive introduction to the R programming language. R is the primary tool used throughout this course for statistical analysis and data visualization. See also Week 1 of the main lectures for additional R and RStudio basics.
:::

# Why R? {background-color="#2c3e50"}

## Why Use R? {.smaller}

-   Good general scripting tool for statistics and mathematics
-   Powerful, flexible, and free
-   Runs on all computer platforms
-   New enhancements coming out all the time
-   Superb data management & graphics capabilities
-   Reproducibility --- keep your scripts to see exactly what was done
-   You can write your own functions
-   Lots of online help available
-   Can use a nice GUI front end such as `RStudio`
-   Can embed your `R` analyses in dynamic, polished files using `Quarto Markdown`

## R Scripts and Quarto Markdown {.smaller}

-   Often we want to write scripts that can just be run --- `R scripts`
-   We can also embed code in Quarto Markdown files for more annotations
-   This improves interpretability and reproducibility
-   You can insert `R code chunks` into `Quarto Markdown` documents
-   A series of R commands that will be executed
-   Can add comments using hashtags `#`
-   Can have pipes (`|>`) to connect one step to the next

# R Basics {background-color="#2c3e50"}

## Basic Math in R {.smaller}

::: panel-tabset
### Output

```{r}
#| label: basic-math-output
#| echo: false
#| eval: true

4 * 4
(4 + 3 * 2^2)
```

### Code

```{r}
#| label: basic-math-code
#| echo: true
#| eval: false

4 * 4
(4 + 3 * 2^2)
```

### Interpretation

-   Commands can be submitted through the terminal, console, or scripts
-   R follows the normal priority of mathematical evaluation (PEMDAS)
-   Parentheses control order of operations just as in algebra
:::

## Assigning Variables {.smaller}

::: panel-tabset
### Output

```{r}
#| label: variables-output
#| echo: false
#| eval: true

x <- 2
x * 3
y <- x * 3
y - 2
```

### Code

```{r}
#| label: variables-code
#| echo: true
#| eval: false

x <- 2
x * 3
y <- x * 3
y - 2
```

### Interpretation

-   Variables are assigned values using the `<-` operator
-   Variable names must begin with a letter
-   R is case sensitive (`x` and `X` are different)
-   The result of calculations can be assigned to new variables
-   Names like `3y <- 3` or `3*y <- 3` will not work
:::

## Arithmetic Operations and Functions {.smaller}

::: panel-tabset
### Output

```{r}
#| label: functions-output
#| echo: false
#| eval: true

y <- 67
print(y)

x <- 124
z <- (x * y)^2
print(z)
```

### Code

```{r}
#| label: functions-code
#| echo: true
#| eval: false

# Common math functions
x + 2
x^2
log(x)    # natural log (built-in function)
sqrt(x)

# Assigning results
y <- 67
print(y)
x <- 124
z <- (x * y)^2
print(z)
```

### Interpretation

-   `log` is a built-in function --- the object goes in parentheses
-   Functions can take arguments that modify behavior (e.g., `log(x, base = 10)`)
-   The outcome of calculations can be assigned to new variables
-   Use `print()` to display values explicitly
:::

# Data Types and Structures {background-color="#2c3e50"}

## Strings {.smaller}

::: panel-tabset
### Output

```{r}
#| label: strings-output
#| echo: false
#| eval: true

x <- "I Love"
print(x)
y <- "Bioengineering"
print(y)
z <- c(x, y)
print(z)
```

### Code

```{r}
#| label: strings-code
#| echo: true
#| eval: false

x <- "I Love"
print(x)
y <- "Bioengineering"
print(y)
z <- c(x, y)    # c stands for concatenate
print(z)
```

### Interpretation

-   Characters need to be set off by quotation marks
-   The `c()` function stands for `concatenate`
-   Reusing variable names overwrites previous assignments
-   Good practice: use descriptive, unique names for each variable
:::

## Vectors {.smaller}

::: panel-tabset
### Output

```{r}
#| label: vectors-output
#| echo: false
#| eval: true

x <- c(2, 3, 4, 2, 1, 2, 4, 5, 10, 8, 9)
print(x)

y <- x^2
print(y)
```

### Code

```{r}
#| label: vectors-code
#| echo: true
#| eval: false

# Create a vector with c()
x <- c(2, 3, 4, 2, 1, 2, 4, 5, 10, 8, 9)
print(x)

# Operations work element-wise
y <- x^2
print(y)

# Plot vectors
plot(x, y)
```

### Interpretation

-   R thinks in terms of **vectors** (a list of values)
-   Vectors can be assigned directly using the `c()` function
-   Mathematical operations on vectors work element-wise
-   Writing programs with vectors in mind will simplify most things
:::

## Types of Vectors {.smaller}

::: panel-tabset
### Types

| Type   | Description                              | Example                    |
|--------|------------------------------------------|----------------------------|
| `int`  | Integers                                 | `1L, 2L, 3L`              |
| `dbl`  | Doubles (real numbers)                   | `3.14, 2.718`              |
| `chr`  | Character strings                        | `"hello", "DNA"`           |
| `lgl`  | Logical (TRUE/FALSE)                     | `TRUE, FALSE, NA`          |
| `fctr` | Factors (categorical variables)          | `factor(c("a","b","a"))`   |
| `dttm` | Date-times                               | `Sys.time()`               |
| `date` | Dates                                    | `Sys.Date()`               |

### Special Values

-   Logical vectors can take: `FALSE`, `TRUE`, or `NA` (not available)
-   Doubles have four special values: `NA`, `NaN` (not a number), `Inf`, `-Inf`
-   Integer and double vectors are collectively known as **numeric** vectors
-   In R, numbers are doubles by default
:::

## Basic Statistics on Vectors {.smaller}

::: panel-tabset
### Output

```{r}
#| label: stats-output
#| echo: false
#| eval: true

x <- c(2, 3, 4, 2, 1, 2, 4, 5, 10, 8, 9)
cat("Mean:", mean(x), "\n")
cat("Median:", median(x), "\n")
cat("Variance:", var(x), "\n")
cat("SD:", sd(x), "\n")
cat("Sum:", sum(x), "\n")
cat("Length:", length(x), "\n")
```

### Code

```{r}
#| label: stats-code
#| echo: true
#| eval: false

mean(x)
median(x)
var(x)
sd(x)        # standard deviation
log(x)
sqrt(x)
sum(x)
length(x)
sample(x, replace = TRUE)  # random sample with replacement
```

### Interpretation

-   Many built-in functions operate on vectors
-   `sample()` has an argument `replace = T` --- arguments modify function behavior
-   There are many arguments for each function, some of which are defaults
-   Use `?function_name` to see the help page for any function
:::

## Creating Vectors with Sequences {.smaller}

::: panel-tabset
### Output

```{r}
#| label: sequences-output
#| echo: false
#| eval: true

seq_1 <- seq(0.0, 10.0, by = 0.1)
head(seq_1, 20)

seq_2 <- seq(10.0, 0.0, by = -0.1)
head(seq_2, 20)
```

### Code

```{r}
#| label: sequences-code
#| echo: true
#| eval: false

# Ascending sequence
seq_1 <- seq(0.0, 10.0, by = 0.1)
print(seq_1)

# Descending sequence
seq_2 <- seq(10.0, 0.0, by = -0.1)
print(seq_2)

# Element-wise operations
seq_square <- (seq_2)^2
print(seq_square)
```

### Interpretation

-   `seq()` creates sequences with a specified start, end, and step size
-   The three arguments in the parentheses are: start, end, and step
-   Operations on sequences work element-wise, just like simple vectors
:::

## Getting Help in R {.smaller}

::: panel-tabset
### Methods

``` r
help(mean)           # Open help page
?mean                # Same as help()
example(mean)        # Show examples
help.search("mean")  # Search help pages
apropos("mean")      # Find functions with "mean" in name
args(mean)           # Show function arguments
```

### Interpretation

-   Help pages exist for all functions explaining parameters and usage
-   Function calls have a simple structure: function name + parentheses + parameters
-   The `example()` function is especially useful for seeing how functions work
-   When in doubt, check the help page!
:::

# Sampling from Distributions {background-color="#2c3e50"}

## Drawing Random Samples {.smaller}

::: panel-tabset
### Output

```{r}
#| label: distributions-output
#| echo: false
#| eval: true

x <- rnorm(10000, 0, 10)
hist(x, main = "Random Normal Distribution",
     xlab = "Value", col = "steelblue", border = "white")
```

### Code

```{r}
#| label: distributions-code
#| echo: true
#| eval: false

# Draw from normal distribution
x <- rnorm(10000, 0, 10)
hist(x)

# Draw uniform random integers
y <- sample(1:10000, 10000, replace = TRUE)

# Combine and plot
xy <- cbind(x, y)
plot(x, y)
```

### Interpretation

-   `rnorm(n, mean, sd)` draws `n` random values from a normal distribution
-   `sample()` draws random values with equal probability
-   `cbind()` binds columns together into a matrix
-   These tools are fundamental for simulation and statistical analysis
:::

## Overlaying Density Curves {.smaller}

::: panel-tabset
### Output

```{r}
#| label: density-output
#| echo: false
#| eval: true

x <- rnorm(1000, 0, 100)
hist(x, xlim = c(-500, 500), probability = TRUE,
     main = "Histogram with Density Curve",
     col = "steelblue", border = "white")
curve(dnorm(x, 0, 100), xlim = c(-500, 500), add = TRUE, col = 'red', lwd = 2)
```

### Code

```{r}
#| label: density-code
#| echo: true
#| eval: false

x <- rnorm(1000, 0, 100)
hist(x, xlim = c(-500, 500), probability = TRUE)
curve(dnorm(x, 0, 100), xlim = c(-500, 500), add = TRUE, col = 'red', lwd = 2)
```

### Interpretation

-   `dnorm()` generates the probability density function
-   `curve()` plots a function --- `add = TRUE` overlays on existing plot
-   The red curve shows the theoretical distribution we sampled from
-   Histograms with density overlays help verify sampling behavior
:::

# Data Frames {background-color="#2c3e50"}

## Creating Data Frames {.smaller}

::: panel-tabset
### Output

```{r}
#| label: dataframes-output
#| echo: false
#| eval: true

hydrogel_concentration <- factor(c("low", "high", "high", "high",
                                    "medium", "medium", "medium", "low"))
compression <- c(3.4, 3.4, 8.4, 3, 5.6, 8.1, 8.3, 4.5)
conductivity <- c(0, 9.2, 3.8, 5, 5.6, 4.1, 7.1, 5.3)

mydata <- data.frame(hydrogel_concentration, compression, conductivity)
row.names(mydata) <- c("Sample_1", "Sample_2", "Sample_3", "Sample_4",
                        "Sample_5", "Sample_6", "Sample_7", "Sample_8")
print(mydata)
```

### Code

```{r}
#| label: dataframes-code
#| echo: true
#| eval: false

# Set up vectors as columns
hydrogel_concentration <- factor(c("low", "high", "high", "high",
                                    "medium", "medium", "medium", "low"))
compression <- c(3.4, 3.4, 8.4, 3, 5.6, 8.1, 8.3, 4.5)
conductivity <- c(0, 9.2, 3.8, 5, 5.6, 4.1, 7.1, 5.3)

# Create data frame
mydata <- data.frame(hydrogel_concentration, compression, conductivity)
row.names(mydata) <- c("Sample_1", "Sample_2", "Sample_3", "Sample_4",
                        "Sample_5", "Sample_6", "Sample_7", "Sample_8")
print(mydata)
```

### Interpretation

-   Data frames are the primary data structure for tabular data in R
-   Vectors become columns when combined with `data.frame()`
-   `factor()` creates categorical variables with fixed levels
-   `row.names()` assigns labels to each row
:::

## Reading and Writing Data {.smaller}

::: panel-tabset
### Reading

``` r
# Comma-separated files
YourFile <- read.table('yourfile.csv', header = TRUE, row.names = 1, sep = ',')
# OR use the convenience function:
YourFile <- read.csv('yourfile.csv')

# Tab-separated files
YourFile <- read.table('yourfile.txt', header = TRUE, row.names = 1, sep = '\t')

# Tidyverse versions (return tibbles)
YourFile <- read_csv('yourfile.csv')
YourFile <- read_tsv('yourfile.txt')

# Excel files
library(readxl)
YourFile <- read_excel('yourfile.xlsx')
```

### Writing

``` r
# Write to CSV
write.table(YourFile, "yourfile.csv", quote = FALSE, row.names = TRUE, sep = ",")
# OR:
write.csv(YourFile, "yourfile.csv")

# Write to TSV
write.table(YourFile, "yourfile.txt", quote = FALSE, row.names = TRUE, sep = "\t")

# Tidyverse version
write_csv(YourFile, "yourfile.csv")
```

### Interpretation

-   `header = TRUE` indicates the first row contains column names
-   `row.names = 1` indicates the first column contains row names
-   `sep` specifies the delimiter (`,` for CSV, `\t` for TSV)
-   The tidyverse `read_csv()` and `read_tsv()` functions return tibbles (enhanced data frames)
:::

## Indexing Data Frames {.smaller}

::: panel-tabset
### Output

```{r}
#| label: indexing-output
#| echo: false
#| eval: true

# Using the mydata frame from before
print(mydata[, 2])         # Column 2
print(mydata$compression)  # By name
print(mydata[2, ])         # Row 2
```

### Code

```{r}
#| label: indexing-code
#| echo: true
#| eval: false

# Access by column number
print(mydata[, 2])

# Access by column name (preferred)
print(mydata$compression)

# Access by row number
print(mydata[2, ])

# Access specific cell
print(mydata[2, 3])

# Subset based on condition
print(mydata$compression > 5)

# Plot two columns
plot(mydata$compression, mydata$conductivity)
```

### Interpretation

-   `[row, column]` notation accesses specific elements
-   `$` notation accesses columns by name (preferred --- more readable)
-   Leaving a dimension empty selects all (e.g., `[, 2]` = all rows, column 2)
-   Logical conditions return TRUE/FALSE vectors for subsetting
:::

# Base R Visualization {background-color="#2c3e50"}

## Basic Plotting {.smaller}

::: panel-tabset
### Output

```{r}
#| label: plot-output
#| echo: false
#| eval: true

seq_1 <- seq(0.0, 10.0, by = 0.1)
plot(seq_1, xlab = "space", ylab = "function of space",
     type = "p", col = "red", main = "Basic R Plot")
```

### Code

```{r}
#| label: plot-code
#| echo: true
#| eval: false

seq_1 <- seq(0.0, 10.0, by = 0.1)
plot(seq_1,
     xlab = "space",
     ylab = "function of space",
     type = "p",        # "p" = points, "l" = lines, "b" = both
     col = "red",
     main = "Basic R Plot")
```

### Interpretation

-   `plot()` is the main high-level plotting function in base R
-   Arguments like `xlab`, `ylab`, `col`, and `type` control appearance
-   `type` options: `"p"` (points), `"l"` (lines), `"b"` (both), `"h"` (histogram-like)
-   Most people use `ggplot2` for publication-quality graphics (see Week 2)
:::

## Multi-Panel Figures {.smaller}

::: panel-tabset
### Output

```{r}
#| label: multipanel-output
#| echo: false
#| eval: true

seq_1 <- seq(0.0, 10.0, by = 0.1)
seq_2 <- seq(10.0, 0.0, by = -0.1)
seq_square <- (seq_2)^2

par(mfrow = c(2, 2))
plot(seq_1, xlab = "time", ylab = "p in population 1", type = "p", col = 'red')
plot(seq_2, xlab = "time", ylab = "p in population 2", type = "p", col = 'green')
plot(seq_square, xlab = "time", ylab = "p2 in population 2", type = "p", col = 'blue')
plot(seq_1^2, xlab = "time", ylab = "p in population 1", type = "l", col = 'orange')
```

### Code

```{r}
#| label: multipanel-code
#| echo: true
#| eval: false

par(mfrow = c(2, 2))  # 2 rows, 2 columns
plot(seq_1, xlab = "time", ylab = "p in pop 1", type = "p", col = 'red')
plot(seq_2, xlab = "time", ylab = "p in pop 2", type = "p", col = 'green')
plot(seq_square, xlab = "time", ylab = "p2 in pop 2", type = "p", col = 'blue')
plot(seq_1^2, xlab = "time", ylab = "p in pop 1", type = "l", col = 'orange')
```

### Interpretation

-   `par(mfrow = c(rows, cols))` sets up a multi-panel layout
-   Subsequent `plot()` calls fill panels left-to-right, top-to-bottom
-   This is useful for comparing related visualizations side by side
:::

## Summary Statistics and Figures {.smaller}

::: panel-tabset
### Output

```{r}
#| label: summary-output
#| echo: false
#| eval: true

summary(mydata$compression)
boxplot(compression ~ hydrogel_concentration, data = mydata,
        col = "steelblue", ylab = "Compression",
        xlab = "Hydrogel Concentration",
        main = "Compression by Hydrogel Concentration")
```

### Code

```{r}
#| label: summary-code
#| echo: true
#| eval: false

# Summary statistics
summary(mydata$compression)

# Histogram
hist(mydata$compression)

# Boxplot by group
boxplot(compression ~ hydrogel_concentration, data = mydata,
        col = "steelblue",
        ylab = "Compression",
        xlab = "Hydrogel Concentration",
        main = "Compression by Hydrogel Concentration")
```

### Interpretation

-   `summary()` provides min, max, quartiles, mean, and median
-   The formula notation `y ~ x` means "y grouped by x"
-   Boxplots show median, IQR, and potential outliers
-   These are quick exploratory tools --- use `ggplot2` for publication graphics
:::

## Tidy Data Principles {.smaller}

::: {.callout-note title="Rules of Thumb for Data Organization"}
-   Store data in nonproprietary formats (plain ASCII text / flat files)
-   Leave an uncorrected original file when doing analyses
-   Use descriptive names for your data files and variables
-   Include a header line with descriptive variable names
-   Maintain effective metadata about the data (data dictionary)
-   When you add observations, add rows
-   When you add variables, add columns
-   A column of data should contain only one data type
:::

## Types of Data {.smaller}

|            |           |               |              |
|:----------:|:---------:|:-------------:|:------------:|
| Categorical|           | Quantitative  |              |
| Ordinal    | Nominal   | Ratio         | Interval     |
| small, medium, large | apples, oranges | kilograms, dollars | temperature, calendar year |
| ordered character | character | numeric | integer |

`Factor` is a special type of character variable that R uses to represent categorical variables with fixed possible values.
