---
title: "Single Cell RNA-seq Workshop - Lecture Template"
author: "*Author name*"
format:
  revealjs:
    footer: Single Cell RNA-seq Workshop
    transition: fade
    transition-speed: slow
    theme: default
    slide-number: true
    code-fold: false
    code-overflow: wrap
    code-line-numbers: true
    highlight-style: github
    embed-resources: true
    css: cresko_style.css
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false

# Core data manipulation and visualization
library(tidyverse)
library(knitr)

# scRNA-seq packages (uncomment what the lecture needs)
# library(Seurat)
# library(SingleCellExperiment)
# library(scran)
# library(scater)

theme_set(theme_minimal(base_size = 14))
set.seed(2026)
```

# Lecture Title {background-color="#2c3e50"}

## Topics

::: incremental
-   *Topic 1*
-   *Topic 2*
-   *Topic 3*
:::

::: callout-note
**Readings:** *Add readings / references here*
:::

## Packages for This Lecture

::: panel-tabset
### Install

```{r}
#| label: packages-install
#| eval: false
#| echo: true

# Install packages (run once)
install.packages(c("tidyverse"))
# BiocManager::install(c("SingleCellExperiment", "scran", "scater"))
```

### Load

```{r}
#| label: packages-load
#| eval: false
#| echo: true

library(tidyverse)
# library(Seurat)
# library(SingleCellExperiment)
```
:::

# Section 1 — Motivation & Background {background-color="#2c3e50"}

## Why this matters

*Motivate the topic. What biological or methodological question are we addressing? Why does a single cell view change the answer?*

## Key concepts

::: incremental
-   *Concept 1*
-   *Concept 2*
-   *Concept 3*
:::

# Section 2 — Methods {background-color="#2c3e50"}

## Approach

*Describe the analytical approach at a conceptual level.*

## Code demonstration

```{r}
#| label: demo
#| eval: false

# Replace with the demonstration code for this lecture
```

# Section 3 — Interpretation & Pitfalls {background-color="#2c3e50"}

## Interpreting results

*What do the outputs mean? What are common misinterpretations?*

## Common pitfalls

::: incremental
-   *Pitfall 1*
-   *Pitfall 2*
-   *Pitfall 3*
:::

# Wrap-Up {background-color="#2c3e50"}

## Summary

::: incremental
-   *Takeaway 1*
-   *Takeaway 2*
-   *Takeaway 3*
:::

## Further reading

-   *Reference / link 1*
-   *Reference / link 2*
